---
title: "tantale dev notes"
output: html_document
---

Personal scratchpad: reminders and copy-paste-able code for solo dev
sessions. Not a vignette, not built by `R CMD build`/`check`/pkgdown
(`dev/` is in `.Rbuildignore`), not meant to be knit start to finish --
run chunks one at a time, in whichever order the task needs. `eval =
FALSE` everywhere on purpose, so an accidental knit does nothing.

For the *why* behind any of this, see `dev/CLAUDE.md` (standing rules)
and `dev/restructuring-notes.md` (the ledger, full history -- read its
`START HERE` block, not the whole file).

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Before touching anything

```{r env-check}
library(tantale)
tantale_setup() # confirms conda binary/root/env and every external tool + pinned version
```

```{r clean-check}
# Uncommitted work? Anything unexpected lying around?
system("git status --short")
```

## Load, document, install

```{r load-all}
devtools::load_all() # fast, in-session only -- what you use while editing R/
```

```{r document}
devtools::document() # regenerate roxygen docs + NAMESPACE after changing @-tags
```

```{r install}
# Required before ANY pkgdown/docs rebuild that follows a change to R/ --
# quarto's article rendering spawns a separate R session that does
# library(tantale) against the INSTALLED copy, not devtools::load_all()'s
# patched session. Skipping this costs a confusing "could not find
# function" buried in a generic "Error running quarto CLI" (see ledger
# START HERE, 2026-09-20).
devtools::install(quick = TRUE, upgrade = FALSE)
```

```{r install-confirm}
# Confirm the install actually took, in a FRESH session -- not this one.
system('Rscript -e \'exists("tales_compare_distal", where = asNamespace("tantale"))\'')
```

## Tests

```{r test-one}
# Targeted run -- the default. Full suite only if the change ripples broadly.
devtools::load_all()
testthat::test_file("tests/testthat/test_tales_class.R")
```

```{r test-all}
# Whole suite -- slow (several minutes), only when a change could plausibly
# reach broadly (e.g. touching tales_class.R, the anomaly checks, or a
# residue-type change).
devtools::test()
```

```{r golden-rebaseline-reminder}
# If a golden-baseline test reports a diff: use the `golden-rebaseline`
# skill, not a manual accept. Explain every changed row BEFORE accepting --
# an accepted-but-unexplained snapshot is indistinguishable from a correct
# one (ledger §8.1d/§8.3).
```

## R CMD check

```{r check}
devtools::check()
```

```{r pkgdown-check}
# After adding or retiring an exported topic -- a dangling _pkgdown.yml
# entry is a hard error in build_site(), and R CMD check does not look at
# that file at all.
pkgdown::check_pkgdown()
```

## Building the docs site

```{r about-release-mode}
# `docs/` is published in RELEASE mode directly -- there is no docs/dev/
# any more (dropped 2026-09-20, ledger §15). `_pkgdown.yml`'s
# `development: mode:` is hardcoded to "release" (since 2026-09-21,
# tested directly first) for exactly this reason, so every build_*
# call below is a bare call -- no override needed, no docs/dev/ risk.
```

```{r cache-clear}
# Only needed after actually changing tell_tales()/tales_compare_distal()/
# tales_group_kmedoids() -- the shared article cache is NOT content-hashed
# (ledger §15), so a stale cache silently keeps serving old results
# otherwise.
unlink("vignettes/articles/_cache", recursive = TRUE)
```

```{r build-one-article}
# Iterating on a single article -- the normal case while writing/editing one.
pkgdown::build_article("articles/tale_msa")
```

```{r build-home-reference-news}
# The non-article parts of the site. Cheap, safe to run any time.
pkgdown::build_home()
pkgdown::build_reference()
pkgdown::build_news()
```

```{r build-all-articles}
# The full articles rebuild. Do NOT use pkgdown::build_articles()/
# build_site() for this -- both funnel every .qmd under vignettes/
# through ONE `quarto render` project pass whose internal order is not
# alphabetical, not _pkgdown.yml's navbar order, and not controllable via
# a committed vignettes/_quarto.yaml (all three tried and falsified,
# ledger §15). tale_classification.qmd MUST render before the other
# three, since it's the only one that computes the shared pipeline; a
# whole-project pass gives no such guarantee (tale_target_prediction.qmd
# was observed rendering second, right after getting_started.qmd).
#
# This loop instead renders each file exactly once, in an order you
# control -- classification first, everything else after -- with no
# double-rendering and no dependence on quarto's own ordering at all.
article_names <- c(
  "articles/tale_classification", # MUST be first -- computes + caches the pipeline
  "articles/tale_mining",
  "articles/tales_class",
  "articles/tale_msa",
  "articles/tales_msa_class",
  "articles/tale_target_prediction",
  "getting_started"
)
for (nm in article_names) {
  pkgdown::build_article(nm)
}
pkgdown::build_articles_index()
```

```{r build-everything}
# A true from-scratch full rebuild, in the same order build_site_local()
# itself uses internally, but with the article step swapped for the loop
# above. Confirm `docs/` is clean in git status before AND after.
pkgdown::init_site()
pkgdown::build_home()
pkgdown::build_reference()
for (nm in c(
  "articles/tale_classification", "articles/tale_mining", "articles/tales_class",
  "articles/tale_msa", "articles/tales_msa_class", "articles/tale_target_prediction",
  "getting_started"
)) {
  pkgdown::build_article(nm)
}
pkgdown::build_articles_index()
pkgdown::build_news()
```

## Version bump

```{r version-bump}
# Pre-publication scheme: 0.9.x, bumping the 4th digit (0.9.9NNN) for
# routine work -- 1.0.0 at actual release. Keep the third component
# >= 9000 -- not because docs/ builds need it any more (mode: release
# in _pkgdown.yml forces release publishing regardless of version), but
# because a value below 9000 changes what pkgdown considers a
# "development version" in other contexts too (e.g. its own
# version-comparison logic), and nothing has re-checked that this still
# doesn't matter now that docs/dev/ is gone.
usethis::use_version("dev") # bumps the 4th digit; confirm DESCRIPTION after
```
