---
title: "tantale function maps"
format:
  html:
    embed-resources: true
    page-layout: full
    code-fold: true
    code-summary: "Code"
execute:
  warning: false
---

Maps of the package for finding one's way around it. Dev-only (`dev/`
is in `.Rbuildignore`); the data-flow map also appears on the site. Ledger §29 has
the decisions behind them.

- **Data flow:** the exported functions and the objects they pass to each
  other, from reading TALE sequences to classes, alignments and target
  predictions.
- **Call graph:** which function calls which in `R/`, internals included.
- **Matrices and lists of vectors:** the functions, internals included,
  that take or return a matrix or a plain list of vectors, recorded from
  the test suite.

Render the page from the package root with
`quarto render dev/function-graph.qmd`.

## Data flow

Which object each exported function takes and returns. Boxes are object
classes, dots are functions. An arrow into a function is an argument
(labelled when it is not the first one); an arrow out of it is the return
value (labelled `$name` for an element of a returned list).

The links were recorded by running the test suite with every exported
function traced (`dev/function-graph-dataflow.R`), so they show what the
functions actually received and returned. Rerun that script after adding
an exported function or changing what one takes or returns; the lists
under the graph show when the recording is out of date. Plain R types
(tibbles, character vectors, plots) are left out of the drawing and given
in each function's tooltip, apart from one link through files on disk:
`tell_tales()` writes a results folder that `tales_from_telltales()` reads.

Validators, predicates, and the `print`, `format` and `[` methods are left
out: they take an object and return it unchanged, or return a logical.
The site shows this map at the top of the reference index, and a reduced
version on its home page; `Rscript pkgdown/function_map.R` redraws both
after the recorder has run.

```{r dataflow}
#| code-summary: "Code: data-flow view"
# The same code draws the site's maps (pkgdown/function_map.R).
source(file.path("..", "pkgdown", "function_map.R"))
fm <- function_map_data("..")
flow <- fm$flow
exported_now <- fm$exported_now
s3_now <- fm$s3_now
recorded_on <- fm$recorded_on
function_map_widget(fm)
```

`r recorded_on`.

```{r dataflow-coverage}
#| results: asis
#| code-summary: "Code: recording coverage"
not_recorded <- setdiff(exported_now, flow$fn)
gone <- setdiff(unique(flow$fn), c(exported_now, s3_now))
code_list <- function(x, suffix = "") {
  if (length(x)) paste0("`", x, suffix, "`", collapse = ", ") else "none"
}
cat(paste0(
  "- Exported functions the test suite never calls, so absent from the recording: ",
  code_list(not_recorded, "()"), "\n",
  "- Recorded functions no longer exported or registered: ", code_list(gone), "\n"))
```

## Call graph

The call graph is rebuilt from the source files on every render. Nothing
is loaded or run: the files in `R/` are parsed.

```{r setup}
#| code-summary: "Code: parse R/ into nodes and edges"
r_dir <- file.path("..", "R")
files <- sort(list.files(r_dir, pattern = "\\.[Rr]$", full.names = TRUE))

# Exported functions and registered S3 methods, from NAMESPACE.
ns_lines <- readLines(file.path("..", "NAMESPACE"))
exports <- sub("^export\\((.*)\\)$", "\\1", grep("^export\\(", ns_lines, value = TRUE))
s3 <- regmatches(ns_lines, regexec("^S3method\\((.*),(.*)\\)$", ns_lines))
s3 <- do.call(rbind, lapply(s3[lengths(s3) == 3], function(m) {
  data.frame(generic = gsub("\"|^.*::", "", m[2]), class = m[3])
}))
s3$method <- paste(s3$generic, s3$class, sep = ".")

# The comment block right above a definition: its first line serves as a
# title in the tooltip (the roxygen title for documented functions).
comment_title <- function(src, line) {
  i <- line - 1
  block <- character()
  while (i >= 1 && grepl("^\\s*#", src[i])) {
    block <- c(src[i], block)
    i <- i - 1
  }
  block <- trimws(sub("^\\s*#'?", "", block))
  block <- block[nzchar(block) & !grepl("^@", block)]
  if (length(block)) block[1] else ""
}

# Top-level definitions of the form `name <- function(...)`.
defs <- list()
for (f in files) {
  ex <- parse(f, keep.source = TRUE)
  sr <- attr(ex, "srcref")
  src <- readLines(f)
  for (i in seq_along(ex)) {
    e <- ex[[i]]
    if (is.call(e) && as.character(e[[1]])[1] %in% c("<-", "=") &&
        is.call(e[[3]]) && identical(e[[3]][[1]], as.name("function"))) {
      defs[[length(defs) + 1]] <- data.frame(
        name = as.character(e[[2]]), file = basename(f),
        line1 = sr[[i]][1], line2 = sr[[i]][3],
        title = comment_title(src, sr[[i]][1]))
    }
  }
}
defs <- do.call(rbind, defs)
stopifnot(!anyDuplicated(defs$name))
defs$kind <- ifelse(defs$name %in% exports, "exported",
                    ifelse(defs$name %in% s3$method, "S3 method", "internal"))

# Edges, read off each definition's tokens.
#   call:      name(...), unqualified or tantale::name(...), or an
#              infix operator defined in the package (`%||%`)
#   reference: name passed as a value, e.g. vapply(x, .helper, ...), and
#              not shadowed by a local variable of the same name
edges <- list()
for (f in files) {
  pd <- getParseData(parse(f, keep.source = TRUE))
  pd <- pd[pd$terminal & pd$token != "COMMENT", ]
  pd <- pd[order(pd$line1, pd$col1), ]
  d <- defs[defs$file == basename(f), ]
  for (k in seq_len(nrow(d))) {
    tk <- pd[pd$line1 >= d$line1[k] & pd$line2 <= d$line2[k], ]
    n <- nrow(tk)
    prev_tok <- c("", tk$token[-n])
    prev2_tok <- c("", "", tk$token[seq_len(max(n - 2, 0))])
    prev2_txt <- c("", "", tk$text[seq_len(max(n - 2, 0))])
    next_tok <- c(tk$token[-1], "")
    next_txt <- c(tk$text[-1], "")
    foreign <- prev_tok %in% c("NS_GET", "NS_GET_INT") & prev2_txt != "tantale"
    calls <- tk$text[tk$token %in% c("SYMBOL_FUNCTION_CALL", "SPECIAL") & !foreign]
    locals <- tk$text[tk$token == "SYMBOL_FORMALS" |
                        (tk$token == "SYMBOL" &
                           (next_tok %in% c("LEFT_ASSIGN", "EQ_ASSIGN") |
                              next_txt == "%<>%" | prev2_tok == "FOR"))]
    refs <- tk$text[tk$token == "SYMBOL" & !prev_tok %in% c("'$'", "'@'") &
                      !foreign & !tk$text %in% locals]
    cl <- intersect(calls, defs$name)
    rf <- setdiff(intersect(refs, defs$name), calls)
    edges[[k + length(edges)]] <- data.frame(
      from = rep(d$name[k], length(cl) + length(rf)), to = c(cl, rf),
      type = rep(c("call", "reference"), c(length(cl), length(rf))))
  }
}
edges <- do.call(rbind, edges)

# S3 dispatch from generics defined in the package to their methods.
own_generics <- intersect(s3$generic, defs$name)
dispatch <- s3[s3$generic %in% own_generics & s3$method %in% defs$name, ]
edges <- rbind(edges, data.frame(from = dispatch$generic, to = dispatch$method,
                                 type = rep("dispatch", nrow(dispatch))))
edges <- edges[edges$from != edges$to, ]

edges$cross_file <- defs$file[match(edges$from, defs$name)] !=
  defs$file[match(edges$to, defs$name)]

defs$callers <- vapply(defs$name, function(x) sum(edges$to == x), integer(1))
defs$callees <- vapply(defs$name, function(x) sum(edges$from == x), integer(1))
```

`R/` has `r length(files)` files, `r length(unique(defs$file))` of which
define functions, `r nrow(defs)` in all:
`r sum(defs$kind == "exported")` exported,
`r sum(defs$kind == "S3 method")` S3 methods and
`r sum(defs$kind == "internal")` internal. They are linked by
`r nrow(edges)` edges, of which `r sum(edges$cross_file)`
cross from one file to another.

The graph opens with one node per file, sized by its number of functions.
An arrow from A to B means code in A calls code in B.

- **Double-click a file** to open it into its functions. **Double-click a
  function** to fold its file back up. The buttons above the graph open or
  fold every file at once. A file that holds a single function shows that
  function in place of a file node.
- **Click a function** to highlight its direct callers and callees and fade
  the rest. The drop-down list selects a function by name. Open its file
  first: a function inside a folded file cannot be highlighted.
- Drag to pan, scroll to zoom, drag a node to move it.

Colours give the kind of function (legend at the top left). Grey arrows
are calls. Dashed arrows are the other two kinds of link: purple for a
function passed by name as an argument (for example to `vapply()`), green
for S3 dispatch from a generic defined in the package to its methods.

```{r graph}
#| code-summary: "Code: draw the graph"
library(visNetwork)

kind_colour <- c(exported = "#2a78c1", "S3 method" = "#1b9e77",
                 internal = "#c9c9c9")
file_colour <- "#e8a33d"

nodes <- data.frame(
  id = defs$name,
  label = defs$name,
  group = defs$kind,
  file = defs$file,
  size = 10 + 2 * sqrt(defs$callers),
  title = sprintf(
    "<b>%s</b><br>%s:%d (%d lines)<br>%s<br>%d callers, %d callees%s",
    htmltools::htmlEscape(defs$name), defs$file, defs$line1,
    defs$line2 - defs$line1 + 1, defs$kind, defs$callers, defs$callees,
    ifelse(nzchar(defs$title),
           paste0("<br><i>", htmltools::htmlEscape(defs$title), "</i>"), ""))
)

vis_edges <- data.frame(
  from = edges$from, to = edges$to,
  arrows = "to",
  dashes = edges$type != "call",
  color = unname(c(call = "#7a7a7a", reference = "#b07aa1",
                   dispatch = "#1b9e77")[edges$type]),
  title = edges$type
)

# Folding by file is done here rather than with visClusteringByGroup(),
# which keys on `group` and repaints opened nodes in vis.js's default
# palette. vis.js does not make a cluster of a single node, so a file
# holding one function shows that function.
fold_by_file <- sprintf("
function(el, x) {
  var network = document.getElementById('graph' + el.id).chart;
  var nodes = network.body.data.nodes;
  var files = Array.from(new Set(nodes.get().map(function(n) { return n.file; }))).sort();
  function fold(f) {
    var n = nodes.get({filter: function(d) { return d.file === f; }}).length;
    if (n < 2 || network.findNode('file:' + f).length) return;
    network.cluster({
      joinCondition: function(d) { return d.file === f; },
      clusterNodeProperties: {
        id: 'file:' + f, label: f, shape: 'dot', color: '%s', borderWidth: 2,
        size: 12 + 4 * Math.sqrt(n),
        title: '<b>' + f + '</b><br>' + n + ' functions<br>double-click to open'
      }
    });
  }
  function openAll() {
    files.forEach(function(f) {
      if (network.isCluster('file:' + f)) network.openCluster('file:' + f);
    });
  }
  files.forEach(fold);
  network.on('doubleClick', function(p) {
    if (p.nodes.length !== 1) return;
    var id = p.nodes[0];
    if (network.isCluster(id)) {
      network.openCluster(id, {releaseFunction: function(c, pos) { return pos; }});
    } else {
      fold(nodes.get(id).file);
    }
  });
  var bar = document.createElement('div');
  bar.style.margin = '0.3em 0';
  [['Open all files', openAll], ['Fold all files', function() { files.forEach(fold); }]]
    .forEach(function(b) {
      var btn = document.createElement('button');
      btn.textContent = b[0];
      btn.style.marginRight = '0.5em';
      btn.onclick = function() { b[1](); setTimeout(function() { network.fit(); }, 300); };
      bar.appendChild(btn);
    });
  el.parentNode.insertBefore(bar, el);
}", file_colour)

legend_nodes <- data.frame(label = c("exported", "S3\nmethod", "internal", "file"),
                           color = c(unname(kind_colour), file_colour),
                           shape = "dot")

visNetwork(nodes, vis_edges, width = "100%", height = "85vh") |>
  visGroups(groupname = "exported", color = kind_colour[["exported"]], shape = "dot") |>
  visGroups(groupname = "S3 method", color = kind_colour[["S3 method"]], shape = "dot") |>
  visGroups(groupname = "internal", color = kind_colour[["internal"]], shape = "dot") |>
  visNodes(font = list(size = 14)) |>
  visEdges(smooth = FALSE, arrows = list(to = list(scaleFactor = 0.5))) |>
  visPhysics(solver = "forceAtlas2Based",
             forceAtlas2Based = list(gravitationalConstant = -80, springLength = 120),
             stabilization = list(iterations = 400)) |>
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1,
                                     algorithm = "hierarchical"),
             nodesIdSelection = list(enabled = TRUE, main = "Select a function")) |>
  visInteraction(navigationButtons = TRUE, tooltipDelay = 150) |>
  visLegend(addNodes = legend_nodes, useGroups = FALSE, position = "left",
            width = 0.1) |>
  htmlwidgets::onRender(fold_by_file)
```

## Functions

Sortable and searchable. `callers` and `callees` count distinct functions
in the package, whatever the edge type.

```{r table}
#| code-summary: "Code: function table"
library(reactable)
tab <- defs[order(defs$file, defs$line1),
            c("name", "file", "line1", "kind", "callers", "callees", "title")]
names(tab)[3] <- "line"
reactable(tab, searchable = TRUE, filterable = TRUE, pagination = FALSE,
          height = 600, compact = TRUE, striped = TRUE,
          columns = list(name = colDef(minWidth = 180, style = list(fontFamily = "monospace")),
                         title = colDef(minWidth = 300),
                         line = colDef(width = 60),
                         callers = colDef(width = 75),
                         callees = colDef(width = 75)))
```

## Matrices and lists of vectors

Functions, exported or internal, that take or return a matrix, an
unclassed list of atomic vectors (`list(c("NI", "HD"), c("NG"))`), or an
unclassed list of matrices. These are the pre-class representations of
alignments and RVD strings. Data frames and classed objects (`tales`,
`tales_msa`, `pairwise_distances`, ggplot objects) do not count, even
though some are lists underneath.

Recorded by running the test suite with every function in the namespace
traced (`dev/function-graph-shapes.R`): an argument is listed when a test
passed it a value of one of these shapes, a return value when the function
returned one (`$name` for an element of a returned list). Rerun the script
after changing what a function takes or returns.

`input` and `output` give the structure a function takes and returns
("none" when neither side is one of these shapes; two structures joined
by `+` when arguments differ). Rows are sorted by `output`. Expand a row
for each argument's name and the element type (`matrix<character>`), and
the source file.

A `list of scalars` is a list whose elements all
have length 1. It covers two different things: one string per array (the
space-separated repeat codes `.build_repeat_msa()` and `as_tales()`
take), and records of file paths or settings (`tell_tales()`'s internal
`paths` and `params`).

```{r shapes}
#| code-summary: "Code: matrix and list-of-vectors table"
shapes_file <- "function-graph-shapes.tsv"
shapes <- read.delim(shapes_file, comment.char = "#", colClasses = "character")
shapes_recorded_on <- sub("^# ", "", readLines(shapes_file, n = 1))
never_called <- shapes$fn[shapes$direction == "never called"]
shapes <- shapes[shapes$direction %in% c("in", "out"), ]
# Where an argument was a list of vectors in some calls and a list of
# scalars in others, the scalar calls are short instances of the same
# thing and are not listed separately.
slot <- paste(shapes$fn, shapes$direction, shapes$arg)
short <- startsWith(shapes$shape, "list of scalars") &
  slot %in% slot[startsWith(shapes$shape, "list of vectors")]
shapes <- shapes[!short, ]

describe_side <- function(d) {
  if (!nrow(d)) return("")
  # one entry per argument, its shapes pooled over calls
  by_arg <- split(d$shape, d$arg)
  paste(vapply(names(by_arg), function(a) {
    s <- paste(sort(unique(by_arg[[a]])), collapse = " | ")
    if (a == "<value>") s else paste0(a, ": ", s)
  }, ""), collapse = "; ")
}
# The structure alone, without element types: "matrix<double>" -> "matrix".
structure_of <- function(x) {
  x <- unique(sub("<.*$", "", x))
  if (length(x)) paste(sort(x), collapse = " + ") else "none"
}
shape_tab <- do.call(rbind, lapply(split(shapes, shapes$fn), function(d) {
  ins <- d[d$direction == "in", ]
  outs <- d[d$direction == "out", ]
  data.frame(input = structure_of(ins$shape), output = structure_of(outs$shape),
             function_name = d$fn[1],
             takes = describe_side(ins), returns = describe_side(outs))
}))
shape_tab <- merge(shape_tab, defs[, c("name", "kind", "file")],
                   by.x = "function_name", by.y = "name", all.x = TRUE)
shape_tab$kind[is.na(shape_tab$kind)] <- "not in R/ any more"
shape_tab$`exported?` <- c(exported = "yes", `S3 method` = "S3 method",
                           internal = "no")[shape_tab$kind]
shape_tab$`exported?`[is.na(shape_tab$`exported?`)] <- shape_tab$kind[is.na(shape_tab$`exported?`)]
# Sorted by output, then input, then name; "none" after every shape.
shape_tab <- shape_tab[order(shape_tab$output == "none", shape_tab$output,
                             shape_tab$input == "none", shape_tab$input,
                             shape_tab$function_name),
                       c("input", "output", "exported?", "function_name",
                         "takes", "returns", "file", "kind")]
rownames(shape_tab) <- NULL
mono <- list(fontFamily = "monospace")
reactable(shape_tab, filterable = TRUE, pagination = FALSE, compact = TRUE,
          striped = TRUE,
          columns = list(input = colDef(minWidth = 170),
                         output = colDef(minWidth = 150),
                         `exported?` = colDef(width = 100),
                         function_name = colDef(minWidth = 220, style = mono),
                         takes = colDef(show = FALSE), returns = colDef(show = FALSE),
                         file = colDef(show = FALSE), kind = colDef(show = FALSE)),
          details = function(i) {
            r <- shape_tab[i, ]
            htmltools::div(
              style = "padding: 4px 12px; font-size: 0.9em",
              htmltools::div(htmltools::strong("takes: "),
                             htmltools::code(if (nzchar(r$takes)) r$takes else "none")),
              htmltools::div(htmltools::strong("returns: "),
                             htmltools::code(if (nzchar(r$returns)) r$returns else "none")),
              htmltools::div(htmltools::strong("file: "), r$file))
          })
```

`r shapes_recorded_on`.
`r nrow(shape_tab)` functions take or return one of these shapes:
`r sum(shape_tab$kind == "exported")` exported,
`r sum(shape_tab$kind == "S3 method")` S3 methods and
`r sum(shape_tab$kind == "internal")` internal.

**Not covered by the recording.** The test suite never calls the
`r length(never_called)` functions below, so their shapes are unknown;
some of them may take or return a matrix or a list of vectors.

```{r shapes-never-called}
#| code-summary: "Code: functions the tests never call"
unseen <- defs[defs$name %in% never_called, c("name", "kind", "file", "title")]
unseen <- unseen[order(unseen$kind != "exported", unseen$file, unseen$name), ]
if (nrow(unseen)) {
  reactable(unseen, pagination = FALSE, compact = TRUE,
            columns = list(name = colDef(minWidth = 190, style = list(fontFamily = "monospace")),
                           title = colDef(minWidth = 300)))
} else {
  cat("None.")
}
```

## Internal functions with no caller in `R/`

Exported functions and S3 methods are entry points by design. An internal
function that nothing in `R/` calls may be called only from tests, called
by R itself (`.onAttach()`), called in a way this parser does not see (see
below), or unused. The `in_tests` column says whether the name appears
anywhere under `tests/`. This list is a starting point for checking. Under
the standing rule, unused code is parked in `R/unused_pending_review.R` or
`inst/legacy/`, never deleted.

```{r orphans}
#| code-summary: "Code: uncalled internals"
orphans <- defs[defs$kind == "internal" & defs$callers == 0,
                c("name", "file", "line1", "title")]
names(orphans)[3] <- "line"
test_src <- unlist(lapply(list.files(file.path("..", "tests"), pattern = "\\.[Rr]$",
                                     recursive = TRUE, full.names = TRUE), readLines))
orphans$in_tests <- vapply(orphans$name, function(x) any(grepl(x, test_src, fixed = TRUE)),
                           logical(1))
if (nrow(orphans)) {
  reactable(orphans, pagination = FALSE, compact = TRUE,
            columns = list(name = colDef(style = list(fontFamily = "monospace"))))
} else {
  cat("None.")
}
```

## What the parser sees and misses

The graph comes from static reading of the source, so it has blind spots.

- **Included:** top-level `name <- function(...)` definitions; direct calls,
  including `tantale::name()` and the package's own infix operator
  `%||%`; functions passed by name as arguments; S3
  dispatch from `as_tales()` (the one generic the package defines) to its
  methods.
- **Methods of base and dplyr generics** (`print.tales`, `[.tales`,
  `dplyr_reconstruct.tales`, ...) show no callers. R reaches them when some
  code calls `print(x)` or `x[i]` on an object of that class, which the
  parser cannot tell from any other `print()` or `[`.
- **Calls built from strings** (`do.call("name", ...)`, `match.fun()`) are
  not followed. None target a package function at the time of writing.
- **Top-level constants** (`TALES_KEY_COLS`, `PAIRWISE_DISTANCES_ID_COLS`,
  ...) and the functions reading them are left out.
- **Class objects** (`tales`, `tales_msa`, `pairwise_distances` and its
  subclasses) as nodes, with edges for which function produces or consumes
  which, are phase 2 of §29 and not built yet.
