---
title: "1. Mining TALE sequences in genomes with tantale"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{1. Mining TALE sequences in genomes with tantale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r package, warning=FALSE, message=FALSE}
library(tantale)
library(dplyr)
```

This is the first of four articles walking through a typical study of TALE
diversity: finding TALE genes in a genome (here), classifying the arrays
found into groups of related sequences, aligning those groups, and
predicting the DNA targets of the TALEs they contain.

This article covers TALE discovery with `tell_tales()`, and something
discovery cannot avoid: real assemblies are not always clean, `tell_tales()`
can sometimes be fooled, and the two ways tantale offers to fix that --
correcting *inside* `tell_tales()` versus correcting the genome sequence
*beforehand* -- are not interchangeable, and neither is guaranteed to work.
`tales_anomalies()` is the tool that tells you which.

Four *Xanthomonas oryzae* genomes are used throughout this article: MAI1,
BAI3 and PXO86 are "clean" assemblies, and BAI3-1-1 is a deliberately
error-prone one -- an assembly of the same BAI3 background with *talC*
deleted, carrying real sequencing/assembly artefacts in its TALE loci. That
second kind is not a corner case: it is exactly the kind of input a mining
pipeline has to cope with in practice.

# Finding TALE loci in genomic DNA

`tell_tales()` searches a genome with `nhmmer`, using three profile HMMs
tuned to the N-terminus, the repeat unit, and the C-terminus of a TALE CDS.
Hits are merged, grouped into candidate arrays by proximity, and each
array's longest ORF is handed to AnnoTALE to split into parts and call its
RVD sequence. `NTERM`/`CTERM` markers are appended to the RVD string
wherever a terminus was identified this way, so a downstream alignment
knows where an array actually starts and ends.

```{r discover_mai1, message = FALSE, warning = FALSE}
out <- fs::dir_create(file.path(tempdir(), "tale_mining"))

mai1_fa <- system.file("extdata", "MAI1.fa", package = "tantale", mustWork = TRUE)
mai1_dir <- file.path(out, "MAI1")

invisible(tell_tales(subject_file = mai1_fa, output_dir = mai1_dir))
```

The result on disk is a directory of reports and fasta files; the result in
R is a `tales` object, one row per part:

```{r tales_from_mai1}
mai1 <- suppressWarnings(tales_from_telltale(mai1_dir))
mai1
```

`plot()` on a `tales` gives a compact overview of what was found -- every
array's parts, positioned within the array, coloured by domain type and
filled by amino-acid length, with the RVD printed on each repeat:

```{r plot_mai1, fig.width=9, fig.height=4}
plot(mai1)
```

## What `tell_tales()` writes to disk

Besides the RVD sequences and the TALE ORFs themselves, `tell_tales()`
writes three tab-separated reports, which `tales_from_telltale()` reads
back to build the object above. They are worth knowing about directly, since
they carry a few things the `tales` object does not:

| file | one row per | worth knowing |
|---|---|---|
| `hitsReport.tsv` | raw `nhmmer` hit | `codon_count`, `frameshift_count` per hit -- before any merging or correction |
| `domainsReport.tsv` | domain, after AnnoTALE | what AnnoTALE actually parsed out of the ORF |
| `arrayReport.tsv` | candidate array | `has_all_domains`, `has_aberrant_repeat`, `orf_coverage`, and (with correction) `predicted_ins_count`/`predicted_dels_count` |

```{r array_report}
array_report <- readr::read_tsv(file.path(mai1_dir, "arrayReport.tsv"),
                                show_col_types = FALSE)
array_report %>% select(array_id, has_all_domains, has_aberrant_repeat,
                        longest_orf_length, orf_coverage)
```

`orf_coverage` is the longest ORF found as a percentage of the whole
candidate array region -- a clean array covers nearly all of it. That
number is about to matter a great deal.

# When discovery goes wrong: detecting anomalies

MAI1 is a good assembly, and every array above came back complete. That is
not guaranteed. Running the same search on BAI3-1-1 -- deliberately
error-prone -- without any correction shows what an assembly artefact does
to the pipeline:

```{r discover_bai311_raw, message = FALSE, warning = FALSE}
bai311_fa <- system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE)
bai311_raw_dir <- file.path(out, "BAI3-1-1_raw")

invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_raw_dir,
                     cterm_min_score = 300))
bai311_raw <- suppressWarnings(tales_from_telltale(bai311_raw_dir))
```

`tales_anomalies()` reports the arrays that are odd rather than merely
absent -- missing sequence data, impossible terminus arrangements,
coordinate disagreements:

```{r anomalies_raw}
tales_anomalies(bai311_raw)
```

Two arrays have no `rvd` at all: AnnoTALE could not parse a repeat-array
structure out of their ORF. This is a frameshift, not a truncation --
compare their `orf_coverage` to a complete array:

```{r orf_coverage_raw}
readr::read_tsv(file.path(bai311_raw_dir, "arrayReport.tsv"),
                show_col_types = FALSE) %>%
  select(array_id, has_all_domains, longest_orf_length, orf_coverage)
```

`ROI_00003` and `ROI_00005` cover barely a fifth and a quarter of their
candidate region -- a single inserted or deleted base early in the array
shifts every codon downstream of it, and the ORF finder simply stops at
the first premature stop codon it meets. Nothing here is a bug: this is
exactly the situation frameshift correction exists for.

# Correcting frameshifts, two ways

tantale offers two distinct routes to a corrected sequence, and they are
not interchangeable: one corrects candidate *arrays* after they have
already been found, from inside `tell_tales()`; the other corrects the
*genome* before discovery even starts. Both can fail, in different ways,
which is exactly why the check above is worth repeating after either one.

## Correcting inside `tell_tales()`

`correct_array = TRUE` runs `DECIPHER::CorrectFrameshifts()` on each
candidate array, aligning it against a shipped reference set of real TALE
proteins and inferring the indels needed to remove premature stops.
`max_comparisons` caps how many references each array may be aligned
against, and is the main control on how long this takes -- full docs on
`tell_tales()` measure the trade-off directly: at the default (all
references) a set of four arrays took 252 seconds; capped to 20, 10
seconds -- but at a cap of 2, the aligner cannot reach a good reference and
invents indels wholesale. Capped to 20 is normally ample; it is used here
to keep this article fast to build:

```{r correct_inside, message = FALSE, warning = FALSE}
bai311_corr_dir <- file.path(out, "BAI3-1-1_corrected")

invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_corr_dir,
                     cterm_min_score = 300,
                     correct_array = TRUE, max_comparisons = 20))
bai311_corr <- suppressWarnings(tales_from_telltale(bai311_corr_dir))
tales_anomalies(bai311_corr)
```

Correction did fix the two frameshifted arrays -- `ROI_00003` and
`ROI_00005` are no longer flagged. But a *different* array, `ROI_00001`,
now is: it was clean in the uncorrected run above, and correcting against
only 20 references was, for this particular array, not enough to find a
good match. The correction did not fail by leaving the array alone; it
failed by aligning it against a poor reference, which is worse, because
the output still looks like a corrected ORF right up until AnnoTALE tries
to parse domains out of it and cannot.

Spending the full search on this same genome -- all ~500 shipped
references rather than 20, about eight minutes instead of ten seconds --
clears every anomaly, `ROI_00001` included, because the aligner can then
actually reach a reference close enough to it. There is no shortcut around
that trade-off: a small `max_comparisons` is a real risk, not a rounding
error, and checking `tales_anomalies()` after correcting is how that risk
gets caught rather than silently shipped downstream.

## Correcting the genome, before discovery

`correct_tales()` takes a different, faster route: rather than aligning
already-found candidate arrays against reference proteins, it wraps the
Java `TALEcorrection` tool to scan the *whole input genome* directly against
profile HMMs and repair frameshifts it finds, before `tell_tales()` ever
runs. That makes it a genome-wide pre-processing step rather than a
per-array one -- worth trying first on a large or especially error-prone
assembly (an ONT-sequenced genome, say), precisely because it does not need
candidate arrays to already exist:

```{r correct_outside, message = FALSE, warning = FALSE}
bai311_java_fa <- file.path(out, "BAI3-1-1_java_corrected.fa")

corrections <- correct_tales(uncorrected_path = bai311_fa,
                             corrected_path = bai311_java_fa,
                             return_corrections = TRUE)
nrow(corrections)
head(corrections, 4)
```

`tell_tales()` then runs on the corrected genome exactly as it would on any
other -- no `correct_array` needed, because the correction already happened
upstream:

```{r discover_after_java_correct, message = FALSE, warning = FALSE}
bai311_java_dir <- file.path(out, "BAI3-1-1_java_tell_tales")

invisible(tell_tales(subject_file = bai311_java_fa, output_dir = bai311_java_dir,
                     cterm_min_score = 300, correct_array = FALSE))
bai311_java <- suppressWarnings(tales_from_telltale(bai311_java_dir))
tales_anomalies(bai311_java)
```

A third outcome again: `ROI_00003` is fixed, but `ROI_00005` still is not
-- the opposite pattern from the `max_comparisons = 20` run above, which
fixed both original frameshifts but broke a third array instead. Neither
correction path is a silver bullet, and which one a given array needs is
not predictable in advance. What is true of both is that `tales_anomalies()`
is the right tool to check afterwards, regardless of which correction path
was used, or whether one was used at all.

## Moving on with what you have

Whichever route was taken, `tales(x, sanitize = TRUE)` drops whatever is
still flagged so an analysis can proceed on what is clean, while keeping a
clear record of what and why:

```{r sanitize}
bai311_clean <- tales(bai311_java, sanitize = TRUE)
n_distinct(bai311_java$array_id) - n_distinct(bai311_clean$array_id)
```

# Next

The next article picks up from a set of `tales` objects like the ones
built here -- across several genomes this time -- and classifies their
arrays into groups of related sequences with `tales_compare()` and
`tales_group()`.
