---
title: "2. Classifying TALE sequences from genomes with tantale"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{2. Classifying TALE sequences from genomes with tantale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r render, include=FALSE, eval = FALSE}
rmarkdown::render(input = "/home/cunnac/Lab-Related/MyScripts/tantale/vignettes/2_tale_classification.Rmd")

pkgdown::build_site(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    lazy = FALSE,
                    )

pkgdown::build_article(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    name = "2_tale_classification",
                    quiet = FALSE
                    )


```



```{r package, warning=FALSE, message=FALSE}
library(tantale)
library(ggplot2)
library(tidyverse)
library(Biostrings)
library(parallel)
library(magrittr)
library(DT)
library(gplots)
library(ape)
library(ggtree)
library(corrr)
library(ggcorrplot)
```

Based on the results of the first article "Mining TALE sequences in genomes with tantale", this second section will cover steps downstream of *tal* loci prediction. It will illustrate TALE classification into groups of related sequences with functions in package `tantale`.\
As a reminder, the genomes used for this article are MAI1, BAI3, BAI3-1-1 (an error-prone genome with TalC deleted) and PXO86, an outgroup Asian strain.\


```{r load previous session}
outdir <- fs::dir_create("~/TEMP/test_tantale") # tempdir(check = TRUE)
load(file.path(outdir, "mining.RData"))
```


# Classify the various TALE sequences into groups of related sequences

To classify TALE groups, we use the function `tales_compare()`, an R implementation of the original DisTal Perl program (see below), which outputs similarity between TALEs and similarity between repeats. The TALE similarity table can be taken as input of `tales_group()` for clustering groups, and the repeats similarity can be used for plotting multiple alignment of Tals within group.

> `tales_compare()` superseded the original Perl-based `runDistal()`/`buildDisTalGroups()` functions, which have been removed from the package (along with the ~90MB of bundled Perl modules they required). Before removal, their output was benchmarked against `tales_compare()`'s three pairwise-alignment backends and found to correlate highly (see the archived comparison in an earlier version of this vignette, or git history for `R/distal.R`).

## Run the R implementation of DisTal

First we use the tell_tales output directories for each genome and get the information about TALE parts (ie domains). The final output is a tibble with a single part per row.

```{r get tale parts, fig.width = 10, fig.height=10}
# distalrOutDir <- file.path(outdir, "distalr")
# unlink(distalrOutDir, recursive = TRUE)
# dir.create(distalrOutDir, showWarnings = F)


tale_parts <- lapply(list.dirs(telltale_outdir, recursive = FALSE), function(ttGenomeDir) {
  genomeId <- gsub("_.*correction", "", basename(ttGenomeDir))
  logger::log_debug("Current genome: {genomeId}")
  tale_parts <- tales_from_telltale(ttGenomeDir) %>%
    mutate(array_id = paste0(genomeId, "_", array_id))
  return(tale_parts)
}) %>%
  bind_rows()

DT::datatable(head(tale_parts), caption = "An example of the content of the tale_parts tibble.")


```

With this kind of representation it is quite easy to make summary plots like this one for example:


```{r Overview of TALE composition by genome, fig.width = 10, fig.height=10}
partsForPlots <- tale_parts %>%
  mutate(label = if_else(domain_type == "repeat", rvd, ""),
         aa_seq_length = factor(nchar(aa_seq))
  )
partsForPlots %>% ggplot(mapping = aes(fill = aa_seq_length,
                                       color = domain_type,
                                       label = label,
                                       y = array_id,
                                       x = position_in_array)) +
  scale_color_viridis_d(option = "rocket") +
  scale_fill_discrete() +
  scale_x_continuous(breaks = 1:50, minor_breaks = NULL) +
  geom_point(shape = 21, size = 5, stroke = 0.9) +
  ggnewscale::new_scale_color() +
  ggnewscale::new_scale_fill() +
  geom_text(size = 2.1, color = "white") + 
  facet_grid(seqnames~ ., scales = "free_y", space = "free") +
  labs(title = "Overview of TALE composition by genome") +
  theme_light()
```

Now, with this `tibble`, we can also run the R implementation of the Distal Perl code.


```{r doing the same with tales_compare, fig.width = 10, fig.height=10}
partsForDistalr <- tale_parts

system.time({distalr_bs_output <- tales_compare(partsForDistalr, ncores = 6,
                             aln_method = "Biostrings")})

system.time({distalr_deci_output <- tales_compare(partsForDistalr, ncores = 6,
                              aln_method = "DECIPHER")})

system.time({distalr_mms_output <- tales_compare(partsForDistalr, ncores = 6,
                              aln_method = "mmseq2",
                              conda_bin = "/home/cunnac/bin/miniconda3/condabin/conda")})

```

Clearly, the "DECIPHER" method for pairwise alignment of domains aa sequences is much faster than all the others.



## Contrast the output of the various distal functions


Now, we can take a look at whether the results of the original Distal script are consistent with our R implementation of the program.


```{r correlation of TALE similarity from various distal implementations, fig.width = 4, fig.height=4}
talSimOutAgregated <- list(distalDECI = distalr_deci_output$tale_distances,
     distalBIOS = distalr_bs_output$tale_distances,
     distalMMS = distalr_mms_output$tale_distances
) %>% bind_rows(.id = "method")

talSimOutAgregatedLong <- reshape2::dcast(talSimOutAgregated %>% select(method, TAL1, TAL2, Sim),
                                          formula = method ~ TAL1 + TAL2, value.var = "Sim"
                                            )
rownames(talSimOutAgregatedLong) <- talSimOutAgregatedLong$method
talSimOutAgregatedLong$method <- NULL
talSimOutAgregatedLong %<>% as.matrix() %>% t()
rownames(talSimOutAgregatedLong) <- NULL

corr_matrix <- cor(talSimOutAgregatedLong)
knitr::kable(corr_matrix)
ggcorrplot(corr_matrix)


talSimOutAgregated <- reshape2::dcast(talSimOutAgregated %>% select(method, TAL1, TAL2, Sim),
                                          formula = TAL1 + TAL2 ~ method, value.var = "Sim"
                                            ) %>% as_tibble()
```


```{r pair plot of TALE similarity from various distal implementations, fig.width = 6, fig.height= 6}
pairs(talSimOutAgregated[3:5], pch = 20, cex = 0.5, lower.panel = NULL)
```



The correlation of the TALE pairwise similarity scores output by the various `tales_compare()` backends is remarkably high.

## Allocate TALEs to groups of related sequences

```{r tale trees, fig.width = 6, fig.height=6}
k <- 22
groupsDECI <- invisible(tales_group(tal_sim = distalr_deci_output$tale_distances,
                     plot_tree = TRUE, k = k, k_range = 10:40,
                     method = "k-medoids"))

groupsBIOS <- invisible(tales_group(tal_sim = distalr_bs_output$tale_distances,
                     plot_tree = TRUE, k = k, k_range = 10:40,
                     method = "k-medoids"))

groupsMMS <- invisible(tales_group(tal_sim = distalr_mms_output$tale_distances,
                     plot_tree = TRUE, k = k, k_range = 10:40,
                     method = "k-medoids"))


grp <- tales_group(tal_sim = distalr_deci_output$tale_distances,
                     plot_tree = TRUE, k = k,
                     method = "hclust")
grp <- tales_group(tal_sim = distalr_bs_output$tale_distances,
                     plot_tree = TRUE, k = k,
                     method = "hclust")
grp <- tales_group(tal_sim = distalr_mms_output$tale_distances,
                     plot_tree = TRUE, k = k,
                     method = "hclust")

```

Because PXO86 is the only Asian strain and these TALEs are rather distantly related to African ones, the part of the trees with PXO86 TALEs are not really meaningful in this context.


```{r test_tale_group}
taleGroups <- groupsDECI %>% as_tibble() %>% arrange(group, name)
knitr::kable(head(taleGroups, n= 12))
```

Based on the silhouette values, `tales_group()` will take the value at the elbow of the curve. In this case we manually took k = `r k`.



```{r save session}
save.image(file.path(outdir, "mining.RData"))
```


