---
title: "Getting started with tantale"
format:
  html:
    toc: true
    code-fold: show
vignette: >
  %\VignetteIndexEntry{Getting started with tantale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

A TALE (Transcription Activator-Like Effector) is a modular protein: a central array of near-identical ~34-residue repeats,
each one specifying a single DNA base through two variable residues (its
RVD), flanked by an N- and a C-terminal region. tantale's job is to find
these arrays in genomic sequence, and to represent them as data you can
subset, compare and plot like any other tibble, with one row per repeat
or terminus.

This article is a two-minute tour. It loads a small result already
sitting on disk, so there are no external tools to install first. The
[package website](https://scunnac.github.io/tantale) has the full set of
articles, which start from raw genomic FASTA and carry a handful of real
*Xanthomonas oryzae* genomes through discovery, classification, alignment
and target prediction.

```{r package, message = FALSE}
library(tantale)
```

## A `tales` object

`tell_tales()` searches a genome for TALE-coding regions and writes its
findings to a directory of reports and FASTA files;
`tales_from_telltale()` reads that directory back into R. The package
ships one such directory already computed (four arrays found across two
short input sequences), so this tour can start from the object and skip
a multi-minute search:

```{r load}
example_dir <- system.file("extdata", "tellTaleExampleOutput", package = "tantale")
x <- tales_from_telltale(example_dir)
x
```

Every row is one *part* of one TALE *array* (the domains composing the full protein, ie the *array*): a repeat or a terminus. `summary()`
gives the array-level view:

```{r summary}
summary(x)
```

## Checking for trouble

Real assemblies are not always clean: a frameshift can leave AnnoTALE
unable to parse a repeat structure out of a candidate array, or a terminus
can be truncated. `tales_anomalies()` reports which arrays, if any, need a
second look:

```{r anomalies}
tales_anomalies(x)
```

Zero rows means none did, here. The [mining
article](https://scunnac.github.io/tantale/articles/tale_mining.html)
covers what to do when this table is not empty -- tantale has two
different ways to correct a frameshifted array.

## Looking at the arrays

`plot()` on a `tales` object lays out every array's parts in order. The
outline of each circle gives the domain type, its fill the part's length
in amino acids, and each repeat carries its RVD. The last repeat of each
array is shorter (20 residues here): it is the half-repeat that ends every
TALE repeat region:

```{r}
#| label: fig-composition
#| fig-cap: "Domain composition of the four TALE arrays in the bundled example."
#| fig-width: 8
#| fig-height: 3
plot(x)
```

## Where to next

This is deliberately the smallest possible slice of tantale. From here,
the package website walks through the full pipeline on real genomes:

- [Mining TALE sequences in genomes](https://scunnac.github.io/tantale/articles/tale_mining.html) -- discovery from raw FASTA, and correcting the frameshifts real assemblies produce
- [The `tales` class](https://scunnac.github.io/tantale/articles/tales_class.html) -- what the object built above actually is, in depth
- [Genuine truncTALEs and frameshift correction](https://scunnac.github.io/tantale/articles/trunctale_correction.html) -- how the two frameshift-correction methods treat naturally truncated TALEs
- [Classifying TALE sequences from genomes](https://scunnac.github.io/tantale/articles/tale_classification.html) -- grouping arrays across several genomes by relatedness
- [Multiple alignment of TALE arrays](https://scunnac.github.io/tantale/articles/tale_msa.html) and [the `tales_msa` class](https://scunnac.github.io/tantale/articles/tales_msa_class.html)
- [Predicting TALE targets](https://scunnac.github.io/tantale/articles/tale_target_prediction.html) -- from an aligned group to a predicted DNA binding site

::: {.callout-note}
## These four articles form one analysis
Classification, both alignment articles, and target prediction all work
over the same three genomes and share one discovery/comparison/grouping
result. The classification article computes it and caches it in
`vignettes/articles/_cache/`, outside the installed package; the other
three read that cache. To render one of those three on its own from a
clean checkout, render the classification article first.
:::
