---
title: "Getting started with tantale"
format:
  html:
    toc: true
    code-fold: show
vignette: >
  %\VignetteIndexEntry{Getting started with tantale}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

A TALE (transcription activator-like effector) gene, once translated, is
a modular protein: a central array of near-identical ~34-residue repeats,
each one specifying a single DNA base through two variable residues (its
RVD), flanked by an N- and a C-terminal region. tantale's job is to find
these arrays in genomic sequence, and to represent them as data you can
subset, compare and plot like any other tibble -- one row per repeat or
terminus, not one opaque object per TALE.

This article is a two-minute tour, not a walkthrough: it loads a small
result already sitting on disk rather than running discovery live, so it
has no external tools to install first. The [package
website](https://scunnac.github.io/tantale) has the real thing -- a set of
articles that start from raw genomic FASTA and carry a handful of real
*Xanthomonas oryzae* genomes all the way through discovery, classification,
alignment and target prediction.

```{r package, message = FALSE}
library(tantale)
```

## A `tales` object

`tell_tales()` searches a genome for TALE-coding regions and writes its
findings to a directory of reports and FASTA files;
`tales_from_telltale()` reads that directory back into R. The package
ships one such directory already computed -- four arrays found across two
short input sequences -- so this tour can start from the object rather
than from a multi-minute search:

```{r load}
example_dir <- system.file("extdata", "tellTaleExampleOutput", package = "tantale")
x <- tales_from_telltale(example_dir)
x
```

Every row is one *part* of one array -- a repeat, or a terminus -- not one
row per TALE. `summary()` gives the array-level view:

```{r summary}
summary(x)
```

## Checking for trouble

Real assemblies are not always clean: a frameshift can leave AnnoTALE
unable to parse a repeat structure out of a candidate array, or a terminus
can go missing. `tales_anomalies()` reports which arrays, if any, need a
second look:

```{r anomalies}
tales_anomalies(x)
```

Zero rows means none did, here. The [mining
article](https://scunnac.github.io/tantale/articles/tale_mining.html)
covers what to do when this table is not empty -- tantale has two
different ways to correct a frameshifted array, and they do not always
agree with each other.

## Looking at the arrays

`plot()` on a `tales` object lays out every array's parts in order,
coloured by domain type, with each repeat's RVD printed on it:

```{r}
#| label: fig-composition
#| fig-cap: "Domain composition of the four TALE arrays in the bundled example."
#| fig-width: 8
#| fig-height: 3
plot(x)
```

## Where to next

This is deliberately the smallest possible slice of tantale. From here,
the package website walks through the full pipeline on real genomes:

- [Mining TALE sequences in genomes](https://scunnac.github.io/tantale/articles/tale_mining.html) -- discovery from raw FASTA, and correcting the frameshifts real assemblies produce
- [The `tales` class](https://scunnac.github.io/tantale/articles/tales_class.html) -- what the object built above actually is, in depth
- [Classifying TALE sequences from genomes](https://scunnac.github.io/tantale/articles/tale_classification.html) -- grouping arrays across several genomes by relatedness
- [Multiple alignment of TALE arrays](https://scunnac.github.io/tantale/articles/tale_msa.html) and [the `tales_msa` class](https://scunnac.github.io/tantale/articles/tales_msa_class.html)
- [Predicting TALE targets](https://scunnac.github.io/tantale/articles/tale_target_prediction.html) -- from an aligned group to a predicted DNA binding site

::: {.callout-note}
## These four articles are one analysis, not four
Classification, both alignment articles, and target prediction all work
over the same three genomes, and the discovery/comparison/grouping result
is shared rather than recomputed four times: the classification article
computes it and caches the result (`vignettes/articles/_cache/`, not part
of the installed package); the other three read that cache. If you are
rendering one of those three in isolation from a clean checkout, rather
than through a full site build, render the classification article first.
:::
