---
title: "Comparing `tales_compare`'s pairwise-alignment backends"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

Here is a small test set of tale cds sequences (5 = 2 groups of 2 + outlier)

```{r}
library(tantale)
tellTaleSubjSeqsFile <- system.file("extdata", "bai3_sample_tal_genomic_regions.fasta", package = "tantale", mustWork = T)
telltale_dir <- system.file("extdata", "tellTaleExampleOutput", package = "tantale", mustWork = T)
#
# tantale::tell_tales(subject_file = tellTaleSubjSeqsFile,
#                           output_dir = telltale_dir
# )
```

# Parse annotale output (rvd, aa, dna, domain type, array id) in a tell_tales output dir

```{r parse tale parts from tell_tales output dir}
tales <- tales_from_telltale(system.file("extdata", "tellTaleExampleOutput", package = "tantale", mustWork = T))
```


## Run `tales_compare` to investigate TALE protein domains composition relatedness

`tales_compare()` returns a list of three typed objects: the stamped `tales`,
the `domain_distances` between domains, and the `tale_distances` between
arrays. The distances are stored as distances -- self-comparison is 0.

```{r use tales_compare}
dres <- tales_compare(tales, ncores = 6, aln_method = "Biostrings")

# dres2 <- tales_compare(tales, ncores = 6, aln_method = "mmseq2",
#                        conda_bin = "/home/cunnac/bin/miniconda3/condabin/conda")

dres3 <- tales_compare(tales, ncores = 6, aln_method = "DECIPHER")
```

```{r compare the backends}
str(lapply(dres, class))

# the two backends should broadly agree on which arrays are close
biostringsD <- as.matrix(dres$tale_distances)
decipherD   <- as.matrix(dres3$tale_distances)
cor(as.vector(biostringsD), as.vector(decipherD[rownames(biostringsD), colnames(biostringsD)]))
```
