---
title: "TALEs domains sequences alignment with MAFFT"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r render, include=FALSE, eval = FALSE}
rmarkdown::render(input = "/home/cunnac/Lab-Related/MyScripts/tantale/vignettes/multiple_alignments.Rmd")

pkgdown::build_site(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    lazy = TRUE)

pkgdown::build_article(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    name = "4_multiple_alignments",
                    quiet = FALSE
                    )
```



```{r package, warning=FALSE, message=FALSE}
library(tidyverse)
library(tantale)

```

This demo will illustrate how to use the `buildRepeatMsa` function to produce TALE sequence multiple algnments.\

Additionally, using real life distalr output, it will details the various type of aesthetics that can be used to plot those TALEs msa.

# Producing simple TALE sequence multiple algnments

Here we start with RVD sequences of the African Xoo TALE TalA:

```{r align rvd strings}
aln <- buildRepeatMsa(inputSeqs = system.file("extdata", "TalA_RVDSeqs_AnnoTALE.fasta",
                                              package = "tantale", mustWork = TRUE),
                      sep = "-",
                      distalRepeatSims = NULL,
                      mafftOpts = "--localpair --maxiterate 1000 --quiet --reorder --op 0 --ep 5 --thread 1",
                      gapSymbol = "-")
```

The output shown as a simple matrix :

```{r kable rvd align}
knitr::kable(aln)
```

Now, it is more evolutionarily relevant to look at Distal parts/domain/repeat (this is used somehow interchangeably with roughly the same meaning) code sequences. Needless to say that to obtain TALE parts code sequences, you need beforehand to run either the original Distal perl script with `runDistal` or our R implementation `distalr` with DNA sequences. In this cas we just use a toy example provided with tantale.


```{r align parts strings}
aln <- buildRepeatMsa(inputSeqs = system.file("extdata", "small_Out_CodedRepeats.fa",
                                              package = "tantale", mustWork = TRUE),
                      sep = " ",
                      distalRepeatSims = NULL,
                      mafftOpts = "--localpair --maxiterate 1000 --quiet --reorder --op 0 --ep 5 --thread 1",
                      gapSymbol = "-")
```

**NOTE**, the difference in the `sep` parameter value.

The output shown as a simple matrix :

```{r kable parts align}
knitr::kable(aln)
```


# A detailled illustration of one of the TALE msa plotting functions

First we get the necessary info from object from a previous run:

```{r get info}
distalrOut <- readRDS(file = system.file("extdata", "sampleDistalrOutput.rds",
                                              package = "tantale", mustWork = TRUE))
repeatMsaByGroup <- readRDS(file = system.file("extdata", "sampleRepeatMsaByGroup.rds",
                                              package = "tantale", mustWork = TRUE))

repeatAlign <- repeatMsaByGroup[[4]]
rvdAlign <- convertRepeat2RvdAlign(repeatAlign = repeatAlign,
                                   repeat2RvdMapping = getRepeat2RvdMappingFromDistalr(distalrOut$taleParts))
```

This is what you get if you provide all the arguments:

```{r detailled illustration of ggplotTalesMsa, fig.width=12, fig.height=2.5}
p1 <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = distalrOut$tal.similarity,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = rvdAlign,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)
```

The returned object can be modified :

```{r dismanteling a aplot}
class(p1)
str(p1$plotlist, max.level = 1)
```

Plotting the original alignment

```{r The alignment part of the aplot, fig.width=12, fig.height=2.5}
p1$plotlist[[1]]
```

Now we can modify the color scales and replot to check this is what we want.

```{r The modified alignment part of the aplot, fig.width=12, fig.height=2.5}
cat("modifying the original alignment as a ggplot")
p2 <- p1$plotlist[[1]] + ggplot2::scale_fill_viridis_c() + ggplot2::scale_color_discrete()
p2
```

We can extract the tree part of the object.


```{r The tree part of the aplot, fig.width=2.5, fig.height=2.5}
p1$plotlist[[2]]
```

And we can recombine all again :

```{r reassembling align and tree, fig.width=12, fig.height=2.5}
p1$plotlist[[1]] <- p2
p1
```


Pay attention to the arguments provided to the plotting function.
In the various calls to `ggplotTalesMsa`, arguments are gradually removed to exemplify the type of alignments that are thus displayed.

```{r detailled illustration of ggplotTalesMsa others, fig.width=12, fig.height=2.5}
p <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = distalrOut$tal.similarity,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = rvdAlign,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatClust" # "repeatClust"
)

p <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = distalrOut$tal.similarity,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = rvdAlign,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)

p <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = distalrOut$tal.similarity,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = NULL,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatClust" # "repeatClust"
)

# fillType has no effect
p <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = distalrOut$tal.similarity,
               repeatSim = NULL,
               rvdAlign = rvdAlign,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)


p <- ggplotTalesMsa(repeatAlign = NULL,
               talsim = distalrOut$tal.similarity,
               repeatSim = NULL,
               rvdAlign = rvdAlign,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)


p <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = distalrOut$tal.similarity,
               repeatSim = NULL,
               rvdAlign = NULL,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)

p <- ggplotTalesMsa(repeatAlign = NULL,
               talsim = NULL,
               repeatSim = NULL,
               rvdAlign = rvdAlign,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)


p <- ggplotTalesMsa(repeatAlign = repeatAlign,
               talsim = NULL,
               repeatSim = NULL,
               rvdAlign = NULL,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)




# Single sequence align
p <- ggplotTalesMsa(repeatAlign = repeatAlign[3, , drop = FALSE],
               talsim = distalrOut$tal.similarity,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = rvdAlign[3, , drop = FALSE],
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)

p <- ggplotTalesMsa(repeatAlign = repeatAlign[3, , drop = FALSE],
               talsim = NULL,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = rvdAlign[3, , drop = FALSE],
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatClust" # "repeatClust"
)


p <- ggplotTalesMsa(repeatAlign = repeatAlign[3, , drop = FALSE],
               talsim = NULL,
               repeatSim = distalrOut$repeat.similarity,
               rvdAlign = NULL,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)

p <- ggplotTalesMsa(repeatAlign = repeatAlign[3, , drop = FALSE],
               talsim = NULL,
               repeatSim = NULL,
               rvdAlign = NULL,
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)

p <- ggplotTalesMsa(repeatAlign = NULL,
               talsim = NULL,
               repeatSim = NULL,
               rvdAlign = rvdAlign[3, , drop = FALSE],
               repeat.clust.h.cut = 90,
               refgrep = NULL,
               consensusSeq = FALSE,
               fillType = "repeatSim" # "repeatClust"
)
```


# A couple of utility functions for TALE msa

Compute the consensus of the columns in a TALE alignment:

```{r get consensus}
taleAlignConsensus(repeatAlign)
taleAlignConsensus(rvdAlign)
```

Figure out what elements in the TALE alignment corresponds to the consensus:

```{r match consensus}
matchConsensus(repeatAlign)
matchConsensus(repeatAlign, returnLong = FALSE)
```



