---
title: "TALEs domains sequences alignment with MAFFT"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{TALEs domains sequences alignment with MAFFT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r render, include=FALSE, eval = FALSE}
rmarkdown::render(input = "/home/cunnac/Lab-Related/MyScripts/tantale/vignettes/multiple_alignments.Rmd")

pkgdown::build_site(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    lazy = TRUE)

pkgdown::build_article(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    name = "p3_tantale_objects",
                    quiet = FALSE
                    )
```



```{r package, warning=FALSE, message=FALSE}
library(tidyverse)
library(tantale)

```

This demo will illustrate how to use the `tales_align()` function to produce TALE sequence multiple algnments.\

Additionally, using real life `tales_compare()` output, it will details the various type of aesthetics that can be used to plot those TALEs msa.

# Producing simple TALE sequence multiple algnments

Here we start with RVD sequences of the African Xoo TALE TalA:

```{r align rvd strings}
# A fasta reaches an alignment by composing the two steps: as_tales() ingests
# it, tales_align() aligns it. as.matrix() then materialises the grid.
tales <- as_tales(system.file("extdata", "TalA_RVDSeqs_AnnoTALE.fasta",
                              package = "tantale", mustWork = TRUE),
                  sep = "-", residue_col = "rvd")
rvd_msa <- tales_align(tales, residue_col = "rvd",
                   mafft_opts = "--localpair --maxiterate 1000 --quiet --reorder --op 0 --ep 5 --thread 1")
rvd_aln <- as.matrix(rvd_msa, gap = "-")
```

The output shown as a simple matrix :

```{r kable rvd align}
knitr::kable(rvd_aln)
```

Now, it is more evolutionarily relevant to look at Distal parts/domain/repeat (this is used somehow interchangeably with roughly the same meaning) code sequences. Needless to say that to obtain TALE parts code sequences, you need beforehand to run `tales_compare()` with DNA sequences. In this cas we just use a toy example provided with tantale.


```{r align parts strings}
tales <- as_tales(system.file("extdata", "small_Out_CodedRepeats.fa",
                              package = "tantale", mustWork = TRUE),
                  sep = " ", residue_col = "dom_code")
repeat_msa <- tales_align(tales, residue_col = "dom_code",
                   mafft_opts = "--localpair --maxiterate 1000 --quiet --reorder --op 0 --ep 5 --thread 1")
repeat_aln <- as.matrix(repeat_msa)
```

**NOTE**, the difference in the `sep` parameter value.

The output shown as a simple matrix :

```{r kable parts align}
knitr::kable(repeat_aln)
```


# A couple of utility functions for TALE msa

Both matrices above -- an RVD alignment and a repeat-code alignment --
work the same way with the two utilities below, since neither cares which
residue vocabulary it is given.

Compute the consensus of the columns in a TALE alignment:

```{r get consensus}
tales_consensus(rvd_aln)
tales_consensus(repeat_aln)
```

Figure out what elements in the TALE alignment correspond to the consensus:

```{r match consensus}
tales_consensus_match(repeat_aln)
tales_consensus_match(repeat_aln, long = FALSE)
```

# Plotting an alignment

Rendering a `tales_msa` as a heatmap -- cell colour, cell text, and the
tree and consensus panels that can sit alongside it -- is `plot()`, via
`plot.tales_msa()`. It has enough independent arguments (which layer fills
the cells, which labels them, whether a dendrogram or a consensus row is
attached) to deserve its own worked treatment rather than a condensed one
here: see the
[`tales_msa` class article](../articles/tales-msa-class.html) for that,
built against the same kind of small, real, `tell_tales()`-derived
alignment as this vignette.



