---
title: "TALEs domains sequences alignment with MAFFT"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{TALEs domains sequences alignment with MAFFT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r render, include=FALSE, eval = FALSE}
rmarkdown::render(input = "/home/cunnac/Lab-Related/MyScripts/tantale/vignettes/multiple_alignments.Rmd")

pkgdown::build_site(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    lazy = TRUE)

pkgdown::build_article(pkg = "/home/cunnac/Lab-Related/MyScripts/tantale",
                    name = "p3_tantale_objects",
                    quiet = FALSE
                    )
```



```{r package, warning=FALSE, message=FALSE}
library(tidyverse)
library(tantale)

```

This demo will illustrate how to use the `tales_align()` function to produce TALE sequence multiple algnments.\

Additionally, using real life `tales_compare()` output, it will details the various type of aesthetics that can be used to plot those TALEs msa.

# Producing simple TALE sequence multiple algnments

Here we start with RVD sequences of the African Xoo TALE TalA:

```{r align rvd strings}
# A fasta reaches an alignment by composing the two steps: as_tales() ingests
# it, tales_align() aligns it. as.matrix() then materialises the grid.
tales <- as_tales(system.file("extdata", "TalA_RVDSeqs_AnnoTALE.fasta",
                              package = "tantale", mustWork = TRUE),
                  sep = "-", residue_col = "rvd")
msa <- tales_align(tales, residue_col = "rvd",
                   mafft_opts = "--localpair --maxiterate 1000 --quiet --reorder --op 0 --ep 5 --thread 1")
aln <- as.matrix(msa, gap = "-")
```

The output shown as a simple matrix :

```{r kable rvd align}
knitr::kable(aln)
```

Now, it is more evolutionarily relevant to look at Distal parts/domain/repeat (this is used somehow interchangeably with roughly the same meaning) code sequences. Needless to say that to obtain TALE parts code sequences, you need beforehand to run `tales_compare()` with DNA sequences. In this cas we just use a toy example provided with tantale.


```{r align parts strings}
tales <- as_tales(system.file("extdata", "small_Out_CodedRepeats.fa",
                              package = "tantale", mustWork = TRUE),
                  sep = " ", residue_col = "dom_code")
msa <- tales_align(tales, residue_col = "dom_code",
                   mafft_opts = "--localpair --maxiterate 1000 --quiet --reorder --op 0 --ep 5 --thread 1")
aln <- as.matrix(msa, gap = "-")
```

**NOTE**, the difference in the `sep` parameter value.

The output shown as a simple matrix :

```{r kable parts align}
knitr::kable(aln)
```


# A detailled illustration of one of the TALE msa plotting functions

First we get the necessary info from object from a previous run:

```{r get info}
distalrOut <- readRDS(file = system.file("extdata", "sampleDistalrOutput.rds",
                                              package = "tantale", mustWork = TRUE))
repeatMsaByGroup <- readRDS(file = system.file("extdata", "sampleRepeatMsaByGroup.rds",
                                              package = "tantale", mustWork = TRUE))

repeat_align <- repeatMsaByGroup[[4]]
rvd_align <- repeat_to_rvd_align(repeat_align = repeat_align,
                                   rvd_map = repeat_to_rvd_map_distalr(distalrOut$tale_parts))
```

This is what you get if you provide all the arguments:

```{r detailled illustration of plot_tales_msa, fig.width=12, fig.height=2.5}
p1 <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = rvd_align,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)
```

The returned object can be modified :

```{r dismanteling a aplot}
class(p1)
str(p1$plotlist, max.level = 1)
```

Plotting the original alignment

```{r The alignment part of the aplot, fig.width=12, fig.height=2.5}
p1$plotlist[[1]]
```

Now we can modify the color scales and replot to check this is what we want.

```{r The modified alignment part of the aplot, fig.width=12, fig.height=2.5}
cat("modifying the original alignment as a ggplot")
p2 <- p1$plotlist[[1]] + ggplot2::scale_fill_viridis_c() + ggplot2::scale_color_discrete()
p2
```

We can extract the tree part of the object.


```{r The tree part of the aplot, fig.width=2.5, fig.height=2.5}
p1$plotlist[[2]]
```

And we can recombine all again :

```{r reassembling align and tree, fig.width=12, fig.height=2.5}
p1$plotlist[[1]] <- p2
p1
```


Pay attention to the arguments provided to the plotting function.
In the various calls to `plot_tales_msa`, arguments are gradually removed to exemplify the type of alignments that are thus displayed.

```{r detailled illustration of plot_tales_msa others, fig.width=12, fig.height=2.5}
p <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = rvd_align,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_clust" # "repeat_clust"
)

p <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = rvd_align,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)

p <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = NULL,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_clust" # "repeat_clust"
)

# fill_type has no effect
p <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = NULL,
               rvd_align = rvd_align,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)


p <- plot_tales_msa(repeat_align = NULL,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = NULL,
               rvd_align = rvd_align,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)


p <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = NULL,
               rvd_align = NULL,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)

p <- plot_tales_msa(repeat_align = NULL,
               tal_sim = NULL,
               repeat_sim = NULL,
               rvd_align = rvd_align,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)


p <- plot_tales_msa(repeat_align = repeat_align,
               tal_sim = NULL,
               repeat_sim = NULL,
               rvd_align = NULL,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)




# Single sequence align
p <- plot_tales_msa(repeat_align = repeat_align[3, , drop = FALSE],
               tal_sim = distalrOut$tal.similarity,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = rvd_align[3, , drop = FALSE],
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)

p <- plot_tales_msa(repeat_align = repeat_align[3, , drop = FALSE],
               tal_sim = NULL,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = rvd_align[3, , drop = FALSE],
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_clust" # "repeat_clust"
)


p <- plot_tales_msa(repeat_align = repeat_align[3, , drop = FALSE],
               tal_sim = NULL,
               repeat_sim = distalrOut$repeat.similarity,
               rvd_align = NULL,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)

p <- plot_tales_msa(repeat_align = repeat_align[3, , drop = FALSE],
               tal_sim = NULL,
               repeat_sim = NULL,
               rvd_align = NULL,
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)

p <- plot_tales_msa(repeat_align = NULL,
               tal_sim = NULL,
               repeat_sim = NULL,
               rvd_align = rvd_align[3, , drop = FALSE],
               h_cut = 90,
               ref_pattern = NULL,
               consensus = FALSE,
               fill_type = "repeat_sim" # "repeat_clust"
)
```


# A couple of utility functions for TALE msa

Compute the consensus of the columns in a TALE alignment:

```{r get consensus}
tales_consensus(repeat_align)
tales_consensus(rvd_align)
```

Figure out what elements in the TALE alignment corresponds to the consensus:

```{r match consensus}
tales_consensus_match(repeat_align)
tales_consensus_match(repeat_align, long = FALSE)
```



