---
title: "Showcase functions for predicting TALE DNA-binding sites on target sequences"
author: "Seb"
date: "`r Sys.time()`"
output:
  html_document:
    theme: cerulean
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tantale)
library(magrittr)
```

# Common parameters values for testing

```{r Common parameters}
rvdSeqsFile= system.file("extdata", "cladeIII_sweet_targeting_control_TALEs.fa", package = "tantale", mustWork = T)
rvdSeqsXstrings= Biostrings::readBStringSet(system.file("extdata", "cladeIII_sweet_targeting_control_TALEs.fa", package = "tantale", mustWork = T))
subjDnaSeqFile= system.file("extdata", "cladeIII_sweet_promoters.fasta", package = "tantale", mustWork = T)
```

# Testing latest version of Talvez (3.2) that can run searches on both strands

```{r Testing Talvez}
outDir = tempfile(); dir.create(outDir)

talvezPreds <- talvez(rvdSeqs = rvdSeqsFile, subjDnaSeqFile = subjDnaSeqFile, outDir = NULL)
talvezPreds <- talvez(rvdSeqs = rvdSeqsXstrings, subjDnaSeqFile = subjDnaSeqFile, optParam = "-t 1 -l 20", outDir = outDir)
talvezPreds <- talvez(rvdSeqs = rvdSeqsXstrings, subjDnaSeqFile = subjDnaSeqFile, optParam = "-t 0 -l 19", outDir = NULL)
knitr::kable(head(talvezPreds))
```

Checking that positions are correct:

```{r}
predsGr <- talvezPreds %>% #dplyr::filter(strand == "-") %>%
  GenomicRanges::makeGRangesFromDataFrame(seqnames.field = "subjSeqId", keep.extra.columns = TRUE)
stopifnot(as.character(BSgenome::getSeq(Biostrings::readDNAStringSet(subjDnaSeqFile), predsGr)) == predsGr$ebeSeq)
```


# Testing preditale

```{r Testing preditale}
outDir = tempfile(); dir.create(outDir)
optParam = "Strand=\"forward strand\""

preditalePreds <- preditale(rvdSeqs = rvdSeqsFile, subjDnaSeqFile = subjDnaSeqFile, optParam = optParam, outDir = outDir)
try(preditalePreds <- preditale(rvdSeqs = rvdSeqsFile, subjDnaSeqFile = subjDnaSeqFile, optParam = optParam, outDir = outDir))
preditalePreds <- preditale(rvdSeqs = rvdSeqsXstrings, subjDnaSeqFile = subjDnaSeqFile, outDir = NULL)
knitr::kable(head(preditalePreds))
```

Checking that positions are correct:

```{r check positions}
predsGr <- preditalePreds %>% #dplyr::filter(strand == "-") %>%
  GenomicRanges::makeGRangesFromDataFrame(seqnames.field = "subjSeqId", keep.extra.columns = TRUE)
stopifnot(as.character(BSgenome::getSeq(Biostrings::readDNAStringSet(subjDnaSeqFile), predsGr)) == predsGr$ebeSeq)
```



# Easily plot rvd seqs alingments to predicted EBEs along a single subject DNA sequence.


```{r testing plotTaleTargetPred, eval = TRUE}

talvezPreds <- talvez(rvdSeqs = rvdSeqsXstrings, subjDnaSeqFile = subjDnaSeqFile, optParam = "-t 10 -l 19", outDir = outDir)
preditalePreds <- preditale(rvdSeqs = rvdSeqsXstrings, subjDnaSeqFile = subjDnaSeqFile, outDir = NULL) %>% print(n=Inf)

grFilter1 <- "SWEET14p_BT07_Sense:340-448" # Functionally Validated 
grFilter2 <- "SWEET13p_BT7_Sense:360-490"


```

Let's first plot a region on the promoter of OsSWEET14 where all predictions were previously functionnaly validated 

```{r plotTaleTargetPred on pSWEET14 with preditale, fig.width = 15, fig.height=4, eval = TRUE}
sw14PrediPlot <- plotTaleTargetPred(predResults = preditalePreds, subjDnaSeqFile = subjDnaSeqFile, filterRange = grFilter1)
class(sw14PrediPlot)
sw14PrediPlot
```

Here is the same region and set of TALE RVD sequences but with `talvez()` predictions.

```{r plotTaleTargetPred on pSWEET14 with talvez, fig.width = 15, fig.height=4, eval = TRUE}
plotTaleTargetPred(predResults = talvezPreds, subjDnaSeqFile = subjDnaSeqFile, filterRange = grFilter1)
```

Now, a more speculative situation but that has the merit to showcase a region with a set of TALEs that are predicted to bind on either strand of the DNA sequence.

```{r plotTaleTargetPred on pSWEET14, fig.width = 15, fig.height=4, eval = TRUE}
plotTaleTargetPred(predResults = talvezPreds, subjDnaSeqFile = subjDnaSeqFile, filterRange = grFilter2)
```




