---
title: "plotting"
author: "Tram"
date: "4/18/2020"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tantale)
```

# get distal runned
```{r}
test_path <- system.file("test", package = "tantale", mustWork = T)

combine_input <- system.file("test", "telltale_output", "combined_correctedRepeatAASeqs.fas", package = "tantale", mustWork = T)

distal_outdir <- file.path(test_path, "distal_output")
distal_output <- runDistal(combine_input, outdir = distal_outdir, overwrite = F)

groups_output <- buildDisTalGroups(distal_outdir, 4.5, overwrite = F)
```

# plots repcode
```{r}
all_tal_sim <- distal_output$tal.similarity

subset.talsim <- function(talsim, tal.names) {
  talsim_grp <- talsim[talsim$TAL1 %in% tal.names, ]
  talsim_grp <- talsim_grp[talsim_grp$TAL2 %in% tal.names, ]
  talsim_grp <- as.matrix(reshape2::acast(talsim_grp, TAL1 ~ TAL2, value.var="Sim")) # melt then unmelt ...
  taldist_grp <- 100 -talsim_grp
  taldist_grp <- taldist_grp[order(rownames(taldist_grp)), ]
}


saved_files_path <- file.path(test_path, "distal_images")
for (i in names(groups_output$SeqOfRepsAlignments)) {
  clustID_matrix <- tantale::convertRepeat2ClusterIDAlign(repeatSim = distal_output$repeat.similarity, repeatAlign =  groups_output$SeqOfRepsAlignments[[i]])
  # imgFile <- file.path(saved_files_path, paste0("Repcode_", i, ".svg"))
  # svg(imgFile, width = 10, height = 6)
  # taldist <- subset.talsim(all_tal_sim, rownames(groups_output$SeqOfRepsAlignments[[i]]))
  tantale::heatmap_msa(talsim = all_tal_sim, forMatrix = clustID_matrix, forCellNote = groups_output$SeqOfRepsAlignments[[i]], forCol = distal_output$repeats.cluster$color, plot.type = "repeat.clusters", main = i)
}

```


# plot rvd

```{r}
for (i in names(groups_output$SeqOfRepsAlignments)) {
  # taldist <- subset.talsim(all_tal_sim, rownames(groups_output$SeqOfRepsAlignments[[i]]))
  # if (all(taldist == 0)) next()
  # imgFile <- file.path(saved_files_path, paste0("RVD_", i, ".svg"))
  # svg(imgFile, width = 10, height = 6)
  tantale::heatmap_msa(talsim = all_tal_sim, forMatrix = groups_output$SeqOfDistancesAlignments[[i]], forCellNote = groups_output$SeqOfRvdAlignment[[i]], plot.type = "repeat.similarity", main = i)
  # if(inherits(possibleError, "error")) {dev.off()}
  # dev.off()
}

```

No plots if I don't check taldist
`Error in (function (side, at = NULL, labels = TRUE, tick = TRUE, line = NA,  : no locations are finite`
->SC: What do you mean?



# test plotting with annotale
```{r}
annotale_fa <- system.file("test", "distal_with_annotale", "combined_CDS_Seqs.fas", package = "tantale", mustWork = T)
anno_distal_outdir <- system.file("test", "distal_with_annotale", package = "tantale", mustWork = T)
anno_distal_output <- runDistal(combine_input, outdir = anno_distal_outdir)

anno_groups_output <- buildDisTalGroups(anno_distal_outdir, 4.8, overwrite = T)
```
Error in builDistalGroups(): All seqs in ALIGNS files are empty

->SC:  WEIRD... Did you try to run the perl code directly to see what happens?


# plot from annotale output


```{r annotale_output}
distal_dir <- system.file("test/distal_output", package = "tantale", mustWork = T)
cds_to_distal <- file.path(distal_dir, "cds.fasta")
unlink(cds_to_distal)
file.create(cds_to_distal)

rvd_to_distal <- file.path(distal_dir, "rvd.fasta")
unlink(rvd_to_distal)
file.create(rvd_to_distal)

annotale_dir <- system.file("test/annotale_output", package = "tantale", mustWork = T)
annotale_cds_files <- list.files(annotale_dir, "TALE_DNA_sequences_(.+)\\.fasta", recursive = T, full.names = T)
file.append(cds_to_distal, annotale_cds_files)
cds <- readDNAStringSet(cds_to_distal)
names(cds) <- gsub(" \\[.+\\]", "", names(cds))
writeXStringSet(cds, cds_to_distal)

annotale_rvd_files <- list.files(annotale_dir, "TALE_RVDs\\.fasta", recursive = T, full.names = T)
file.append(rvd_to_distal, annotale_rvd_files)

rvdVectors <- readBStringSet(rvd_to_distal)
rvdVectors <- BStringSet(sapply(rvdVectors, function(r) paste("NTERM", r, "CTERM", sep = "-")), use.names = T)
names(rvdVectors) <- gsub(" \\[.+\\]", "", names(rvdVectors))
rvdVectors <- sapply(rvdVectors, function(r) strsplit(toString(r), split = "-"))
```

```{r distal}
distal_output <- runDistal(cds_to_distal, outdir = distal_dir, overwrite = T)

groups <- 9
taleGroups <- groupTales(distal_output$tal.similar, numGroups =  groups, plot = TRUE)

distal_group <- buildDisTalGroups(distal_dir, num.groups = groups, overwrite = T)

repeatVectors <- distal_output$coded.repeats.str
rep2rvd <- getRepeat2RvdMapping(talesRepeatVectors = repeatVectors, talesRvdVectors = rvdVectors)
```

```{r plots, fig.width = 15, fig.height=4}
plot_dir <- file.path(distal_dir, "plots")
dir.create(plot_dir)

# repcode
for (i in names(distal_group$SeqOfRepsAlignments)) {
  clustID_matrix <- tantale::convertRepeat2ClusterIDAlign(repeatSim = distal_output$repeat.similarity, repeatAlign =  distal_group$SeqOfRepsAlignments[[i]], h.cut = 85)
  if (nrow(clustID_matrix) < 2) next()
  imgFile <- file.path(plot_dir, paste0("Repcode_", i, ".svg"))
  tantale::heatmap_msa(talsim = distal_output$tal.similarity, forMatrix = clustID_matrix, forCellNote = distal_group$SeqOfRepsAlignments[[i]], plot.type = "repeat.clusters", main = i, save.path = imgFile)
}

# rvd
for (i in names(distal_group$SeqOfRepsAlignments)) {
  imgFile <- file.path(plot_dir, paste0("RVD_", i, ".svg"))
  rvd_msa <- convertRepeat2RvdAlign(repeatAlign = distal_group$SeqOfRepsAlignments[[i]], repeat2RvdMapping = rep2rvd)
  if (nrow(rvd_msa) < 2) next()
  tantale::heatmap_msa(talsim = distal_output$tal.similarity, forMatrix = distal_group$SeqOfDistancesAlignments[[i]], forCellNote = rvd_msa, plot.type = "repeat.similarity", main = i, save.path = imgFile)
}
```

```{r}
test <- file.path(distal_dir, "test")
distal_output <- runDistal("/home/baotram/tal/DisTAL1.2_MultipleAlignment/DisTAL_and_Additionalscripts/SampleTALSaa.fa", outdir = distal_dir, overwrite = T)

groups <- 9
taleGroups <- groupTales(distal_output$tal.similar, numGroups =  groups, plot = TRUE)

distal_group <- buildDisTalGroups(distal_dir, num.groups = groups, overwrite = T)
```

# plot from telltale output

```{r telltale2_output}
distal_dir <- system.file("test/distal_output", package = "tantale", mustWork = T)
unlink(distal_dir, recursive = T)
dir.create(distal_dir)
cds_to_distal <- file.path(distal_dir, "telltale2_cds.fasta")
unlink(cds_to_distal)
file.create(cds_to_distal)



telltale2_dir <- system.file("test/telltale2_output", package = "tantale", mustWork = T)
telltale2_cds_files <- list.files(telltale2_dir, "putativeTalCds.fasta", recursive = T, full.names = T)
telltale2_cds_files <- telltale2_cds_files[!grepl("temp_annotale", telltale2_cds_files, fixed = T)]
for (input in telltale2_cds_files) {
  id <- basename(dirname(input))
  seqs <- readBStringSet(input, seek.first.rec = T)
  names(seqs) <- paste0(id, "_", names(seqs))
  names(seqs) <- stringr::str_trim(names(seqs))
  seqs <- seqs[width(seqs) != 0]
  writeXStringSet(seqs, cds_to_distal, append = T)
}

readDNAStringSet(cds_to_distal)

rvd_to_distal <- file.path(distal_dir, "telltale2_rvd.fasta")
unlink(rvd_to_distal)
annotale_rvd_files <- list.files(telltale2_dir, "rvdSequences.fas", recursive = T, full.names = T)
for (input in annotale_rvd_files) {
  id <- basename(dirname(input))
  seqs <- readBStringSet(input, seek.first.rec = T)
  names(seqs) <- paste0(id, "_", names(seqs))
  names(seqs) <- stringr::str_trim(names(seqs))
  seqs <- seqs[width(seqs) != 0]
  writeXStringSet(seqs, rvd_to_distal, append = T)
}

rvdVectors <- readBStringSet(rvd_to_distal)
rvdVectors <- sapply(rvdVectors, function(r) strsplit(toString(r), split = "-"))
```

```{r distal}
distal_output <- runDistal(cds_to_distal, outdir = distal_dir, overwrite = T)

groups <- 9
taleGroups <- groupTales(distal_output$tal.similar, numGroups =  groups, plot = TRUE)

distal_group <- buildDisTalGroups(distal_dir, num.groups = groups, overwrite = T)

repeatVectors <- distal_output$coded.repeats.str
rep2rvd <- getRepeat2RvdMapping(talesRepeatVectors = repeatVectors, talesRvdVectors = rvdVectors)

```

