---
title: "TALE multiple aligments mith MAFFT?"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = TRUE)
library(tantale)
```

# Forewords


Fortitiuosly found out that MAFFT can generate mutliple alignment of character sequences : https://mafft.cbrc.jp/alignment/software/textcomparison.html

MAFFT `--text` mode actually accepts characters 0x01 – 0xFF excluding > (0x3E), = (0x3D), < (0x3C), - (0x2D), Space (0x20), Carriage Return (0x0d) and Line Feed (0x0a).  So the maximum size of alphabet is 248.

So this should in principle be possible to convert RVD sequences or repeat unit sequences in single characters strings and run it...


RVD seq -> hexadecimal code -> ASCII


# [Ascii code table in R](https://www.r-bloggers.com/ascii-code-table-in-r/)

accept an ascii character and return the decimal value

```{r}
asc <- function(x) { strtoi(charToRaw(x),16L) }
```

return the ascii character given it’s decimal code value

```{r}
chr <- function(n) { rawToChar(as.raw(n)) }
```


# Testing `buildRepeatMsa` 

```{r}
#taleRvdSeqsFile <- system.file("extdata", "TalF_RVDSeqs_AnnoTALE.fasta", package = "tantale")
taleRvdSeqsFile <- system.file("extdata", "TalA_RVDSeqs_AnnoTALE.fasta", package = "tantale")

#taleRepeatSeqsFile <- system.file("extdata", "Out_CodedRepeats.fa", package = "tantale")
taleRepeatSeqsFile <- system.file("extdata", "small_Out_CodedRepeats.fa", package = "tantale")

#distalRepeatDistMatFile <- system.file("extdata", "BigRepDist.mat", package = "tantale")
distalRepeatDistMatFile <- system.file("extdata", "Out_Repeatmatrix.mat", package = "tantale")

buildRepeatMsa(inputSeqs=taleRvdSeqsFile)
buildRepeatMsa(inputSeqs=taleRvdSeqsFile, mafftOpts = "--quiet")
buildRepeatMsa(inputSeqs=taleRvdSeqsFile, distalRepeatSims=distalRepeatDistMatFile)
buildRepeatMsa(inputSeqs=taleRepeatSeqsFile, distalRepeatSims=distalRepeatDistMatFile)
buildRepeatMsa(inputSeqs=taleRepeatSeqsFile)
```


# Testing `pickRefName` 

```{r}
rvdAlign <- buildRepeatMsa(inputFile=taleRvdSeqsFile)
repeatAlign <- buildRepeatMsa(inputFile=taleRepeatSeqsFile, distalRepeatDistMatFile=distalRepeatDistMatFile)

pickRefName(rvdAlign[ c(5, 4, 3, 2, 1), ])
pickRefName(rvdAlign[ c(5, 4, 3, 2, 1), ], refTag = "MAI")
pickRefName(rvdAlign[ c(5, 4, 3, 2, 1), ], refTag = "CFBP1947")

```

