---
title: "Predicting TALE targets"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

The previous articles found TALE arrays, grouped them, and aligned
related ones. This last article asks the question a TALE's RVD sequence
exists to answer: what DNA sequence does it bind? Each RVD contacts one
base of the target through a well-characterised, largely one-to-one code
(`NI` prefers A, `HD` prefers C, `NG` prefers T, `NN` prefers G or A, and
so on), so a whole RVD sequence predicts an *EBE* (Effector Binding
Element) as a short DNA motif with one position per RVD.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

tantale wraps two independent predictors that use this code differently
enough to be worth contrasting:
[TALVEZ](https://doi.org/10.1371/journal.pone.0068464), a
position-weight-matrix scan, and
[PrediTALE](https://doi.org/10.1371/journal.pcbi.1007206), which also
models where a real EBE tends to sit relative to a transcription start
site. Both are wrapped as `talvez()` and `preditale()`, which take the
same inputs and share their core columns (`taleId`, `subjSeqId`, `start`,
`end`, `strand`, `score`, `ebeSeq`), so their predictions can be compared
directly. Their scores are on different scales, though, and each adds
its own column: `rank` for TALVEZ, `pval` for PrediTALE.

## Getting RVD sequences to predict with

RVD sequences come straight from a `tales` object via
`tales_rvd_strings()`, with the termini dropped by default, since only
the repeats bind DNA. This reuses the same three genomes as
[the classification](tale_classification.html) and
[alignment](tale_msa.html) articles, from the classification article's
cached discovery result (see
[the getting-started article](tantale.html) for how these four
articles are linked).

```{r cached_pipeline, include = FALSE}
read_cache <- function(path) {
  if (!fs::file_exists(path)) {
    cli::cli_abort(c(
      "Cached pipeline result not found at {.path {path}}.",
      "i" = "Render {.file tale_classification.qmd} first (or run a full {.code pkgdown::build_site()}), which computes and caches it."
    ))
  }
  readRDS(path)
}
all_tales <- read_cache("_cache/discovery.rds")
```

```{r rvds}
rvds <- tales_rvd_strings(all_tales)
length(rvds)
```

The target sequences are promoter regions of three clade III *SWEET*
genes, a well-studied class of rice susceptibility genes that TALEs from
*Xanthomonas oryzae* are known to target, from several rice varieties:

```{r subj_file}
subj_file <- system.file("extdata", "cladeIII_sweet_promoters.fasta",
                         package = "tantale", mustWork = TRUE)
Biostrings::readBStringSet(subj_file) |> names()
```

## TALVEZ

```{r talvez, message = FALSE}
talvez_preds <- talvez(rvd_seqs = rvds, subj_file = subj_file, opt_param = "-t 0 -l 19")
nrow(talvez_preds)
```

```{r talvez_show}
talvez_preds |> arrange(desc(score)) |> head(5)
```

## PrediTALE

```{r preditale, message = FALSE, results = "hide"}
preditale_preds <- preditale(rvd_seqs = rvds, subj_file = subj_file)
```

```{r preditale_show}
nrow(preditale_preds)
preditale_preds |> arrange(desc(score)) |> head(5)
```

::: {.callout-tip}
## PrediTALE's `pval`
PrediTALE additionally reports a `pval` per prediction, since it models
the null distribution of scores directly. It is the column to use when
the question is "is this a credible site at all", beyond "which of these
is the best".
:::

## Visualising predictions against the target sequence

`plot_target_preds()` draws predicted RVD-to-base correspondences
directly against the target DNA, coloured by how well each RVD's known
base preference matches the base underneath it, so a prediction
supported along its whole length is visually distinguishable from one
held up by only one or two well-matching RVDs.

::: {.callout-important}
## Keep the plotted window narrow
Every RVD gets its own box, one per base of the target. A window wide
enough to span several unrelated predictions squeezes every box down to
a sliver too narrow for its own label. Pick a window around *one* region
of interest (here, a site where TALEs from different strains converge)
instead of the whole promoter at once.
:::

The same locus turns up more than once across strains, as expected,
since [the classification article](tale_classification.html) found this
comparison dominated by one-locus-per-strain groups. Three arrays, one
from each strain and all members of the same classification group,
predict the *same* site:

```{r best_predictions}
converging <- preditale_preds |>
  filter(taleId %in% c("BAI3_ROI_00008", "BAI3-1-1_ROI_00007", "MAI1_ROI_00008"),
        subjSeqId == "SWEET14p_Nipponbare_Sense")
converging |> select(subjSeqId, start, end, score, taleId)
```

```{r}
#| label: fig-target-preds
#| fig-cap: "Three orthologous TALEs, one from each strain, predicted to bind the exact same site in the SWEET14 promoter."
#| fig-width: 12
#| fig-height: 3.6
plot_target_preds(preds = converging, subj_file = subj_file,
                  filter_range = "SWEET14p_Nipponbare_Sense:460-505")
```

Each RVD is printed in a box over the base it is predicted to contact;
predictions on the sense strand are drawn above the sequence and those on
the antisense strand below it, since a TALE can bind either strand of
its target. The first box, `OO`, marks position 0: the base just before
the first repeat's target, where TALEs strongly prefer a T. All three
predictions in @fig-target-preds land on the same site with identical
scores, as expected of one locus conserved across
the three strains.

## Closing the loop

This set of articles followed one path through a TALE study:
`tell_tales()` to find arrays (correcting them, and checking for what
correction could not fix, along the way), `tales_compare_distal()` and
`tales_group_hclust()`/`tales_group_kmedoids()` to relate them,
`tales_align()` to see that relatedness directly, and
`talvez()`/`preditale()` to ask what they do. None of these
steps requires the ones before it in code (each function documented
here takes ordinary `tales` objects or plain sequences), but together
they cover most of what a study of TALE diversity in a set of genomes
actually needs.
