---
title: "Mining TALE sequences in genomes"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

This is the first of a set of articles walking through a typical study of
TALE diversity: finding TALE genes in a genome (here), classifying the
arrays found into groups of related sequences, aligning those groups, and
predicting the DNA targets of the TALEs they contain. Two of the articles
in the set are deep dives into a single class (`tales`, `tales_msa`)
rather than the next step in the walkthrough; they are linked from the
point where they become relevant, not numbered into the sequence.

This article covers TALE discovery with `tell_tales()`, and something
discovery cannot avoid: real assemblies are not always clean,
`tell_tales()` can sometimes be fooled, and tantale offers two ways to fix
that which are not interchangeable and neither of which is guaranteed to
work. `tales_anomalies()` is the tool that tells you which.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

::: {.callout-note}
## The genomes used throughout
Four *Xanthomonas oryzae* genomes appear across this whole set of
articles: **MAI1**, **BAI3** and **PXO86** are "clean" assemblies, and
**BAI3-1-1** is a deliberately error-prone one -- an assembly of the same
BAI3 background with *talC* deleted, carrying real sequencing/assembly
artefacts in its TALE loci. That last kind is not a corner case: it is
exactly the kind of input a mining pipeline has to cope with in practice,
and it is what the second half of this article is about.
:::

## Finding TALE loci in genomic DNA

`tell_tales()` searches a genome with `nhmmer`, using three profile HMMs
tuned to the N-terminus, the repeat unit, and the C-terminus of a TALE
CDS. Hits are merged, grouped into candidate arrays by proximity, and each
array's longest ORF is handed to AnnoTALE to split into parts and call its
RVD sequence. `NTERM`/`CTERM` markers are appended to the RVD string
wherever a terminus was identified this way, so a downstream alignment
knows where an array actually starts and ends.

A scratch directory holds everything this set of articles builds:

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_mining"))
```

Start with MAI1, a clean assembly, to see what discovery looks like when
nothing goes wrong:

```{r mai1_path}
mai1_fa <- system.file("extdata", "MAI1.fa", package = "tantale", mustWork = TRUE)
mai1_dir <- file.path(out, "MAI1")
```

```{r discover_mai1, message = FALSE}
invisible(tell_tales(subject_file = mai1_fa, output_dir = mai1_dir))
```

The result on disk is a directory of reports and fasta files. The result
in R is a `tales` object -- one row per part, not per TALE:

```{r tales_from_mai1}
mai1 <- suppressWarnings(tales_from_telltale(mai1_dir))
mai1
```

::: {.callout-tip}
## What a `tales` object actually is
Rows, columns, subsetting, and the `dom_code` column that the rest of
tantale is built on are covered in depth in
[the `tales` class article](tales_class.html) -- read it once this
walkthrough has given you an object to look at.
:::

`plot()` on a `tales` gives a compact overview of what was found: every
array's parts, positioned within the array, coloured by domain type and
filled by amino-acid length, with the RVD printed on each repeat.

```{r}
#| label: fig-mai1-composition
#| fig-cap: "Domain composition of the nine TALE arrays found in MAI1."
#| fig-width: 9
#| fig-height: 4
plot(mai1)
```

@fig-mai1-composition already shows something worth noticing: array
lengths vary, and every array here runs the full width of its row --
there are no gaps, because nothing has been aligned yet. `position_in_array`
just counts parts within each array independently.

## What `tell_tales()` writes to disk

Besides the RVD sequences and the TALE ORFs themselves, `tell_tales()`
writes three tab-separated reports, which `tales_from_telltale()` reads
back to build the object above. They are worth knowing about directly,
since they carry a few things the `tales` object does not.

```{r report_table, echo = FALSE}
knitr::kable(tibble::tribble(
  ~file, ~`one row per`, ~`worth knowing`,
  "`hitsReport.tsv`", "raw `nhmmer` hit", "`codon_count`, `frameshift_count` per hit -- before any merging or correction",
  "`domainsReport.tsv`", "domain, after AnnoTALE", "what AnnoTALE actually parsed out of the ORF",
  "`arrayReport.tsv`", "candidate array", "`has_all_domains`, `has_aberrant_repeat`, `orf_coverage`, and (with correction) `predicted_ins_count`/`predicted_dels_count`"
))
```

`arrayReport.tsv` is the one this article leans on most, so it is worth
reading directly rather than only through the `tales` object:

```{r array_report}
array_report <- readr::read_tsv(file.path(mai1_dir, "arrayReport.tsv"),
                                show_col_types = FALSE)
array_report %>%
  select(array_id, has_all_domains, longest_orf_length, orf_coverage)
```

`orf_coverage` is the longest ORF found, as a percentage of the whole
candidate array region -- a clean array covers nearly all of it. That
number is about to matter a great deal.

## When discovery goes wrong: detecting anomalies {#sec-anomalies}

MAI1 is a good assembly, and every array above came back complete. That is
not guaranteed. Running the same search on BAI3-1-1 -- deliberately
error-prone -- shows what an assembly artefact does to the pipeline.

```{r bai311_path}
bai311_fa <- system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE)
bai311_raw_dir <- file.path(out, "BAI3-1-1_raw")
```

```{r discover_bai311_raw, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_raw_dir,
                     cterm_min_score = 300))
```

```{r tales_from_bai311_raw}
bai311_raw <- suppressWarnings(tales_from_telltale(bai311_raw_dir))
```

`tales_anomalies()` reports the arrays that are odd rather than merely
absent -- missing sequence data, impossible terminus arrangements,
coordinate disagreements:

```{r anomalies_raw}
tales_anomalies(bai311_raw)
```

Two arrays have no `rvd` at all: AnnoTALE could not parse a repeat-array
structure out of their ORF. This is a frameshift, not a truncation --
compare their `orf_coverage` to a complete array:

```{r orf_coverage_raw}
readr::read_tsv(file.path(bai311_raw_dir, "arrayReport.tsv"),
                show_col_types = FALSE) %>%
  select(array_id, has_all_domains, longest_orf_length, orf_coverage)
```

`ROI_00003` and `ROI_00005` cover barely a fifth and a quarter of their
candidate region -- a single inserted or deleted base early in the array
shifts every codon downstream of it, and the ORF finder simply stops at
the first premature stop codon it meets. Nothing here is a bug: this is
exactly the situation frameshift correction exists for.

## Correcting frameshifts, two ways

tantale offers two distinct routes to a corrected sequence, and they are
not interchangeable: one corrects candidate *arrays* after they have
already been found, from inside `tell_tales()`; the other corrects the
*genome* before discovery even starts. Both can fail, in different ways,
which is exactly why @sec-anomalies's check is worth repeating after
either one -- and, run on the same genome, they will turn out to disagree
with each other about which array still needs help.

### Correcting inside `tell_tales()`

`correct_array = TRUE` runs `DECIPHER::CorrectFrameshifts()` on each
candidate array, aligning it against a shipped reference set of real TALE
proteins and inferring the indels needed to remove premature stops.

::: {.callout-important}
## `max_comparisons` trades reliability for speed
`max_comparisons` caps how many references each array may be aligned
against, and is the main control on how long correction takes. The full
docs on `tell_tales()` measure the trade-off directly: at the default
(all references) a set of four arrays took 252 seconds; capped to 20, 10
seconds -- but at a cap of 2, the aligner cannot reach a good reference
and invents indels wholesale. Capped to 20 is normally ample, and is used
below to keep this article fast to build -- but the section it fixes
also shows exactly what "ample, not certain" means in practice.
:::

```{r correct_inside_path}
bai311_corr_dir <- file.path(out, "BAI3-1-1_corrected")
```

```{r correct_inside, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_corr_dir,
                     cterm_min_score = 300,
                     correct_array = TRUE, max_comparisons = 20))
bai311_corr <- suppressWarnings(tales_from_telltale(bai311_corr_dir))
```

```{r anomalies_corr_inside}
tales_anomalies(bai311_corr)
```

Correction did fix the two frameshifted arrays -- `ROI_00003` and
`ROI_00005` are no longer flagged. But a *different* array, `ROI_00001`,
now is: it was clean in the uncorrected run above, and correcting against
only 20 references was, for this particular array, not enough to find a
good match. The correction did not fail by leaving the array alone; it
failed by aligning it against a poor reference, which is worse, because
the output still looks like a corrected ORF right up until AnnoTALE tries
to parse domains out of it and cannot.

Spending the full search on this same genome -- all ~500 shipped
references rather than 20, about eight minutes instead of ten seconds --
clears every anomaly, `ROI_00001` included, because the aligner can then
actually reach a reference close enough to it. There is no shortcut
around that trade-off: a small `max_comparisons` is a real risk, not a
rounding error, and checking `tales_anomalies()` after correcting is how
that risk gets caught rather than silently shipped downstream.

### Correcting the genome, before discovery

`correct_tales()` takes a different, faster route: rather than aligning
already-found candidate arrays against reference proteins, it wraps the
Java `TALEcorrection` tool to scan the *whole input genome* directly
against profile HMMs and repair frameshifts it finds, before
`tell_tales()` ever runs. That makes it a genome-wide pre-processing step
rather than a per-array one -- worth trying first on a large or
especially error-prone assembly (an ONT-sequenced genome, say), precisely
because it does not need candidate arrays to already exist.

```{r correct_outside_path}
bai311_java_fa <- file.path(out, "BAI3-1-1_java_corrected.fa")
```

```{r correct_outside, message = FALSE}
corrections <- correct_tales(uncorrected_path = bai311_fa,
                             corrected_path = bai311_java_fa,
                             return_corrections = TRUE)
```

`correct_tales()` reports every substitution it made, genome-wide, as a
table -- this genome needed `r nrow(corrections)`:

```{r corrections_head}
head(corrections, 4)
```

`tell_tales()` then runs on the corrected genome exactly as it would on
any other -- no `correct_array` needed, because the correction already
happened upstream:

```{r discover_after_java_correct_path}
bai311_java_dir <- file.path(out, "BAI3-1-1_java_tell_tales")
```

```{r discover_after_java_correct, message = FALSE}
invisible(tell_tales(subject_file = bai311_java_fa, output_dir = bai311_java_dir,
                     cterm_min_score = 300, correct_array = FALSE))
bai311_java <- suppressWarnings(tales_from_telltale(bai311_java_dir))
```

```{r anomalies_java}
tales_anomalies(bai311_java)
```

A third outcome again: `ROI_00003` is fixed, but `ROI_00005` still is not
-- the opposite pattern from the `max_comparisons = 20` run above, which
fixed both original frameshifts but broke a third array instead. Neither
correction path is a silver bullet, and which one a given array needs is
not predictable in advance.

### A clean correction without an eight-minute wait {#sec-best-correction}

Three results so far, none of them fully clean: `max_comparisons = 20`
trades a fix for a new break; `correct_tales()` alone leaves one array
short. Two ways forward, both measured rather than assumed:

```{r correct_moderate_path}
bai311_best_dir <- file.path(out, "BAI3-1-1_best")
```

```{r correct_moderate, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_best_dir,
                     cterm_min_score = 300,
                     correct_array = TRUE, max_comparisons = 50))
bai311_best <- suppressWarnings(tales_from_telltale(bai311_best_dir))
```

```{r anomalies_best}
tales_anomalies(bai311_best)
```

Zero anomalies -- raising the cap from 20 to 50 references was enough for
every array in this genome, in a little over a minute rather than eight.
`bai311_best` is the version the rest of this set of articles builds on.

::: {.callout-tip}
## A second way there: chain both correction paths
Running `correct_tales()` first and feeding *its* output into
`tell_tales(correct_array = TRUE)` also reaches zero anomalies, and lets
the per-array step use a smaller `max_comparisons` again, because the
genome-wide pass has already removed most of the damage `tell_tales()`
would otherwise have to correct for on its own. Measured on this same
genome (not re-run live here, to keep this article's build time down):

| route | time | anomalies |
|---|---|---|
| `max_comparisons = 50` alone | ~79 s | 0 |
| `correct_tales()` (~29 s) then `max_comparisons = 20` (~50 s) | ~79 s combined | still 1 (`ROI_00001`) |
| `correct_tales()` (~29 s) then `max_comparisons = 50` (~66 s) | ~95 s combined | 0 |

For this genome the two routes land at about the same cost. The chained
route is worth reaching for on a genome messy enough that neither
correction alone gets close: each pass only has to clean up what the
other left behind, rather than solving the whole problem itself.
:::

## How much did either correction actually help? {#sec-improvement}

Anomaly counts say whether AnnoTALE could parse an array at all, which is
a pass/fail signal. `orf_coverage` is more informative, since it is a
number and can be compared *before* an array even becomes an anomaly.
Read it back for the two originally-broken arrays, across all three runs:

```{r coverage_helper}
read_coverage <- function(dir, method) {
  readr::read_tsv(file.path(dir, "arrayReport.tsv"), show_col_types = FALSE) |>
    select(array_id, orf_coverage) |>
    filter(array_id %in% c("ROI_00003", "ROI_00005")) |>
    mutate(method = method)
}
```

```{r coverage_assemble}
coverage <- bind_rows(
  read_coverage(bai311_raw_dir,  "uncorrected"),
  read_coverage(bai311_corr_dir, "max_comparisons = 20"),
  read_coverage(bai311_java_dir, "correct_tales()"),
  read_coverage(bai311_best_dir, "max_comparisons = 50")
) |>
  mutate(method = factor(method, levels = c(
    "uncorrected", "max_comparisons = 20", "correct_tales()", "max_comparisons = 50"
  )))
```

```{r coverage_table}
coverage |>
  tidyr::pivot_wider(names_from = method, values_from = orf_coverage) |>
  knitr::kable()
```

```{r}
#| label: fig-coverage-improvement
#| fig-cap: "ORF coverage of the two frameshifted BAI3-1-1 arrays, before and after each correction attempt. max_comparisons = 50 is the only one of the three that recovers both arrays fully."
#| fig-width: 8
#| fig-height: 4
ggplot2::ggplot(coverage, ggplot2::aes(x = method, y = orf_coverage, fill = method)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ array_id) +
  ggplot2::scale_fill_viridis_d(guide = "none") +
  ggplot2::labs(x = NULL, y = "ORF coverage (%)") +
  ggplot2::theme_light() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
```

@fig-coverage-improvement is the honest version of "did correction help":
`max_comparisons = 20` and `correct_tales()` both recover most of
`ROI_00003`'s coding sequence outright, but only partly help
`ROI_00005` -- and only `max_comparisons = 50` reaches full coverage on
both, which is exactly why @sec-best-correction settled on it. What is
true of every attempted route is that `tales_anomalies()`, not a coverage
number by itself, is what actually says whether AnnoTALE could use the
result.

## Moving on with what you have

`bai311_best` needs none of this -- it is already clean -- but a route
that stops short of that, such as the `correct_tales()`-only run above,
still needs a decision about what to do with what remains flagged.
`tales(x, sanitize = TRUE)` drops it so an analysis can proceed on what is
clean, while keeping a clear record of what and why:

```{r sanitize}
bai311_clean <- tales(bai311_java, sanitize = TRUE)
n_distinct(bai311_java$array_id) - n_distinct(bai311_clean$array_id)
```

::: {.callout-note}
## What the rest of this set of articles uses
From here on, "BAI3-1-1" means `bai311_best` -- the
`max_comparisons = 50` correction from @sec-best-correction -- alongside
MAI1 and BAI3. PXO86, the fourth genome shipped with the package, is left
out of the classification and alignment articles that follow: it is a
distant Asian outgroup whose TALE repertoire barely overlaps the African
strains', and including it turns a clean set of one-ortholog-per-strain
groups into a mix of real cross-strain groups and PXO86-only paralog
clusters that do not illustrate the same thing.
:::

## Next

The [next article](tale_classification.html) picks up from a set of
`tales` objects like the ones built here -- across several genomes this
time -- and classifies their arrays into groups of related sequences with
`tales_compare()` and `tales_group()`.
