---
title: "Mining TALE sequences in genomes"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

This is the first of a set of articles walking through a typical study of
TALE diversity: finding TALE genes in a genome (here), classifying the
arrays found into groups of related sequences, aligning those groups, and
predicting the DNA targets of the TALEs they contain. The other articles
in the set are side branches: deep dives into a single class (`tales`,
`tales_msa`) and one case study on naturally truncated TALEs. Each is
linked from the point where it becomes relevant.

This article covers TALE discovery with `tell_tales()`. Because of their repetitive nature *tal* genes cannot be assembled from short read sequencing technologies data. This is why now most of the genomes of *tal* bearing bacteria are sequenced with long read technologies which may be error prone. Those assemblies are not always clean, and
`tell_tales()` can sometimes be fooled. tantale offers two ways to correct
frameshifts. They work differently, neither is guaranteed to succeed, and
`tales_anomalies()` tells you which arrays still need attention after
either one.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

::: {.callout-note}
## The genomes used throughout
Four *Xanthomonas oryzae* genomes appear across this set of articles:
**MAI1**, **BAI3** and **PXO86** are "clean" assemblies, and **BAI3-1-1**
is a deliberately error-prone one: an assembly of the same BAI3
background with *talC* deleted, carrying real sequencing/assembly
artefacts in its TALE loci. A mining pipeline meets this kind of input
routinely in practice, and it is what the second half of this article is
about.
:::

## Finding TALE loci in genomic DNA

`tell_tales()` searches a genome with `nhmmer`, using three profile HMMs
tuned to the N-terminus, the repeat unit, and the C-terminus of a TALE
CDS. Hits are merged, grouped into candidate arrays by proximity, and each
array's longest ORF is handed to AnnoTALE to split into parts and call its
RVD sequence. `NTERM`/`CTERM` markers are added at either end of the RVD
string wherever a terminus was identified this way, so a downstream
alignment knows where an array actually starts and ends.

A scratch directory holds everything this set of articles builds:

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_mining"))
```

Start with MAI1, a clean assembly, to see what discovery looks like when
nothing goes wrong:

```{r mai1_path}
mai1_fa <- system.file("extdata", "MAI1.fa", package = "tantale", mustWork = TRUE)
mai1_dir <- file.path(out, "MAI1")
```

```{r discover_mai1, message = FALSE}
invisible(tell_tales(subject_file = mai1_fa, output_dir = mai1_dir))
```

The result on disk is a directory of reports and fasta files. The result
in R is a `tales` object, with one row per part (repeat or terminus):

```{r tales_from_mai1}
mai1 <- suppressWarnings(tales_from_telltale(mai1_dir))
mai1
```

::: {.callout-tip}
## What a `tales` object actually is
Rows, columns, subsetting, and the `dom_code` column that the rest of
tantale is built on are covered in depth in
[the `tales` class article](tales_class.html) -- read it once this
walkthrough has given you an object to look at.
:::

`plot()` on a `tales` gives a compact overview of what was found: every
array's parts, positioned within the array, outlined by domain type and
filled by amino-acid length, with the RVD printed on each repeat.

```{r}
#| label: fig-mai1-composition
#| fig-cap: "Domain composition of the nine TALE arrays found in MAI1."
#| fig-width: 9
#| fig-height: 4
plot(mai1)
```

@fig-mai1-composition already shows something worth noticing: array
lengths vary, from 14 to 26 repeats (counting the final half-repeat), and
each array's parts sit side by side with no gaps, because nothing has
been aligned yet. `position_in_array` just counts parts within each array
independently.

## What `tell_tales()` writes to disk

Besides the RVD sequences and the TALE ORFs themselves, `tell_tales()`
writes three tab-separated reports, which `tales_from_telltale()` reads
back to build the object above. They are worth knowing about directly,
since they carry a few things the `tales` object does not.

```{r report_table, echo = FALSE}
knitr::kable(tibble::tribble(
  ~file, ~`one row per`, ~`worth knowing`,
  "`hits_report.tsv`", "raw `nhmmer` hit", "`codon_count`, `frameshift_count` per hit -- before any merging or correction",
  "`domains_report.tsv`", "domain, after AnnoTALE", "what AnnoTALE actually parsed out of the ORF",
  "`array_report.tsv`", "candidate array", "`has_all_domains`, `has_aberrant_repeat`, `orf_coverage`, and (with correction) `predicted_ins_count`/`predicted_dels_count`"
))
```

`array_report.tsv` is the one this article leans on most, so it is worth
reading directly:

```{r array_report}
array_report <- readr::read_tsv(file.path(mai1_dir, "array_report.tsv"),
                                show_col_types = FALSE)
array_report %>%
  select(array_id, has_all_domains, longest_orf_length, orf_coverage)
```

`orf_coverage` is the longest ORF found, as a percentage of the whole
candidate array region; a clean array's ORF covers 91-93% of it here.
That number is about to matter a great deal.

The report has one more row than the `tales` object has arrays.
`ROI_00005` is a candidate region with `has_all_domains` `FALSE` and no
ORF length; it yields no parts, so it is absent from
the object and from @fig-mai1-composition.

## When discovery goes wrong: detecting anomalies {#sec-anomalies}

MAI1 is a good assembly, and every array in the `tales` object above came
back complete. That is not guaranteed. Running the same search on
BAI3-1-1, the deliberately error-prone assembly, shows what an assembly
artefact does to the pipeline.

```{r bai311_path}
bai311_fa <- system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE)
bai311_raw_dir <- file.path(out, "BAI3-1-1_raw")
```

```{r discover_bai311_raw, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_raw_dir,
                     cterm_min_score = 300))
```

A first sign that there is something unexpected with these TALEs predictions are the warnings obtained when importing `tell_tales` results as a `tales` object:

```{r tales_from_bai311_raw}
bai311_raw <- tales_from_telltale(bai311_raw_dir)
```

`tales_anomalies()` reports arrays whose content is inconsistent: missing
sequence data, impossible terminus arrangements, coordinate
disagreements:

```{r anomalies_raw}
tales_anomalies(bai311_raw)
```

Two arrays have no `rvd` at all: AnnoTALE could not parse a repeat-array
structure out of the longest predicted ORF. Their `orf_coverage` shows why:

```{r orf_coverage_raw}
readr::read_tsv(file.path(bai311_raw_dir, "array_report.tsv"),
                show_col_types = FALSE) %>%
  select(array_id, has_all_domains, longest_orf_length, orf_coverage)
```

`ROI_00003` and `ROI_00005` predicted ORFs cover 19% and 23% of their candidate region.

No array in this assembly reaches MAI1's 91-93%: the
others fall between 38% and 70%. This is unusual and a sign also deserving further scrutiny even if `tales_anomalies` did not report any problem.
A close look at the 'array_report.tsv' table indicates that `ROI_00001` and `ROI_00007` have a predicted ORF encoding very few repeats (4-6) even if both termini profiles where found at the DNA level (the "has_all_domains" columns reports `TRUE`).

Because, in practice this barely happens in high quality genomes, in the absence of any further evidence, this could be explained by several in/dels along the CDS that completely break it.

```{r rvd_strings}

readr::read_tsv(file.path(bai311_raw_dir, "array_report.tsv"),
                show_col_types = FALSE) %>%
  select(array_id, has_all_domains, orf_coverage, rvd_string)
```





A premature stop could also be
genuine: some strains carry naturally truncated TALEs (see
[Genuine truncTALEs and frameshift correction](trunctale_correction.html)).
In BAI3-1-1 the corrections below settle it: once repaired, both arrays
recover the coverage of a clean MAI1 array, as expected of assembly
frameshifts.

## Correcting frameshifts, two ways

tantale offers two distinct routes to a corrected sequence: one corrects
candidate *arrays* after they have been found, from inside `tell_tales()`;
the other corrects the *genome* before discovery starts. Both can fail,
in different ways, so @sec-anomalies's check is worth repeating after
either one. Run on the same genome below, they disagree about which
array still needs help.

### Correcting inside `tell_tales()`

`correct_array = TRUE` runs `DECIPHER::CorrectFrameshifts()` on each
candidate array, aligning it against a shipped reference set of real TALE
proteins and inferring the indels needed to remove premature stops.

::: {.callout-important}
## `max_comparisons` trades reliability for speed
`max_comparisons` caps how many references each array may be aligned
against, and is the main control on how long correction takes. The full
docs on `tell_tales()` measure the trade-off directly: four arrays
aligned against a 1057-sequence reference set took 252 seconds uncapped
and 10 seconds capped to 20. At a cap of 2, the aligner cannot reach a
good reference and invents indels wholesale. A cap of 20 is usually
ample, and is used below to keep this article fast to build; the next
section shows a case where it is not enough.
:::

```{r correct_inside_path}
bai311_corr_dir <- file.path(out, "BAI3-1-1_corrected")
```

```{r correct_inside, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_corr_dir,
                     cterm_min_score = 300,
                     correct_array = TRUE, max_comparisons = 20))
```

Something is odd:

```{r correct_inside warnings}
bai311_corr <- tales_from_telltale(bai311_corr_dir)
```

Anomalies reported:

```{r anomalies_corr_inside}
tales_anomalies(bai311_corr)
```

Correction did fix the two frameshifted arrays -- `ROI_00003` and
`ROI_00005` are no longer flagged. But a *different* array, `ROI_00001`,
now is: it was clean in the uncorrected run above, and 20 references
were, for this particular array, not enough to find a good match. The
aligner corrected it against a poor reference instead. That is worse than
leaving it alone, because the output still looks like a corrected ORF
right up until AnnoTALE tries to parse domains out of it and cannot.

Running the full search on this same genome, against all ~500 shipped
references, takes about eight minutes instead of twenty seconds and
clears every anomaly, `ROI_00001` included: the aligner can then reach a
reference close enough to it. A small `max_comparisons` is a real risk,
and checking `tales_anomalies()` after correcting is how it gets caught
before it reaches downstream analyses.

### Correcting the genome, before discovery

`correct_tales()` takes a different, faster route. It wraps the Java
`TALEcorrection` tool, which scans the *whole input genome* against
profile HMMs and repairs the frameshifts it finds, before `tell_tales()`
ever runs. As a genome-wide pre-processing step that needs no candidate
arrays, it is worth trying first on a large or especially error-prone
assembly (an ONT-sequenced genome, say).

```{r correct_outside_path}
bai311_java_fa <- file.path(out, "BAI3-1-1_java_corrected.fa")
```

```{r correct_outside, message = FALSE}
corrections <- correct_tales(uncorrected_path = bai311_fa,
                             corrected_path = bai311_java_fa,
                             return_corrections = TRUE)
```

`correct_tales()` reports every edit it made, genome-wide, as a table of
positions and inserted or deleted bases. This genome needed
`r nrow(corrections)`:

```{r corrections_head}
head(corrections, 4)
```

`tell_tales()` then runs on the corrected genome exactly as it would on
any other, with no `correct_array`, because the correction already
happened upstream:

```{r discover_after_java_correct_path}
bai311_java_dir <- file.path(out, "BAI3-1-1_java_tell_tales")
```

```{r discover_after_java_correct, message = FALSE}
invisible(tell_tales(subject_file = bai311_java_fa, output_dir = bai311_java_dir,
                     cterm_min_score = 300, correct_array = FALSE))
bai311_java <- suppressWarnings(tales_from_telltale(bai311_java_dir))
```

```{r anomalies_java}
tales_anomalies(bai311_java)
```

Zero anomalies: `correct_tales()` fixes both `ROI_00003` and `ROI_00005`
on this genome, without touching any candidate array individually, and
does not break `ROI_00001`.

### A clean correction without an eight-minute wait {#sec-best-correction}

`correct_tales()` above already reached zero anomalies in under a minute.
`max_comparisons = 50` is a second, independent route to the same
outcome, useful when the Java correction tool is not available:

```{r correct_moderate_path}
bai311_best_dir <- file.path(out, "BAI3-1-1_best")
```

```{r correct_moderate, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_best_dir,
                     cterm_min_score = 300,
                     correct_array = TRUE, max_comparisons = 50))
bai311_best <- tales_from_telltale(bai311_best_dir)
```

```{r anomalies_best}
tales_anomalies(bai311_best)
```

Zero anomalies: raising the cap from 20 to 50 references was enough for
every array in this genome, in about a minute. Either this run or
`correct_tales()`'s own result above is a reasonable choice for the rest
of this set of articles; `bai311_best` (built here) is the one used from
here on.



## How much did either correction actually help? {#sec-improvement}

Anomaly counts say whether AnnoTALE could parse an array at all, which is
a pass/fail signal. `orf_coverage` is more informative, since it is a
number and can be compared *before* an array even becomes an anomaly.
Read it back for the two originally-broken arrays, across all three runs:

```{r coverage_helper}
read_coverage <- function(dir, method) {
  readr::read_tsv(file.path(dir, "array_report.tsv"), show_col_types = FALSE) |>
    select(array_id, orf_coverage) |>
    filter(array_id %in% c("ROI_00001", "ROI_00005")) |>
    mutate(method = method)
}
```

```{r coverage_assemble}
coverage <- bind_rows(
  read_coverage(bai311_raw_dir,  "uncorrected"),
  read_coverage(bai311_corr_dir, "max_comparisons = 20"),
  read_coverage(bai311_java_dir, "correct_tales()"),
  read_coverage(bai311_best_dir, "max_comparisons = 50")
) |>
  mutate(method = factor(method, levels = c(
    "uncorrected", "max_comparisons = 20", "correct_tales()", "max_comparisons = 50"
  )))
```

```{r coverage_table}
coverage |>
  tidyr::pivot_wider(names_from = method, values_from = orf_coverage) |>
  knitr::kable()
```

```{r}
#| label: fig-coverage-improvement
#| fig-cap: "ORF coverage of the two frameshifted BAI3-1-1 arrays, before and after each correction attempt. All three routes bring both arrays to the 91-93% of a clean MAI1 array."
#| fig-width: 8
#| fig-height: 4
ggplot2::ggplot(coverage, ggplot2::aes(x = method, y = orf_coverage, fill = method)) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ array_id) +
  ggplot2::scale_fill_viridis_d(guide = "none") +
  ggplot2::labs(x = NULL, y = "ORF coverage (%)") +
  ggplot2::theme_light() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
```

In @fig-coverage-improvement, all three routes bring `ROI_00001` and
`ROI_00005` to identical coverage, 93% and 91%. On these two numbers
alone the three routes look interchangeable. What separates them is the
new anomaly `max_comparisons = 20` introduced in a different array,
`ROI_00001`, and coverage of these two arrays cannot show it; only
`tales_anomalies()` catches it.

## Moving on with what you have

`bai311_best` needs none of this: it is already clean, and so is
`correct_tales()`'s own result above. A route that stops short of that,
such as `bai311_corr` (the `max_comparisons = 20` run, which fixed the
two original frameshifts but broke `ROI_00001`), still needs a decision
about what to do with what remains flagged. `tales(x, sanitize = TRUE)`
drops it so an analysis can proceed on what is clean, with a warning
naming each dropped array and the reason:

```{r sanitize}
bai311_clean <- tales(bai311_corr, sanitize = TRUE)
n_distinct(bai311_corr$array_id) - n_distinct(bai311_clean$array_id)
```

::: {.callout-note}
## What the rest of this set of articles uses
From here on, "BAI3-1-1" means `bai311_best`, the
`max_comparisons = 50` correction from @sec-best-correction, alongside
MAI1 and BAI3. PXO86, the fourth genome shipped with the package, is left
out of the classification and alignment articles that follow: it is a
distant Asian outgroup whose TALE repertoire barely overlaps the African
strains', and including it turns a clean set of one-ortholog-per-strain
groups into a mix of real cross-strain groups and PXO86-only paralog
clusters that do not illustrate the same thing.
:::

## Next

The [next article](tale_classification.html) picks up from a set of
`tales` objects like the ones built here, across several genomes this
time, and classifies their arrays into groups of related sequences with
`tales_compare_distal()` and `tales_group_hclust()`/`tales_group_kmedoids()`.
