---
title: "The tales_msa class"
format:
  html:
    toc: true
    number-sections: false
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The [previous article](tales-class.html) covered the `tales` class: one row
per part, with `dom_code` as the identity layer that makes an array
alignable. This one is about what happens once several arrays *are*
aligned -- the `tales_msa` class -- and about the traffic between the two
classes: how a `tales` is promoted into one, how it demotes back, and what
each class draws when you plot it.

```{r setup, message = FALSE}
library(tantale)
library(dplyr)
```

# Building one

`tales_align()` takes a `tales` and returns a `tales_msa`: the same rows,
plus one new column, `alignment_position`. Nothing else changes -- every
column the input carried, `rvd` and `dom_code` both included, survives the
trip, because the alignment is computed on *one* residue layer
(`residue_col`) but the object it returns is still the whole table.

The example below uses the same small fixture as the previous article --
four arrays, small enough to align in a couple of seconds -- but runs
`tales_compare()` first, since the alignment is more informative once
`dom_code` and the two distance tables exist alongside it:

```{r build_msa, message = FALSE, warning = FALSE}
out <- file.path(tempdir(), "tales_msa_article")
invisible(tell_tales(
  subject_file = system.file("extdata", "bai3_sample_tal_genomic_regions.fasta",
                             package = "tantale", mustWork = TRUE),
  output_dir = out))

x <- suppressWarnings(tales_from_telltale(out))
cmp <- tales_compare(x)
msa <- tales_align(cmp$tales, residue_col = "dom_code")
msa
```

`residue_col = "dom_code"` means MAFFT is shown each array as a sequence of
domain codes -- exactly the alignable-unit argument from the previous
article. Aligning on `"rvd"` instead is possible too, but coarser: distinct
repeats sharing an RVD would be forced to look identical to the aligner,
which is a real loss of resolution whenever it matters (again, see the
previous article's `HD` example).

# What is new: `alignment_position`

A `tales_msa` is a `tales` with one extra column and one extra stored
number:

- **`alignment_position`** -- where this part sits in the alignment, which
  is not the same coordinate as `position_in_array`. The two agree up to
  the first gap and diverge after it.
- **`tales_width()`** -- the alignment's total column count, stored rather
  than derived, because `max(alignment_position)` over a *subset* of arrays
  can under-report a width whose last columns happen to be all gaps in that
  subset.

```{r width_demo}
tales_width(msa)
msa %>% filter(array_id == msa$array_id[1]) %>%
  select(array_id, position_in_array, alignment_position, domain_type)
```

**Gaps are implicit, not stored.** There is no row, anywhere, whose
`alignment_position` is a gap -- a gap is simply a column with no row for
that array. This is why the class validates cheaply and why every `tales`
invariant (the key, the residue columns, the `dom_code` bijection) keeps
holding unchanged on a `tales_msa`: nothing new was added to *rows*, only a
new column that says where each existing row lands. What the alignment
adds on top is its own, narrower, set of rules -- checked by
`validate_tales_msa()` in addition to everything `validate_tales()` already
checks:

- `alignment_position` is a positive integer, present on every row;
- unique within an array (two parts cannot land in the same column);
- **ordered like `position_in_array`** -- an alignment may insert gaps, it
  may never reorder parts;
- never beyond the declared width.

# Views: matrices are computed, not stored

The long table is the only thing actually stored. A rectangular alignment
-- the picture most people have in mind, one row per array, one column per
position -- is a *view*, built on request by `as.matrix()`, and it has to
be asked which layer to render:

```{r as_matrix_demo}
as.matrix(msa, value = "dom_code")[, 1:8]
as.matrix(msa, value = "rvd")[, 1:8]
```

Both matrices have the same shape and the same gaps -- they are two
readings of one geometry, not two independent alignments. That equivalence
is exactly what lets `plot.tales_msa()` (below) show a repeat's identity as
a colour and its RVD as a label on the same cell.

# Coercion between `tales` and `tales_msa`

**Promotion** is `tales_align()`: give it a complete `tales`, get back a
`tales_msa`. There is no other route, because the alignment column has to
come from an actual alignment run -- nothing else could populate it
honestly.

**Demotion** is `as_tales()`, and it is a real operation, not a formality:

```{r demote}
demoted <- as_tales(msa)
class(demoted)
is_tales_msa(demoted)
"alignment_position" %in% names(demoted)
```

`alignment_position` survives as an ordinary column -- demoting does not
erase where things were aligned, it only stops the object *claiming* to be
a coherent single alignment. That claim is exactly what `tales_width()`
encodes, and it is gone: `tales_width(demoted)` is `NULL`. This matters
for more than bookkeeping. A talome-wide summary spanning several
*independently* aligned groups cannot honestly be one `tales_msa` --
each group has its own width and its own coordinate system, so
`alignment_position = 5` would mean unrelated things in two different
groups. Demoting to a plain `tales` and carrying `alignment_position` as
an ordinary column (per group) is the honest shape for that case; the
class is not stretched to cover something it cannot actually guarantee.

**Subsetting degrades the same way, automatically, one step at a time.**
Row subsetting never touches the alignment contract, since every check
above is closed under keeping a subset of rows. Column subsetting can:
dropping `alignment_position` itself leaves something that is still a
valid `tales` (everything else about a part is unaffected) but can no
longer be a `tales_msa`, and the class notices without being told:

```{r subset_degrade}
no_alignment <- msa %>% select(-alignment_position)
class(no_alignment)
```

This is the same graded-degradation mechanism from the previous
article's `[.tales` method, just visible one level up: a `tales_msa` that
loses what makes it an alignment steps down to a `tales` rather than lying
about still being one, and a `tales` that loses its key steps down to a
bare tibble for the same reason. Losing an *optional* column --
`seqnames`, `aa_seq` -- costs nothing at either level.

# Plotting: two methods, one generic

`plot()` dispatches on class, so the two objects draw different things
from the same call.

## `plot.tales()` before and after alignment

A `tales` without an alignment can only be laid out on
`position_in_array`: every array starts its own count at 1, so an
aberrant repeat scatters to wherever it happens to sit in each array.
`position = "alignment"` needs the column that only an aligned (or
demoted-from-aligned) object carries, and lines features up instead:

```{r plot_before_after, fig.width=9, fig.height=4}
plot(x, position = "array")
plot(demoted, position = "alignment")
```

## `plot.tales_msa()`: three things decided independently

The alignment plot draws a heatmap -- one row per array, one column per
position -- and three questions about it are answered separately:

- **what each cell *says*** -- `label`, defaulting to `rvd` whenever that
  is not also the fill layer;
- **what colour that text is** -- always whether the cell matches the
  consensus of its column (`tales_consensus()`), cyan for yes, pink for
  no, regardless of what `fill_type` is showing;
- **what colour the block behind it is** -- `fill_type`, the only one of
  the three that can be unavailable.

| `fill_type` | shows | needs |
|---|---|---|
| `"repeat_clust"` | which cluster the repeat falls in, cut at `h_cut` | `domain_sim` |
| `"repeat_sim"` | protein-sequence similarity to the reference, 0-100 | `domain_sim` |
| `"rvd_sim"` | how alike the RVD's *DNA-binding preference* is to the reference's, -1 to 1 | a `label` layer |

`domain_sim` is the `domain_distances` element of `tales_compare()`'s
result -- the same table `tales_domain_distances()` produces on its own.
Passing `tal_sim` (the `tale_distances` element) in addition adds a
dendrogram panel ordering the rows by relatedness:

```{r plot_msa_default, fig.width=10, fig.height=3}
plot(msa)
```

```{r plot_msa_sim, fig.width=10, fig.height=3}
plot(msa, fill_type = "repeat_sim",
    domain_sim = cmp$domain_distances, tal_sim = cmp$tale_distances)
```

`"repeat_sim"` and `"repeat_clust"` colour by protein-sequence similarity
to a reference row, chosen automatically as the array with the fewest
gaps (ties broken alphabetically) unless `ref_pattern` names one. `rvd_sim`
asks the same "how similar to the reference" question one layer up, at the
level of DNA-binding specificity rather than protein sequence -- and the
two genuinely disagree, because `HD` and `ND` are different repeats with
identical specificity, while repeats differing only at positions 12-13 are
near-identical proteins with opposite target bases. It needs no
`domain_sim` at all, only the RVDs themselves:

```{r plot_msa_rvdsim, fig.width=10, fig.height=3}
plot(msa, fill_type = "rvd_sim")
```

A consensus panel can sit above either view. It reads the same layer the
cells are labelled with, so it is never talking about something the cells
below it are not:

```{r plot_msa_consensus, fig.width=10, fig.height=3.3}
plot(msa, consensus = TRUE,
    domain_sim = cmp$domain_distances, tal_sim = cmp$tale_distances)
```

# Next

Aligning is rarely the last step in a real analysis -- the walkthrough
vignettes take this further, running the same mechanics over whole
genomes and TALE groups rather than one small fixture.
