---
title: "Multiple alignment of TALE arrays"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

The [previous article](tale_classification.html) grouped TALE arrays by
overall similarity. Grouping says arrays are related; it does not say
*how* -- which repeats correspond, where one array has an insertion or a
deletion relative to another. That needs an alignment.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

::: {.callout-note}
## Where this fits
This article is a walkthrough: pick one real classification group and
align it. The mechanics -- what a `tales_msa` actually is, coercion
between `tales` and `tales_msa`, and the plotting options in full --
are a deep dive in [the `tales_msa` class article](tales_msa_class.html),
which continues directly from the alignment built here rather than
starting over.
:::

The AnnoTALE tool this package builds on can assign TALEs to classes, but
cannot insert gaps while doing so -- and TALE arrays evolve substantially
by whole-repeat duplication and deletion, so a method that cannot
represent a gap is blind to exactly the events that matter most (see the
[QueTAL paper](https://doi.org/10.3389/fpls.2015.00545)). `tales_align()`
drives MAFFT in text mode instead, treating each distinct repeat as one
alignable symbol, which lets it open and close gaps freely.

## Re-deriving one group from the previous article

Reproduced here rather than carried over, since each article in this
series stands alone. The three genomes and the correction settings are
exactly [the previous article's](tale_classification.html).

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_msa"))
genome_files <- c(
  MAI1       = system.file("extdata", "MAI1.fa",     package = "tantale", mustWork = TRUE),
  BAI3       = system.file("extdata", "BAI3.fa",     package = "tantale", mustWork = TRUE),
  `BAI3-1-1` = system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE)
)
```

```{r discover_all, message = FALSE}
all_tales <- lapply(names(genome_files), function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir,
                       cterm_min_score = 300,
                       correct_array = TRUE, max_comparisons = 50))
  tales_from_telltale(strain_dir) |>
    mutate(array_id = paste0(strain, "_", array_id))
}) |>
  suppressWarnings() |>
  bind_rows() |>
  tales(sanitize = TRUE)
```

```{r compare_and_group, message = FALSE}
cmp <- tales_compare(all_tales, aln_method = "DECIPHER", ncores = 4)
grouped <- tales_group(cmp$tales, cmp$tale_distances,
                       method = "k-medoids", k = "auto", k_range = 2:20) |>
  suppressMessages()
```

One group holds a member from each of MAI1, BAI3 and BAI3-1-1 -- the same
locus, present in all three related African strains -- and, usefully for
this article, its three copies are not identical:

```{r pick_group}
picked_group <- grouped |> filter(group == 6)
picked_group |> distinct(array_id)
```

## Aligning the group

```{r align_group, message = FALSE}
msa <- tales_align(picked_group, residue_col = "dom_code")
msa
```

```{r}
#| label: fig-msa-default
#| fig-cap: "Alignment of one TALE locus across three related strains, coloured by repeat cluster."
#| fig-width: 9
#| fig-height: 2.3
plot(msa, tal_sim = cmp$tale_distances, domain_sim = cmp$domain_distances)
```

@fig-msa-default already shows something `tales_group()`'s single
distance number could not: **BAI3 and BAI3-1-1 are both missing four
repeats that MAI1 has**, at alignment positions 13-16 -- a real internal
gap, not one at either end of the array. Because BAI3-1-1 is the same
genomic background as BAI3 with *talC* deleted, the two sharing this
exact deletion is consistent with it being inherited from their common
background rather than independent events in each.

## Does a scoring matrix change the alignment?

`tales_align()` accepts a `repeat_sims` argument: a substitution cost
matrix MAFFT uses when scoring which repeats to match against each other,
rather than treating every mismatch as equally bad. Passing one is
optional, and it is worth checking what it actually changes rather than
assuming.

### Aligning on `dom_code`

Without one, every distinct repeat is as different from every other as
any other pair -- MAFFT has no notion that two repeats might be more or
less alike:

```{r align_nomatrix}
msa_plain <- tales_align(picked_group, residue_col = "dom_code")
```

With one, `domain_distances` -- the repeat-level protein similarity
already computed by `tales_compare()` -- tells MAFFT how alike two
repeats actually are:

```{r align_matrix, message = FALSE}
msa_scored <- tales_align(picked_group, residue_col = "dom_code",
                          repeat_sims = cmp$domain_distances)
```

```{r compare_matrices}
identical(as.matrix(msa_plain), as.matrix(msa_scored))
```

They differ. Looking at just the affected row shows how:

```{r compare_matrices_detail}
rbind(
  no_matrix = as.matrix(msa_plain)["BAI3-1-1_ROI_00006", ],
  scored    = as.matrix(msa_scored)["BAI3-1-1_ROI_00006", ]
)
```

Scored, the gap this array carries is one column narrower -- MAFFT could
place its real repeats slightly more compactly once it knew which
substitutions were cheap. This matches what the package documentation
already claims about `repeat_sims`: more compact alignments, fewer
gap columns, on data where it has something to work with.

### Aligning on `rvd`

The built-in RVD similarity matrix works the same way, opted into with
`repeat_sims = "rvd"` rather than a `domain_distances` table -- but scores
DNA-binding specificity instead of protein sequence, so it is worth
checking separately rather than assuming it behaves like the `dom_code`
case:

```{r align_rvd, message = FALSE}
msa_rvd_plain  <- tales_align(picked_group, residue_col = "rvd")
msa_rvd_scored <- tales_align(picked_group, residue_col = "rvd", repeat_sims = "rvd")
identical(as.matrix(msa_rvd_plain), as.matrix(msa_rvd_scored))
```

On this group, aligning on RVDs gives the *same* alignment whether or not
the matrix is supplied. That is a real result, not a shortcut taken here:
the RVD alphabet is much smaller than the repeat-code one (a handful of
distinct RVDs against a handful of distinct repeats, for three arrays
this closely related), so there was little room for a scoring matrix to
change anything. The lesson is not "the RVD matrix never matters" -- it
is that whether either matrix matters depends on the data, which is
exactly why this section checked rather than asserted it.

## Other views of the same alignment

Colouring by similarity to a reference instead of by cluster membership
makes graded divergence visible where cluster identity would only say
"different":

```{r}
#| label: fig-msa-sim
#| fig-cap: "The same alignment, coloured by protein-sequence similarity to the reference array."
#| fig-width: 9
#| fig-height: 2.3
plot(msa, fill_type = "repeat_sim",
    tal_sim = cmp$tale_distances, domain_sim = cmp$domain_distances)
```

And with a consensus panel attached, the column-by-column question "does
this array agree with the group" is answered directly above the
alignment itself:

```{r}
#| label: fig-msa-consensus
#| fig-cap: "The same alignment again, with a consensus row attached above it."
#| fig-width: 9
#| fig-height: 2.6
plot(msa, consensus = TRUE,
    tal_sim = cmp$tale_distances, domain_sim = cmp$domain_distances)
```

## Next

Once related arrays are aligned, the biologically motivated next question
is what DNA sequence their RVDs are predicted to bind -- covered in
[the final article](tale_target_prediction.html) of this series. To go
deeper on the alignment mechanics used here first, see
[the `tales_msa` class article](tales_msa_class.html).
