---
title: "Multiple alignment of TALE arrays"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

The [previous article](tale_classification.html) grouped TALE arrays by
overall similarity. Grouping says arrays are related. Saying *how*
(which repeats correspond, where one array has an insertion or a deletion
relative to another) needs an alignment.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

::: {.callout-note}
## Where this fits
This article is a walkthrough: pick one real classification group and
align it. The mechanics (what a `tales_msa` actually is, coercion
between `tales` and `tales_msa`, and the plotting options in full) are a
deep dive in [the `tales_msa` class article](tales_msa_class.html),
which continues directly from the alignment built here.
:::

The AnnoTALE tool this package builds on can assign TALEs to classes, but
cannot insert gaps while doing so. TALE arrays evolve substantially by
whole-repeat duplication and deletion, so a method that cannot represent
a gap is blind to exactly the events that matter most (see the
[QueTAL paper](https://doi.org/10.3389/fpls.2015.00545)). `tales_align()`
drives MAFFT in text mode instead, treating each distinct domain as one
alignable symbol, which lets it open and close gaps freely.

## The same group, from the previous article

Same three genomes (MAI1, BAI3, BAI3-1-1), same correction settings, and
the same comparison and grouping as
[the classification article](tale_classification.html), reused here from
its cached result; see [the getting-started article](tantale.html)
for how these four articles are linked.

```{r cached_pipeline, include = FALSE}
read_cache <- function(path) {
  if (!fs::file_exists(path)) {
    cli::cli_abort(c(
      "Cached pipeline result not found at {.path {path}}.",
      "i" = "Render {.file tale_classification.qmd} first (or run a full {.code pkgdown::build_site()}), which computes and caches it."
    ))
  }
  readRDS(path)
}
all_tales <- read_cache("_cache/discovery.rds")
cmp       <- read_cache("_cache/compare.rds")
grouped   <- read_cache("_cache/group.rds")
```

This group holds one member from each of MAI1, BAI3 and BAI3-1-1 (the
same locus, present in all three related African strains) and, usefully
for this article, its three copies are not identical:

```{r pick_group}
picked_group <- grouped |> filter(group == 6)
picked_group |> distinct(array_id)
```

## Aligning the group

```{r align_group, message = FALSE}
msa <- tales_align(picked_group, residue_col = "dom_code")
msa
```

```{r}
#| label: fig-msa-default
#| fig-cap: "Alignment of one TALE locus across three related strains, coloured by domain cluster. MAI1's array (marked #) is the reference."
#| fig-width: 9
#| fig-height: 2.3
plot(msa, tale_distances = cmp$tale_distances, domain_distances = cmp$domain_distances)
```

@fig-msa-default already shows something `tales_group_kmedoids()`'s single
distance number could not: **BAI3 and BAI3-1-1 both lack four repeats
that MAI1 has**, at alignment positions 13-16. The gap is internal: all
three arrays end with the same 20-residue half-repeat (column 17), which
lines up across the gap. BAI3-1-1 derives from BAI3, so the two share
this difference.

## Does a scoring matrix change the alignment?

`tales_align()` accepts a `domain_distances` argument: a substitution cost
matrix MAFFT uses when scoring which repeats to match against each other.
Without it, every mismatch counts as equally bad. Passing one is
optional, and it is worth checking what it actually changes.

### Aligning on `dom_code`

Without one, every distinct domain is as different from every other as
any other pair, and MAFFT has no notion that two repeats might be more or
less alike:

```{r align_nomatrix}
msa_plain <- tales_align(picked_group, residue_col = "dom_code")
```

With one, `domain_distances`, the domain-level protein similarity
already computed by `tales_compare_distal()`, tells MAFFT how alike two
repeats actually are:

```{r align_matrix, message = FALSE}
msa_scored <- tales_align(picked_group, residue_col = "dom_code",
                          domain_distances = cmp$domain_distances)
```

```{r compare_matrices}
identical(as.matrix(msa_plain), as.matrix(msa_scored))
```

On this group the two alignments are identical. The domain distances
count the residues a half-repeat lacks against a full repeat, so the
final half-repeat that BAI3 and BAI3-1-1 share with MAI1 stays opposite
its identical copy (column 17), and the four-repeat gap stays internal.
On more divergent arrays, the matrix lets MAFFT prefer matching similar
repeats over dissimilar ones; whether that changes the alignment depends
on the data, so it is worth checking as done here.

### Aligning on `rvd`

The built-in RVD similarity matrix works the same way, opted into with
`domain_distances = "rvd"` instead of a distance table. It scores
DNA-binding specificity instead of protein sequence, so it is worth
checking separately:

```{r align_rvd, message = FALSE}
msa_rvd_plain  <- tales_align(picked_group, residue_col = "rvd")
msa_rvd_scored <- tales_align(picked_group, residue_col = "rvd", domain_distances = "rvd")
identical(as.matrix(msa_rvd_plain), as.matrix(msa_rvd_scored))
```

On this group, aligning on RVDs also gives the *same* alignment whether or
not the matrix is supplied. The RVD alphabet is smaller than the domain-code
one (6 distinct RVDs against 12 distinct repeat codes here), and these
three arrays are closely related, so there was little room for a scoring
matrix to change anything. Whether either matrix matters depends on the
data.

## Other views of the same alignment

Colouring by similarity to the reference array instead of by cluster
membership makes graded divergence visible where cluster identity would
only say "different":

```{r}
#| label: fig-msa-sim
#| fig-cap: "The same alignment, coloured by protein-sequence similarity to the reference array."
#| fig-width: 9
#| fig-height: 2.3
plot(msa, fill_type = "domain_sim",
    tale_distances = cmp$tale_distances, domain_distances = cmp$domain_distances)
```

In @fig-msa-sim, every domain is identical to MAI1's except BAI3-1-1's
N-terminus, which has its own `dom_code` at about 99.7% similarity. The
cluster colouring of @fig-msa-default groups it with the other two
N-termini, so the difference only shows here.

With a consensus panel attached, the column-by-column question "does
this array agree with the group" is answered directly above the
alignment itself:

```{r}
#| label: fig-msa-consensus
#| fig-cap: "The same alignment again, with a consensus row attached above it."
#| fig-width: 9
#| fig-height: 2.6
plot(msa, consensus = TRUE,
    tale_distances = cmp$tale_distances, domain_distances = cmp$domain_distances)
```

## Next

Once related arrays are aligned, the biologically motivated next question
is what DNA sequence their RVDs are predicted to bind, covered in
[the final article](tale_target_prediction.html) of this series. To go
deeper on the alignment mechanics used here first, see
[the `tales_msa` class article](tales_msa_class.html).
