---
title: "3. Multiple alignment of TALE sequences with tantale"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{3. Multiple alignment of TALE sequences with tantale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r package, warning=FALSE, message=FALSE}
library(tantale)
library(dplyr)
```

The [previous article](2_tale_classification.html) grouped TALE arrays by
overall similarity. Grouping says arrays are related; it does not say *how*
-- which repeats correspond, where one array has an insertion or a deletion
relative to another. That needs an alignment. The mechanics of
`tales_align()` and of plotting a `tales_msa` are covered in full, with a
small fixture, in the [`tales_msa` class article](../articles/tales-msa-class.html);
this article instead picks up a real group from the classification just
performed and looks at what its alignment actually shows.

The AnnoTALE tool this package builds on can assign TALEs to classes, but
cannot insert gaps while doing so -- and TALE arrays evolve substantially by
whole-repeat duplication and deletion, so a method that cannot represent a
gap is blind to exactly the events that matter most (see the
[QueTAL paper](https://doi.org/10.3389/fpls.2015.00545)). `tales_align()`
drives MAFFT in text mode instead, treating each distinct repeat as one
alignable symbol, which lets it open and close gaps freely.

# Re-deriving one group from the previous article

Reproduced here rather than carried over, since each article in this series
stands alone:

```{r discover_and_group, message = FALSE, warning = FALSE}
out <- fs::dir_create(file.path(tempdir(), "tale_msa"))
genome_files <- c(
  MAI1     = system.file("extdata", "MAI1.fa",     package = "tantale", mustWork = TRUE),
  BAI3     = system.file("extdata", "BAI3.fa",     package = "tantale", mustWork = TRUE),
  `BAI3-1-1` = system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE),
  PXO86    = system.file("extdata", "PXO86.fa",    package = "tantale", mustWork = TRUE)
)

all_tales <- lapply(names(genome_files), function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir,
                       cterm_min_score = 300))
  tales_from_telltale(strain_dir) %>%
    mutate(array_id = paste0(strain, "_", array_id), strain = strain)
}) %>%
  suppressWarnings() %>%
  bind_rows() %>%
  tales(sanitize = TRUE)

cmp <- tales_compare(all_tales, aln_method = "DECIPHER", ncores = 4) %>%
  suppressMessages()
grouped <- tales_group(cmp$tales, cmp$tale_distances,
                       method = "k-medoids", k = "auto", k_range = 2:20) %>%
  suppressMessages()
```

One group holds a member from each of MAI1, BAI3 and BAI3-1-1 -- the same
locus, present in three related African strains, one of them the
error-prone assembly from the first article:

```{r pick_group}
group9 <- grouped %>% filter(group == 9)
group9 %>% distinct(array_id, strain)
```

# Aligning the group

```{r align_group, message = FALSE, warning = FALSE}
msa <- tales_align(group9, residue_col = "dom_code")
msa
```

```{r plot_default, fig.width=9, fig.height=3}
plot(msa, tal_sim = cmp$tale_distances, domain_sim = cmp$domain_distances)
```

Cells are filled by which repeat cluster each domain falls in and labelled
with its RVD; text colour marks whether that element matches the column's
consensus. Even at three arrays, the alignment already says something
`tales_group()`'s single distance number could not: whether these three
copies of the locus differ by simple substitutions at matched positions, or
by whole repeats present in some arrays and absent from others.

Colouring by similarity to a reference instead, rather than by cluster
membership, makes graded divergence visible where cluster identity would
only say "different":

```{r plot_sim, fig.width=9, fig.height=3}
plot(msa, fill_type = "repeat_sim",
    tal_sim = cmp$tale_distances, domain_sim = cmp$domain_distances)
```

And with a consensus panel attached, the column-by-column question "does
this array agree with the group" is answered directly above the alignment
itself:

```{r plot_consensus, fig.width=9, fig.height=3.3}
plot(msa, consensus = TRUE,
    tal_sim = cmp$tale_distances, domain_sim = cmp$domain_distances)
```

# Next

Once related arrays are aligned, the biologically motivated next question
is what DNA sequence their RVDs are predicted to bind -- covered in the
final article of this series.
