---
title: "4. TALE target predictions with tantale"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{4. TALE target predictions with tantale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r package, warning=FALSE, message=FALSE}
library(tantale)
library(dplyr)
```

The previous three articles found TALE arrays, grouped them, and aligned
related ones. This last article asks the question a TALE's RVD sequence
exists to answer: what DNA sequence does it bind? Each RVD contacts one
base of the target through a well-characterised, largely one-to-one code
(`NI` prefers A, `HD` prefers C, `NG` prefers T, `NN` prefers G or A, and so
on), so a whole RVD sequence predicts an *EBE* -- an Effector Binding
Element -- as a short DNA motif with one position per RVD.

tantale wraps two independent predictors that use this code differently
enough to be worth contrasting: [TALVEZ](https://doi.org/10.1371/journal.pone.0068464),
a position-weight-matrix scan, and
[PrediTALE](https://doi.org/10.1371/journal.pcbi.1007206), which also models
where a real EBE tends to sit relative to a transcription start site. Both
are wrapped as `talvez()` and `preditale()`, taking the same inputs and
returning the same column layout, so their predictions can be compared
directly rather than translated by hand.

# Predicting against a real promoter set

RVD sequences come straight from a `tales` object via `tales_rvd_strings()`
-- the same projection used for alignment input, just with the termini
dropped by default, since prediction concerns the repeat domain only. This
reuses the classification result from the second article:

```{r discover_and_compare, message = FALSE, warning = FALSE}
out <- fs::dir_create(file.path(tempdir(), "tale_target_prediction"))
genome_files <- c(
  MAI1 = system.file("extdata", "MAI1.fa", package = "tantale", mustWork = TRUE),
  BAI3 = system.file("extdata", "BAI3.fa", package = "tantale", mustWork = TRUE),
  PXO86 = system.file("extdata", "PXO86.fa", package = "tantale", mustWork = TRUE)
)

all_tales <- lapply(names(genome_files), function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir))
  tales_from_telltale(strain_dir) %>%
    mutate(array_id = paste0(strain, "_", array_id))
}) %>%
  suppressWarnings() %>%
  bind_rows() %>%
  tales(sanitize = TRUE)

rvds <- tales_rvd_strings(all_tales)
length(rvds)
```

The target sequences are promoter regions of three clade III *SWEET* genes
-- a well-studied class of rice susceptibility genes that TALEs from
*Xanthomonas oryzae* are known to target -- from several rice varieties:

```{r subj_file}
subj_file <- system.file("extdata", "cladeIII_sweet_promoters.fasta",
                         package = "tantale", mustWork = TRUE)
Biostrings::readBStringSet(subj_file) %>% names()
```

## TALVEZ

```{r talvez, message = FALSE}
talvez_preds <- talvez(rvd_seqs = rvds, subj_file = subj_file, opt_param = "-t 0 -l 19")
nrow(talvez_preds)
talvez_preds %>% arrange(desc(score)) %>% head(5)
```

## PrediTALE

```{r preditale, message = FALSE, results = "hide"}
preditale_preds <- preditale(rvd_seqs = rvds, subj_file = subj_file)
```

```{r preditale_show}
nrow(preditale_preds)
preditale_preds %>% arrange(desc(score)) %>% head(5)
```

PrediTALE additionally reports a `pval` per prediction, since it models the
null distribution of scores directly rather than only ranking candidates
against each other -- worth using when the question is "is this a credible
site at all" rather than only "which of these is the best".

# Visualising predictions against the target sequence

`plot_target_preds()` draws predicted RVD-to-base correspondences directly
against the target DNA, coloured by how well each RVD's known base
preference matches the base actually predicted underneath it -- so a
"good" prediction is visually distinguishable from one held up by only one
or two well-matching RVDs. Restrict to one promoter and a window within it
to keep the figure readable:

```{r plot_preds, fig.width=12, fig.height=4}
best_on_target <- preditale_preds %>%
  filter(subjSeqId == "SWEET14p_Nipponbare_Sense") %>%
  arrange(desc(score)) %>%
  head(5)

plot_target_preds(preds = best_on_target, subj_file = subj_file,
                  filter_range = "SWEET14p_Nipponbare_Sense:300-500")
```

Each RVD is printed in a box over the base it is predicted to contact;
predictions on the sense strand are drawn above the sequence and those on
the antisense strand below it, since a TALE can bind either strand of its
target.

# Closing the loop

These four articles followed one path through a TALE study: `tell_tales()`
to find arrays (correcting them, and checking for what correction could not
fix, along the way), `tales_compare()` and `tales_group()` to relate them,
`tales_align()` to see that relatedness directly, and `talvez()`/
`preditale()` to ask what they do. None of these steps requires the ones
before it in code -- each function documented here takes ordinary `tales`
objects or plain sequences -- but together they cover most of what a study
of TALE diversity in a set of genomes actually needs.
