---
title: "Classifying TALE sequences from genomes"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

```{r site_cache, include = FALSE}
# Shared with tale_msa.qmd, tales_msa_class.qmd and tale_target_prediction.qmd,
# which read these files rather than recomputing (see the getting-started
# article). Not content-hashed: delete _cache/ before rebuilding the site
# after changing tell_tales()/tales_compare_distal(). Grouping is cheap and
# always recomputed, so its silhouette plot is drawn on every render.
site_cache_dir <- fs::dir_create("_cache")
discovery_cache <- file.path(site_cache_dir, "discovery.rds")
compare_cache   <- file.path(site_cache_dir, "compare.rds")
group_cache     <- file.path(site_cache_dir, "group.rds")
```

The [previous article](tale_mining.html) covered finding TALE loci in one
genome at a time. A real study usually has several genomes, and the
interesting question becomes "which TALEs, across strains, are versions
of the same thing": alleles of one locus, related by descent. This
article covers that comparison:
`tales_compare_distal()` to quantify how alike every pair of arrays and
domains is, and `tales_group_hclust()`/`tales_group_kmedoids()` to turn
that into discrete groups.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

## Discovering TALEs across several genomes

This is a fresh discovery run, independent of the previous article, over
three related genomes: MAI1, BAI3, and the best-corrected version of
BAI3-1-1 from
[the correction section of the mining article](tale_mining.html#sec-best-correction).

::: {.callout-note}
## Why only three genomes
tantale ships a fourth sample genome, PXO86, a distantly related Asian
outgroup strain. It is left out here deliberately: its TALE repertoire
barely overlaps the African strains' at all, so including it turns a
clean set of one-locus-per-strain groups into a mix of real cross-strain
groups and PXO86-only paralog clusters, which illustrate a different
question from the one this article is asking.
:::

Discovery runs independently per genome, so it is written as one function
applied to each genome name in turn:

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_classification"))
genome_files <- c(
  MAI1       = system.file("extdata", "MAI1.fa",       package = "tantale", mustWork = TRUE),
  BAI3       = system.file("extdata", "BAI3.fa",       package = "tantale", mustWork = TRUE),
  `BAI3-1-1` = system.file("extdata", "BAI3-1-1.fa",   package = "tantale", mustWork = TRUE)
)
```

```{r discover_one}
discover_strain <- function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir,
                       cterm_min_score = 300,
                       correct_array = TRUE, max_comparisons = 50))
  tales_from_telltale(strain_dir) |>
    mutate(array_id = paste0(strain, "_", array_id), strain = strain)
}
```

`correct_array = TRUE` with `max_comparisons = 50` is the route
[the mining article](tale_mining.html#sec-best-correction) found clean
for BAI3-1-1. Run on MAI1 and BAI3 too, it costs a little extra time;
neither needed correcting.

```{r discover_all, message = FALSE}
if (fs::file_exists(discovery_cache)) {
  all_tales <- readRDS(discovery_cache)
} else {
  all_tales <- lapply(names(genome_files), discover_strain) |>
    suppressWarnings() |>
    bind_rows()
}
```

`array_id` is prefixed by strain before the three objects are combined,
since `tell_tales()` numbers regions independently within each genome:
without the prefix, `MAI1`'s `ROI_00001` and `BAI3`'s `ROI_00001` would
collide.

```{r sanitize_all}
all_tales <- tales(all_tales, sanitize = TRUE)
saveRDS(all_tales, discovery_cache)
n_distinct(all_tales$array_id)
```

Nine arrays each come from MAI1 and BAI3, and eight from BAI3-1-1.

## Quantifying relatedness: `tales_compare_distal()`

`tales_compare_distal()` is tantale's R implementation of the DisTAL
comparison used by the original QueTAL suite: it assigns every distinct
domain sequence a `dom_code` (see [the `tales` class article](tales_class.html)
for what that means), computes pairwise distances between distinct
domains, and from those, distances between whole arrays.

```{r tales_compare, message = FALSE}
if (fs::file_exists(compare_cache)) {
  cmp <- readRDS(compare_cache)
} else {
  cmp <- tales_compare_distal(all_tales, aln_method = "DECIPHER", ncores = 4)
  saveRDS(cmp, compare_cache)
}
names(cmp)
```

```{r tale_distances_show}
cmp$tale_distances
```

### Does the choice of backend matter? {#sec-backends}

Comparing domain sequences, the step just run, is the expensive part,
and three backends can do it: `"DECIPHER"` (the default, used above),
`"Biostrings"`, and `"mmseq2"`. All three score pairwise protein
alignments between domains, so the choice should depend on what is
installed and how large the dataset is, and should not change the
answer. A small independent fixture, on which all three run quickly,
checks that:

```{r backend_fixture}
small <- tales_from_telltale(system.file("extdata", "tellTaleExampleOutput",
                                         package = "tantale"))
```

```{r backend_runs, message = FALSE}
cmp_decipher   <- tales_compare_distal(small, aln_method = "DECIPHER")
cmp_biostrings <- tales_compare_distal(small, aln_method = "Biostrings")
cmp_mmseq2     <- tales_compare_distal(small, aln_method = "mmseq2")
```

`as.matrix()` on a `tale_distances` object always returns ids sorted the
same way, so the three backends' outputs line up without any further
work:

```{r backend_correlate}
backend_dists <- tibble::tibble(
  DECIPHER   = as.vector(as.matrix(cmp_decipher$tale_distances)),
  Biostrings = as.vector(as.matrix(cmp_biostrings$tale_distances)),
  mmseq2     = as.vector(as.matrix(cmp_mmseq2$tale_distances))
)
cor(backend_dists)
```

All three agree closely on this fixture (correlations of 0.986 and
above), so the default gives up nothing in accuracy here, at least on
data this size.

## Allocating arrays to groups: `tales_group_hclust()` and `tales_group_kmedoids()`

Two clustering methods turn `tale_distances` into discrete groups. Both
take the `tales` object the distances were computed from, which lets
each check that the two actually correspond, and both return it with the
result attached as a `group` column. Each carries its own
method-specific arguments (`k_range`/`seed` for k-medoids, `plot_tree`
for hclust).

`tales_group_kmedoids()` fits `cluster::pam()` for each candidate group
count in `k_range` and plots the average silhouette value of each fit.
`k = "auto"` picks the elbow of that curve, the point after which adding
groups stops paying off, a feature tantale adds on top of what DisTAL
itself offers:

```{r tales_group}
grouped <- tales_group_kmedoids(cmp$tales, cmp$tale_distances,
                                k_range = 2:20, k = "auto")
saveRDS(grouped, group_cache)
```

```{r group_sizes}
group_sizes <- grouped |> distinct(array_id, group) |> count(group) |> arrange(desc(n))
group_sizes
```

The automatic pick, in red on the silhouette plot, is
`r n_distinct(grouped$group)` groups. The curve is nearly flat just past
it, so a few more groups would fit about as well; the elbow takes the
smallest of those. The result is clean: `r sum(group_sizes$n == 3)` of
`r nrow(group_sizes)` groups have exactly three members, one from each
strain, the one-locus-per-strain pattern this comparison is meant to
recover. The remaining group has only a MAI1 and a BAI3 member, matching
BAI3-1-1's one fewer array.

`tales_group_hclust()` instead cuts a dendrogram at a height chosen to
yield exactly `k` groups, and with `plot_tree = TRUE` draws it. The tree
is a useful check on whether the chosen `k` looks reasonable:

```{r}
#| label: fig-tale-dendrogram
#| fig-cap: "Hierarchical clustering of the same three-genome comparison, cut at the k tales_group_kmedoids() picked automatically above."
#| fig-width: 7
#| fig-height: 5
invisible(tales_group_hclust(cmp$tales, cmp$tale_distances,
                             k = n_distinct(grouped$group),
                             plot_tree = TRUE))
```

## An overview across strains: `talomes_heatmap()`

With arrays assigned to groups, one natural summary is which RVD sequence
variant each strain carries in each group. A strain's *talome*, by
analogy to its genome, is the whole complement of TALEs it carries.

```{r talome_annotation}
tale_annotation <- tibble::tibble(
  array_id = names(tales_rvd_strings(grouped)),
  rvdseq = as.character(tales_rvd_strings(grouped))
) |>
  left_join(distinct(grouped, array_id, group, strain), by = "array_id")
```

```{r}
#| label: fig-talomes-heatmap
#| fig-cap: "RVD sequence variant carried by each strain, in each classification group."
#| fig-width: 8
#| fig-height: 6
talomes_heatmap(tale_annotation, group_col = "group", strain_col = "strain",
                rvd_col = "rvdseq")
```

Each cell in @fig-talomes-heatmap is one strain's RVD sequence variant in
one group, coloured by the variant's rank within that group (the darkest
is the most common); the `#` after each group name counts its distinct
variants. A grey cell means the strain has no member in that group, and
a cell split into several colours would mean a strain carries more than
one variant in the group. Here, BAI3 and BAI3-1-1 carry the same variant
in every group they share. MAI1 carries that same variant in four groups
and a different one in the other five.

## A different lens: comparing predicted binding specificity

Everything above compares TALEs by their *sequence*, domain by domain,
via DisTAL. A TALE's RVDs also predict, base by base, the DNA sequence it
binds, and two arrays can be compared on that prediction directly instead:
`tales_to_universalmotif()` converts each array's RVD sequence into a
position weight matrix (PWM) of DNA-binding preferences, using the same
per-RVD specificity weights `tales_predict_targets()` relies on
(`rvd_dna_specificity`), and `tales_compare_functal()` scores every pair
of PWMs with `universalmotif::compare_motifs()`:

```{r functal_compare, message = FALSE}
picked_group <- grouped |> filter(group == 6)
functal_dissim <- tales_compare_functal(picked_group)
functal_dissim
```

BAI3 and BAI3-1-1 (the same genomic background, with *talC* deleted in
the latter) predict the *identical* binding specificity for this array.
MAI1's version carries the same twelve RVDs followed by four more, and
its dissimilarity of 0.25 reflects those four extra positions. This
result comes from DNA-binding prediction alone, independently of the
DisTAL comparison above, and agrees with it.

Once arrays are a list of real `universalmotif` motifs, that package's own
plotting functions apply directly. `motif_tree()` draws the same kind of
relatedness tree as `tales_group_hclust()`, built from binding-specificity
distance instead of protein-domain distance:

```{r}
#| label: fig-functal-tree
#| fig-cap: "Predicted-binding-specificity relatedness tree for the same three arrays, from universalmotif::motif_tree(). Topology only: branch lengths are not drawn to scale."
#| fig-width: 6
#| fig-height: 3
motifs <- tales_to_universalmotif(picked_group)
universalmotif::motif_tree(motifs, layout = "rectangular", linecol = "none",
                           labels = "name", legend = FALSE) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.02, 0.6)))
```

`view_motifs()` renders the PWMs themselves as sequence logos, showing the
specificity model behind the distances. MAI1's four extra positions are
visible at the end:

```{r}
#| label: fig-functal-logos
#| fig-cap: "Predicted DNA-binding preference at each RVD position, one logo per array."
#| fig-width: 8
#| fig-height: 5
universalmotif::view_motifs(motifs)
```

`tales_to_universalmotif()`'s output is a plain list of `universalmotif`
objects, so the rest of that package's toolkit is reachable the same
way: `universalmotif::scan_sequences()` can search a promoter sequence for
predicted binding sites directly from one of these PWMs, an alternative
angle on what `tales_predict_targets()` does in
[the target-prediction article](tale_target_prediction.html);
`universalmotif::merge_motifs()` can build one consensus binding-site
model for a whole group, complementing what `tales_consensus()` already
does at the RVD/`dom_code` layer; and `universalmotif::average_ic()` is
worth checking before comparing motifs at all, since a PWM built from an
array with many unresolved RVDs carries little real information and can
distort a comparison (`r round(universalmotif::average_ic(motifs[[1]]), 2)`
bits here, above the 0.25-bit minimum `compare_motifs()` applies by
default). Statistical significance testing for these comparisons
(`compare_motifs()`'s `compare.to`/`max.p` arguments) is deliberately not
shown here: `universalmotif`'s default null distributions are calibrated
on real transcription-factor motifs, and applying them to TALE-derived
PWMs would give uncalibrated p-values. A TALE-specific calibration is
separate work.

## Next

Grouping tells you *which* arrays are related. Seeing exactly *how*
(which repeats match, which are inserted or deleted relative to one
another) needs an alignment, which is the subject of
[the next article](tale_msa.html).
