---
title: "Mining TALE sequences in genomes"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

This is the first of a set of articles walking through a typical study of
TALE diversity: finding TALE genes in a genome (here), classifying the
arrays found into groups of related sequences, aligning those groups, and
predicting the DNA targets of the TALEs they contain. The other articles
in the set are side branches: deep dives into a single class (`tales`,
`tales_msa`) and one case study on naturally truncated TALEs. Each is
linked from the point where it becomes relevant.

This article covers TALE discovery with AnnoTALE and `tell_tales()`. Because of their repetitive nature *tal* genes cannot be assembled from short read sequencing technologies data. This is why now most of the genomes of *tal* bearing bacteria are sequenced with long read technologies which may be error prone. Those assemblies are not always clean, and
`tell_tales()` can sometimes be fooled. tantale offers two ways to correct
frameshifts. They work differently, neither is guaranteed to succeed, and
`tales_anomalies()` tells you which arrays still need attention after
either one.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

::: {.callout-note}
## The genomes used throughout
Four *Xanthomonas oryzae* genomes appear across this set of articles:
**MAI1**, **BAI3** and **PXO86** are "clean" assemblies, and **BAI3-1-1**
is a deliberately error-prone one: an assembly of the same BAI3
background with a deleted *tal* locus, carrying real sequencing/assembly
artefacts in its TALE loci. A mining pipeline meets this kind of input
routinely in practice, and it is what the second half of this article is
about.
:::

## AnnoTALE, the standard tool

[AnnoTALE](https://doi.org/10.1038/srep21077) is the standard tool for
finding and annotating *tal* genes in bacterial genome sequences. Its
`predict` stage locates TALE genes in a genome and flags those likely to
be pseudogenes. Its `analyze` stage splits each predicted protein into
its N-terminal region, repeats and C-terminal region, and reads the RVD
of every repeat. A third stage, `build`, assigns TALEs to the classes of
the AnnoTALE nomenclature. tantale runs the first two with
`run_annotale_predict()` and the third with `run_annotale_build()`, and
`tales_from_annotale()` turns the output of `predict` into a `tales`
object. tantale ships AnnoTALE's output for four TALEs of BAI3:

```{r annotale_example}
annotale_example <- system.file("extdata", "annotaleExampleOutput", package = "tantale")
bai3_sample <- tales_from_annotale(annotale_example)
bai3_sample
```

Each terminal region is searched with the profile HMM of the TALE N- or
C-terminal domain, and marked `NTERM` or `CTERM` when it matches (`XXXXX`
otherwise). A match shows that the region is related to a TALE terminus.
A shorter terminus that matches is coded the same way, although it has
probably lost part of its function (see
[Genuine truncTALEs](trunctale_correction.html#sec-two-trunctales)).
On a whole genome, the same object comes from two calls
(not run here, as `predict` takes a couple of minutes per genome):

```{r annotale_genome, eval = FALSE}
mai1_annotale_dir <- file.path(tempdir(), "MAI1_annotale")
run_annotale_predict(tantale_genome("MAI1"),
                     output_dir = mai1_annotale_dir)
mai1_annotale <- tales_from_annotale(mai1_annotale_dir)
```

On the clean BAI3 assembly, AnnoTALE predicts nine TALEs, none flagged. On BAI3-1-1, the error-prone assembly
of the same background, it predicts eight and flags all eight as
putative pseudogenes. `tell_tales()` is built for such genomes. It
locates candidate arrays with DNA profiles of the three TALE regions
before any ORF is predicted, reports the DNA evidence for each array
(which terminus profiles matched, how much of the region the longest ORF
covers), can correct frameshifts in each array before AnnoTALE `analyze`
reads it, and keeps per-array reports and terminus alignments for
inspection. The rest of this article uses `tell_tales()`.

## Finding TALE loci in genomic DNA

`tell_tales()` searches a genome with `nhmmer`, using three profile HMMs
tuned to the N-terminus, the repeat unit, and the C-terminus of a TALE
CDS. Hits are merged, grouped into candidate arrays by proximity, and each
array's longest ORF is handed to AnnoTALE to split into parts and call its
RVD sequence. `NTERM`/`CTERM` markers are added at either end of the RVD
string wherever a terminus was identified this way, so a downstream
alignment knows where an array actually starts and ends.

A scratch directory holds everything this set of articles builds:

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_mining"))
```

Start with MAI1, a clean assembly, to see what discovery looks like when
nothing goes wrong:

```{r mai1_path}
mai1_fa <- tantale_genome("MAI1")
mai1_dir <- file.path(out, "MAI1")
```

```{r discover_mai1, message = FALSE}
invisible(tell_tales(subject_file = mai1_fa, output_dir = mai1_dir))
```

The result on disk is a directory of reports and fasta files. The result
in R is a `tales` object, with one row per part (repeat or terminus):

```{r tales_from_mai1}
mai1 <- suppressWarnings(tales_from_telltales(mai1_dir))
mai1
```

::: {.callout-tip}
## What a `tales` object actually is
Rows, columns, subsetting, and the `dom_code` column that the rest of
tantale is built on are covered in depth in
[the `tales` class article](tales_class.html) -- read it once this
walkthrough has given you an object to look at.
:::

`plot()` on a `tales` gives a compact overview of what was found: every
array's parts, positioned within the array, outlined by domain type and
filled by amino-acid length, with the RVD printed on each repeat.

```{r}
#| label: fig-mai1-composition
#| fig-cap: "Domain composition of the nine TALE arrays found in MAI1."
#| fig-width: 9
#| fig-height: 4
plot(mai1)
```

@fig-mai1-composition already shows something worth noticing: array
lengths vary, from 14 to 26 repeats (counting the final half-repeat), and
each array's parts sit side by side with no gaps, because nothing has
been aligned yet. `position_in_array` just counts parts within each array
independently.

## What `tell_tales()` writes to disk

Besides the RVD sequences and the TALE ORFs themselves, `tell_tales()`
writes three tab-separated reports, which `tales_from_telltales()` reads
back to build the object above. They are worth knowing about directly,
since they carry a few things the `tales` object does not.

```{r report_table, echo = FALSE}
knitr::kable(tibble::tribble(
  ~file, ~`one row per`, ~`worth knowing`,
  "`hits_report.tsv`", "raw `nhmmer` hit", "`codon_count`, `frameshift_count` per hit -- before any merging or correction",
  "`domains_report.tsv`", "domain, after AnnoTALE", "what AnnoTALE actually parsed out of the ORF",
  "`array_report.tsv`", "candidate array", "`nterm_dna_hit`/`cterm_dna_hit`, `nterm_aa_hit`/`cterm_aa_hit`, `has_aberrant_repeat`, `orf_coverage`, and (with correction) `predicted_ins_count`/`predicted_dels_count`"
))
```

`array_report.tsv` is the one this article leans on most, so it is worth
reading directly:

```{r array_report}
array_report <- readr::read_tsv(file.path(mai1_dir, "array_report.tsv"),
                                show_col_types = FALSE)
array_report %>%
  select(array_id, nterm_dna_hit, cterm_dna_hit, longest_orf_length, orf_coverage)
```

`orf_coverage` is the longest ORF found, as a percentage of the whole
candidate array region length; a clean array's ORF covers 91-93% of it here.
That number is about to matter a great deal.

The report has one more row than the `tales` object has arrays.
`ROI_00005` is a candidate region with no terminus hit (`nterm_dna_hit`
and `cterm_dna_hit` both `FALSE`) and no
ORF length; it yields no parts, so it is absent from
the object and from @fig-mai1-composition. Using `tell_tales()` with a more stringent
`min_dna_hits` or `min_array_length` parameter values would probably have got rid of it right away.

## When discovery goes wrong: detecting anomalies {#sec-anomalies}

MAI1 is a good assembly, and every array in the `tales` object above came
back complete. That is not guaranteed. Running the same search on
BAI3-1-1, the deliberately error-prone assembly, shows what an assembly
artefact does to the pipeline.

```{r bai311_path}
bai311_fa <- tantale_genome("BAI3-1-1")
bai311_raw_dir <- file.path(out, "BAI3-1-1_raw")
```

```{r discover_bai311_raw, message = FALSE}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_raw_dir,
                     cterm_min_score = 300))
```

A first sign that there is something unexpected with these TALEs predictions are the warnings obtained when importing `tell_tales` results as a `tales` object:

```{r tales_from_bai311_raw, warning = TRUE}
bai311_raw <- tales_from_telltales(bai311_raw_dir)
```

`tales_anomalies()` lists the arrays concerned and why. It reports every
array that is not a standard TALE (an N-terminus, one or more repeats and
a C-terminus, both termini matching the profile of their TALE domain),
and every array whose content is inconsistent (missing sequence data,
impossible terminus arrangements, coordinate disagreements):

```{r anomalies_raw}
tales_anomalies(bai311_raw)

```

`ROI_00003` and `ROI_00005` have no repeat at all: AnnoTALE could not
parse a repeat-array structure out of the longest predicted ORF. Five
other arrays have a terminus coded `XXXXX`, which does not match the
profile of its TALE domain. Their `orf_coverage` shows why:

```{r orf_coverage_raw}
readr::read_tsv(file.path(bai311_raw_dir, "array_report.tsv"),
                show_col_types = FALSE) %>%
  select(array_id, nterm_dna_hit, cterm_dna_hit,
         longest_orf_length, orf_coverage, rvd_string) %>%
  arrange(array_id) %>%
  knitr::kable()
```

`ROI_00003` and `ROI_00005` predicted ORFs cover 19% and 23% of their candidate region.

No array in this assembly reaches MAI1's 91-93%: the
others fall between 38% and 70%. This is unusual, and `tales_anomalies()` flags every one of them except `ROI_00002`.
A close look at the 'array_report.tsv' table indicates that loci such as `ROI_00001` and `ROI_00007` have a predicted ORF encoding very few repeats (4-6), and only repeats (see `rvd_string`), even if both termini profiles were found at the DNA level (`nterm_dna_hit` and `cterm_dna_hit` are both `TRUE`).

Because, in practice this barely happens in high quality genomes, in the absence of any further evidence, this could be explained by in/del(s) along the CDS that completely break them.


A premature stop could also be
genuine: some strains carry naturally truncated TALEs (see
[Genuine truncTALEs and frameshift correction](trunctale_correction.html)).
In BAI3-1-1 the corrections below settle it: once repaired, both arrays
recover the coverage of a clean MAI1 array, as expected of assembly
frameshifts.

## Correcting frameshifts, two ways

tantale offers two distinct routes to a corrected sequence: one corrects
candidate *arrays* after they have been found, from inside `tell_tales()`;
the other corrects the *genome* before discovery starts. Both can fail,
in different ways, so @sec-anomalies's check is worth repeating after
either one. Run on the same genome below, **they disagree about which
array still needs help**.

### Correcting inside `tell_tales()` {#sec-best-correction}

`correct_array = TRUE` runs `DECIPHER::CorrectFrameshifts()` on each
candidate array, aligning it against a shipped reference set of real TALE
proteins and inferring the indels needed to remove premature stops.
`max_comparisons` caps how many of these references each array is
aligned against; @sec-max-comparisons explains its default, 50.

```{r correct_inside_path}
bai311_corr_dir <- file.path(out, "BAI3-1-1_corrected")
```

```{r correct_inside, message = FALSE, warning = TRUE, results = "hide"}
t_corr <- system.time(
  tell_tales(subject_file = bai311_fa, output_dir = bai311_corr_dir,
             cterm_min_score = 300, correct_array = TRUE)
)
```

The warnings come from the correction itself. Where
`DECIPHER::CorrectFrameshifts()` inserts a base to restore the reading
frame, it writes an `N`, and `tell_tales()` turns it into a `C` so that
AnnoTALE can translate the sequence.

Importing the result as a `tales` object raises no warning, and
`tales_anomalies()` finds nothing:

```{r tales_from_bai311_corr}
bai311_corr <- tales_from_telltales(bai311_corr_dir)
tales_anomalies(bai311_corr)
```

To see what changed, `tales_bind()` combines the uncorrected and the
corrected arrays into one object, which `plot()` draws with a single
legend. The array identifiers must be unique across the objects it
binds, so each object's get a prefix; a `method` column records the
correction each array went through (a factor, which sets the order of
the panels), and `roi` keeps the original identifier:

```{r bind-raw-corrected}
#| warning: false
method_levels <- c("uncorrected", "correct_tales()", "correct_array = TRUE")
tag_method <- function(x, method, prefix) {
  mutate(x, method = factor(method, levels = method_levels),
         roi = array_id, array_id = paste0(prefix, "_", array_id))
}
bai311_compared <- tales_bind(
  tag_method(bai311_raw, "uncorrected", "raw"),
  tag_method(bai311_corr, "correct_array = TRUE", "corr")
)
```

`tales_bind()` warns again about the anomalies of the uncorrected arrays;
that warning is hidden here. `facet_by = "method"` draws one panel per
method:

```{r plot-raw-vs-corrected}
#| fig-width: 12
#| fig-height: 8
plot(bai311_compared, facet_by = "method")
```

The uncorrected talome is extremely messy: most arrays have no canonical
N- and C-termini, and `ROI_00003` and `ROI_00005` essentially encode
shorter N-termini only. In contrast, the corrected arrays all display both
termini with a canonical length, which suggests that these TALEs are likely
to have a standard transcription activation function.

### Correcting the genome, before discovery

`correct_tales()` takes a different, faster route. It wraps the Java
`TALEcorrection` tool, which scans the *whole input genome* against
profile HMMs and repairs the frameshifts it finds, before `tell_tales()`
ever runs. As a genome-wide pre-processing step that needs no candidate
arrays, it is worth trying first on a large or especially error-prone
assembly (an ONT-sequenced genome, say).

```{r correct_outside_path}
bai311_java_fa <- file.path(out, "BAI3-1-1_java_corrected.fa")
```

```{r correct_outside, message = FALSE}
t_java <- system.time(
  corrections <- correct_tales(uncorrected_path = bai311_fa,
                               corrected_path = bai311_java_fa,
                               return_corrections = TRUE)
)
```

`correct_tales()` reports every edit it made, genome-wide, as a table of
positions and inserted or deleted bases. This genome needed
`r nrow(corrections)`:

```{r corrections_head}
head(corrections, 4)
```

`tell_tales()` then runs on the corrected genome exactly as it would on
any other, with no `correct_array`, because the correction already
happened upstream:

```{r discover_after_java_correct_path}
bai311_java_dir <- file.path(out, "BAI3-1-1_java_tell_tales")
```

```{r discover_after_java_correct, message = FALSE}
t_java <- t_java + system.time(
  tell_tales(subject_file = bai311_java_fa, output_dir = bai311_java_dir,
             cterm_min_score = 300, correct_array = FALSE)
)
bai311_java <- tales_from_telltales(bai311_java_dir)
```

```{r anomalies_java}
tales_anomalies(bai311_java)
```

`correct_tales()` fixes most arrays but fails to repair the N-termini of `ROI_00006` and
`ROI_00009`, which still do not match the TALE N-terminal profile.

Compared with the correction made inside `tell_tales()`, bound the same
way:

```{r plot-java-vs-corrected}
#| warning: false
#| fig-width: 12
#| fig-height: 8
tales_bind(tag_method(bai311_java, "correct_tales()", "java"),
           tag_method(bai311_corr, "correct_array = TRUE", "corr")) |>
  plot(facet_by = "method")
```

The two corrections give the same repeats everywhere except at the start
of `ROI_00006` and `ROI_00009`. After `correct_tales()`, both begin with a
24-aa N-terminus followed by a 32-aa repeat read as `nv`, and lack the
`NN` and `HD` repeats that open these arrays after correction inside
`tell_tales()`. The N-termini of five other arrays differ by one residue:
288 aa after `correct_tales()`, 287 aa after correction inside
`tell_tales()`.

Binding all three runs gives the lengths of every part at a glance, by
domain type (columns) and correction method (rows), one colour per
candidate region:

```{r part-lengths}
#| warning: false
#| fig-width: 12
#| fig-height: 7
tales_bind(tag_method(bai311_raw, "uncorrected", "raw"),
           tag_method(bai311_java, "correct_tales()", "java"),
           tag_method(bai311_corr, "correct_array = TRUE", "corr")) |>
  mutate(aa_length = factor(nchar(aa_seq), sort(unique(nchar(aa_seq)))),
         domain_type = factor(domain_type, c("N-terminus", "repeat", "C-terminus"))) |>
  ggplot2::ggplot(ggplot2::aes(aa_length, fill = roi)) +
  ggplot2::geom_bar(position = ggplot2::position_dodge(preserve = "single")) +
  ggplot2::facet_grid(method ~ domain_type, scales = "free", space = "free_x") +
  ggplot2::scale_fill_manual(values = unname(grDevices::palette.colors(8, "Okabe-Ito"))) +
  ggplot2::labs(x = "Length (aa)", y = "Parts", fill = "Region") +
  ggplot2::theme_light()
```

Without correction, the termini take almost any length, from 24 to 288 aa
for the N-termini and from 2 to 226 aa for the C-termini, and two repeats
are only 11 and 14 aa long. After either correction, every C-terminus is
278 aa long, and the repeats come back to the canonical 34 aa, the final
20-aa half-repeats and a few of 32 and 33 aa. The N-termini settle at 287
or 288 aa, except the two 24-aa N-termini that `correct_tales()` leaves.


### What is the best method for correcting the DNA sequences of TALE arrays?

On this genome, correction inside `tell_tales()` gives the better
result: `r n_distinct(bai311_corr$array_id) - n_distinct(tales_anomalies(bai311_corr)$array_id)`
of `r n_distinct(bai311_corr$array_id)` arrays come out as standard
TALEs, against
`r n_distinct(bai311_java$array_id) - n_distinct(tales_anomalies(bai311_java)$array_id)`
of `r n_distinct(bai311_java$array_id)` after `correct_tales()`. It is
also about as fast: on the machine that built this page it took
`r round(t_corr[["elapsed"]])` seconds. The Java route needs two steps,
`correct_tales()` and then the `tell_tales()` run that finds the arrays
in the corrected genome, and together they took
`r round(t_java[["elapsed"]])` seconds.
One genome is a small sample, however. `correct_tales()` repairs the
whole genome in one pass, and the corrected genome can serve other
analyses as well. On a new, error-prone assembly, running both routes
and comparing their `tales_anomalies()` costs a few minutes and shows the
arrays on which they agree.

### `max_comparisons` sets the depth of the search {#sec-max-comparisons}

`max_comparisons` is the main control on how long correction inside
`tell_tales()` takes. `DECIPHER::CorrectFrameshifts()` ranks the
references by a quick distance and aligns each array against the closest
`max_comparisons` of them. When the cap is too low, an array whose good
references rank further down is corrected against a poor one. BAI3-1-1
has one such array, `ROI_00001`. Measured once on this genome, against the
494 shipped references:

| `max_comparisons` | seconds | standard TALEs | `ROI_00001` |
|---|---|---|---|
| no correction | 19 | 0 of 8 | 4 repeats, neither terminus matched |
| 2 to 5 | 19-22 | 7 of 8 | N-terminus unmatched, 19 of its 26 repeats |
| 10, 20 | 26, 34 | 7 of 7 | not parsed by AnnoTALE |
| 50 (default) | 61 | 8 of 8 | N-terminus, 26 repeats, C-terminus |
| 100 | 101 | 8 of 8 | as at 50 |
| all 494 | 455 | 8 of 8 | as at 50 |

At 2 to 5, `tales_anomalies()` flags `ROI_00001`. At 10 and 20,
AnnoTALE cannot split the corrected ORF into parts, so the array never
reaches the `tales` object, and `tales_from_telltales()` warns about it:

```{r max_comparisons_20_path}
bai311_mc20_dir <- file.path(out, "BAI3-1-1_corrected_20")
```

```{r max_comparisons_20, message = FALSE, warning = FALSE, results = "hide"}
invisible(tell_tales(subject_file = bai311_fa, output_dir = bai311_mc20_dir,
                     cterm_min_score = 300, correct_array = TRUE,
                     max_comparisons = 20))
```

```{r max_comparisons_20_import, warning = TRUE}
bai311_mc20 <- tales_from_telltales(bai311_mc20_dir)
```

On MAI1, BAI3 and PXO86, the other genomes shipped with the package, 50
gives the same corrected sequences as the full search. With a reference
set of your own, compare a run at the default with one at
`max_comparisons = NULL` before relying on the cap.

## Moving on with what you have

`bai311_corr` needs none of this: it is already clean. A route that stops
short of that, such as `correct_tales()` above (which left the N-termini
of `ROI_00006` and `ROI_00009` unmatched), still needs a decision about
what to do with what remains flagged. `tales(x, sanitize = TRUE)`
drops it so an analysis can proceed on what is clean, with a warning
naming each dropped array and the reason:

```{r sanitize, warning = TRUE}
bai311_clean <- tales(bai311_java, sanitize = TRUE)
n_distinct(bai311_java$array_id) - n_distinct(bai311_clean$array_id)
```

::: {.callout-note}
## What the rest of this set of articles uses
From here on, "BAI3-1-1" means `bai311_corr`, the correction made
inside `tell_tales()` (@sec-best-correction), alongside
MAI1 and BAI3. PXO86, the fourth genome shipped with the package, is left
out of the classification and alignment articles that follow: it is a
distant Asian outgroup whose TALE repertoire barely overlaps the African
strains', and including it turns a clean set of one-ortholog-per-strain
groups into a mix of real cross-strain groups and PXO86-only paralog
clusters that do not illustrate the same thing.
:::

## Next

The [next article](tale_classification.html) picks up from a set of
`tales` objects like the ones built here, across several genomes this
time, and classifies their arrays into groups of related sequences with
`tales_compare_distal()` and `tales_group_hclust()`/`tales_group_kmedoids()`.
