---
title: "The tales class"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every function in tantale that describes, compares or plots a set of TALEs
takes or returns a `tales` object. This article is a deep dive into what
that object is, how to build one, and what its most important column,
`dom_code`, actually means.

::: {.callout-note}
## Where this fits
This article extends [Mining TALE sequences](tale_mining.html): read it
once that walkthrough has given you a `tales` object to look at, and
return there afterwards to continue with classification. Its sibling deep
dive, on the `tales_msa` class, comes later in the walkthrough.
:::

```{r setup, message = FALSE}
library(tantale)
library(dplyr)
```

# One row per part {#sec-one-row-per-part}

A `tales` object is a long table with **one row per part** of a TALE array:
one row for the N-terminus, one for each central repeat, one for the
C-terminus.

That shape is forced by the biology. A TALE array is built from a variable
number of repeats (as few as a handful, more than thirty in some natural
alleles), so "one row per TALE" would need a ragged number of columns, one
per possible repeat position. In a long table, an array with 12 repeats
and one with 27 are both just more rows, distinguished by `array_id`.

Building one from a real `tell_tales()` run looks like this. The example
below uses the small sequence set shipped with the package, so it runs in a
few seconds without any correction step:

```{r tell_tales_demo, message = FALSE, warning = FALSE}
out <- file.path(tempdir(), "tales_class_article")
invisible(tell_tales(
  subject_file = system.file("extdata", "bai3_sample_tal_genomic_regions.fasta",
                             package = "tantale", mustWork = TRUE),
  output_dir = out))
```

```{r tales_from_demo}
x <- suppressWarnings(tales_from_telltale(out))
x
```

Two columns locate a part within its array, and they answer different
questions:

- `position_in_array` numbers **every** part, termini included, starting at
  1 (the N-terminus, when present).
- `position_in_crd` numbers only the **repeats**, starting at 1 for the
  first one, and is `NA` on the two termini.

On a repeat, `position_in_array` is one ahead of `position_in_crd` when
the array has an N-terminus, and equal to it when it does not. Keeping
only the repeats of an array (`filter(x, domain_type == "repeat")`)
therefore still gives a valid `tales`: `position_in_crd` stays contiguous
from 1 even though `position_in_array` no longer starts there.
Completeness (every part present, `position_in_array` running `1:n`) is a
precondition some functions need, alignment chiefly, and is checked by
`tales_assert_complete()`. The class itself does not require it.

# The column contract

Two columns are required outright, because without them a row cannot be
located at all: `array_id` and `position_in_array`. A third requirement is
looser: at least one of `rvd` or `dom_code` must be present, since a
`tales` with neither has no residues to describe.

Everything else is optional, but validated for shape when it is there:

| column | holds |
|---|---|
| `domain_type` | `"N-terminus"`, `"repeat"`, or `"C-terminus"` |
| `position_in_crd` | position among repeats only, `NA` on termini |
| `aa_seq`, `dna_seq` | the part's amino acid / nucleotide sequence |
| `seqnames` | the source contig, for parts read off a genome |
| `dom_code` | an identifier per distinct domain sequence, added by `tales_assign_domain_codes()` (see below) |
| `source_directory` | which `tell_tales()` run produced this row |

A `tales` built straight from a fasta of RVD sequences (below) has none of
these beyond `rvd` itself, and that is a legitimate, if minimal, object.
Any further column (a strain name, a clade, a host range) is preserved
untouched and never warned about: metadata is expected to accumulate on
this table via ordinary `mutate()` calls.

# Three ways to build one

**From a `tell_tales()` run**, as above, with `tales_from_telltale()`.
This is the richest route: it carries sequences, source coordinates and
both position columns.

**From bare sequences**, with `as_tales()`: a fasta of RVD or
repeat-code strings, sequence names becoming `array_id`:

```{r as_tales_demo}
rvd_fasta <- system.file("extdata", "TalA_RVDSeqs_AnnoTALE.fasta",
                         package = "tantale")
cat(readLines(rvd_fasta, n = 2), sep = "\n")

as_tales(rvd_fasta, sep = "-")
```

This result has only `array_id`, `position_in_array` and `rvd`, because
that is all a bare sequence file can supply. It is still a complete,
valid `tales`.

**From a comparison**, with `tales_assign_domain_codes()` (see below): the
same object, given back with a `dom_code` column added.

All three produce the same class, checked by the same `validate_tales()`,
usable with the same functions afterwards.

# It is a tibble

A `tales` is a tibble with extra structure layered on top. Ordinary dplyr
verbs work on it directly:

```{r dplyr_demo}
x %>%
  filter(domain_type == "repeat") %>%
  count(array_id)
```

Subsetting that would break the class, such as dropping `array_id`,
quietly degrades the result to a plain tibble instead of raising an
error, so a `tales` never claims more than it is. @sec-subsetting looks at
that degradation more closely, once `dom_code` is in the picture.

`print()` and `format()` follow the same idea Biostrings uses for sequence
sets: a header with the essentials (arrays, parts, which residue columns
are present), then the first and last arrays rendered as sequences, eliding
the middle. `summary()` adds the slower numbers (how many distinct RVDs,
how many repeats per array, whether every array has both termini, and,
once `dom_code` is present, how many distinct domains) and folds in
`tales_anomalies()`:

```{r summary_demo}
summary(x)
```

# Anomalies are reported and kept

Real TALE predictions are messy: a terminus can be duplicated, misplaced,
or simply not found. `tales()` loads such arrays anyway, with a warning.
A class that insisted on clean input would force cleaning outside the
package and throw away exactly the signal a user would want to inspect.

"Anomalous" is deliberately a narrower category than "wrong". An array
missing a terminus, for instance, is not flagged here at all: it may
genuinely sit at the edge of a contig. What *is* flagged is a part
arrangement that cannot be biologically real (two N-termini in one array,
say, or one that is not at the start), a part missing the sequence data
it should carry, one amino acid sequence paired with two different RVDs,
and position columns that disagree with each other:

```{r anomaly_demo}
odd <- tibble::tibble(
  array_id = c("a1", "a1", "a1", "a2", "a2", "a2"),
  position_in_array = c(1L, 2L, 3L, 1L, 2L, 3L),
  domain_type = c("N-terminus", "N-terminus", "repeat",
                  "N-terminus", "repeat", "C-terminus"),
  rvd = c("NTERM", "NTERM", "HD", "NTERM", "HD", "CTERM")
)
x_odd <- suppressWarnings(tales(odd))
tales_anomalies(x_odd)
```

`tales(sanitize = TRUE)` drops the flagged arrays instead of merely warning
about them:

```{r sanitize_demo}
tales(odd, sanitize = TRUE)
```

Structural problems, such as a duplicated key or an `NA` where one is not
allowed, remain hard errors either way, because the table cannot be
interpreted at all.

# Projections: turning parts back into sequences

Some consumers want a `tales` as a table; others want each array back as a
single string. Three functions bridge the gap:

- `tales_rvd_strings()`: one RVD string per array, the form
  target-prediction tools consume. Drops the termini by default
  (`rvd_only = TRUE`), since only the repeats bind DNA.
- `tales_coded_strings()`: the `dom_code` equivalent, for ARLEM and
  MAFFT's text mode. Keeps the termini by default, since alignment uses
  them as anchors, and separates codes with a space, which is what those
  tools split on.
- `tales_domain_codes()`: the lookup table itself, with one row per
  distinct `dom_code` and the amino acid sequence and RVD it stands for.

```{r projections_demo}
xa <- tales_assign_domain_codes(x)
tales_coded_strings(xa)["ROI_00002"]
tales_domain_codes(xa) %>% head(4)
```

# What a `dom_code` is {#sec-dom-code}

This is the column the rest of the comparison machinery is built on, and it
is worth being precise about.

## It names a domain, repeats and termini alike

A TALE part is an N-terminus, a repeat, or a C-terminus, and **all three get
`dom_code`s on the same footing**. "Domain" is the word used here because it
covers a repeat and a terminus alike; a column that also numbers C-termini
cannot be called a repeat code. Two parts with identical amino acid
sequences get the same code, whatever kind of part they are.

The distinction matters in practice. In the small example used throughout
this article:

```{r dom_code_counts}
n_distinct(xa$dom_code)
table(xa$domain_type[match(unique(xa$dom_code), xa$dom_code)])
```

`r n_distinct(xa$dom_code)` distinct codes, and only
`r sum(xa$domain_type[!duplicated(xa$dom_code)] == "repeat")` of them are
repeats; the rest are distinct N- and C-termini. Reading the total as a
count of distinct repeats would overstate repeat diversity.

## It is what makes a TALE alignable

Aligning TALE arrays residue by residue is meaningless: the repeats are
near-identical (~34 amino acids, differing mainly at two positions), so
almost anything matches almost anything else. Giving each distinct domain a
symbol turns an array into a *sequence of domains*, which can be aligned
the way a protein sequence is, with insertions and deletions of whole
repeats instead of individual residues. This is why `tales_align()` works
on `dom_code` (or `rvd`), never on `aa_seq` directly.

## It is finer than an RVD

The RVD is residues 12-13 of a repeat, and it is what determines which DNA
base the repeat binds. Two repeats can share an RVD -- the same binding
specificity -- while differing elsewhere in the protein, and they get
**different** `dom_code`s. So `rvd` is the functional layer and `dom_code`
the identity layer, and the class carries both because they answer
different questions.

In this example, the RVD `HD` (which binds cytosine) is shared by several
distinct repeat sequences:

```{r hd_example}
reps <- xa %>% filter(domain_type == "repeat")
reps %>% filter(rvd == "HD") %>% distinct(dom_code, aa_seq) %>% head(4)
```

A terminus, meanwhile, has no RVD at all: its `rvd` column holds a
placeholder (`"NTERM"`, `"CTERM"`, or `"XXXXX"` for a terminus whose CDS
was detected but not identified; see `tales_anchor_codes()`), while its
`dom_code` identifies a real sequence.

## How they are computed

`tales_assign_domain_codes()` groups the parts by `aa_seq` and numbers
the groups (`dplyr::cur_group_id()`).

## The numbers are only meaningful within one call

Because the numbering depends on which distinct sequences are present,
code `42` from one comparison is not code `42` from another: add an
array, remove one, or run it on a different day's data, and the same
protein can land on a different number. The package enforces this.

Every object `tales_assign_domain_codes()` produces carries a namespace
stamp, readable with `tales_namespace()`, derived from the exact set of
distinct sequences that produced it:

```{r namespace_demo}
tales_namespace(xa)
```

Functions that combine codes from two different calls refuse to run. The
rest of this section demonstrates that, one step at a time.

Domain distances over the whole example set are the baseline to mix in
something incompatible with:

```{r namespace_full_distances, message = FALSE}
dd_all <- tales_domain_distances(xa)
```

Splitting the example array set in two and re-assigning codes to one part
gives it a *different* namespace, even though every sequence it contains
was already present in the full set:

```{r namespace_resplit}
sub <- xa[xa$array_id != "ROI_00001", ][setdiff(names(xa), "dom_code")]
xa_sub <- tales_assign_domain_codes(sub)

tales_namespace(xa_sub) == tales_namespace(xa)
```

::: {.callout-warning}
## Mixing namespaces is an error
Using `xa_sub`'s codes against `dd_all`, distances keyed by the other
namespace, would compute a plausible-looking wrong answer. The function
refuses instead:

```{r namespace_mismatch, error = TRUE}
tales_tale_distances(xa_sub, dd_all)
```
:::

# Where `dom_code` leads: comparing TALEs

Minting codes is the first of three steps that together compare a set of
TALEs: `tales_assign_domain_codes()` first, then `tales_domain_distances()`
(how different the distinct domains are from each other, pairwise), then
`tales_tale_distances()` (how different the *arrays* are, aligning their
domain-code sequences using the domain distances as the substitution cost).
`tales_compare_distal()` runs all three and returns the corresponding
`tales`, `domain_distances` and `tale_distances` objects together:

```{r tales_compare_demo, message = FALSE, warning = FALSE}
cmp <- tales_compare_distal(x)
names(cmp)
```

Each step is independently useful (the domain-level distances alone
answer questions about domain diversity, without running the array-level
alignment) and each is documented in its own right; see
`tales_compare_distal()` for the full chain.

# Subsetting keeps the class only while the contract holds {#sec-subsetting}

A `tales` is a tibble, so every subsetting route a tibble supports works on
it too: `[`, `filter()`, `select()`, `arrange()`, `mutate()`. What differs
from a plain tibble is what happens to the *class* along the way.

**Row subsetting never breaks anything.** Every invariant a `tales` checks
still holds on any subset of its rows. This is why completeness
(@sec-one-row-per-part, above) is checked only by the functions that
need it. Filtering to one array, or to its repeats only, is still a
valid `tales`:

```{r subset_rows}
one_array <- xa[xa$array_id == "ROI_00002", ]
is_tales(one_array)
nrow(one_array)
```

**Column subsetting is where it gets interesting.** Dropping a column the
contract needs (`array_id`, or both residue columns at once) leaves
something that can no longer be described as a `tales`. The class then
quietly steps out of the way, without an error:

```{r subset_drop_key}
no_id <- xa %>% select(-array_id)
is_tales(no_id)
class(no_id)
```

The same happens whichever route removes the load-bearing column:
`select()`, `mutate()`, or plain `[, j]`. This is deliberate: a `tales`
object that kept claiming to be one after losing its key would let every
downstream function trust a promise the object could no longer keep.
Dropping an optional column (`seqnames`, `aa_seq`, ...) has no such
consequence, and the class survives untouched.

`mutate()` earns one extra check the others do not need: it can overwrite
`array_id` or `position_in_array` *in place*, which plain subsetting cannot
do, so the result could keep every contract column while no longer being a
valid key. That case is caught separately, immediately after the mutate:

```{r subset_mutate_break_key, error = TRUE}
xa %>% mutate(position_in_array = 1L)
```

The row-level biology is **not** re-checked on every verb:
`tales_anomalies()` does not run again on each `filter()` call. Anomalies
are a property to inspect deliberately (see above).

**Attributes travel with a valid subset**, which matters most for
`dom_code_namespace`. Cutting `xa` down to one array with `[` keeps its
namespace, because the namespace describes *the run the codes came from*:

```{r subset_namespace}
tales_namespace(one_array) == tales_namespace(xa)
```

That holds because `one_array` was cut with `[` and kept its codes;
`tales_assign_domain_codes()` on a subset mints a *different* namespace, as
the mismatch example above showed.

A `tales_msa` carries one further column, `alignment_position`. Losing
*that* while keeping everything else degrades it by exactly one step, to
a plain `tales`.
[The `tales_msa` class article](tales_msa_class.html) covers that graded
degradation in full, together with the reverse direction: how a `tales` is
promoted into an alignment in the first place.

# Next

This article stands on its own; where to go next depends on what you were
after. To continue the walkthrough this extends,
[return to classifying TALE sequences](tale_classification.html). To go
deeper on alignment, [the `tales_msa` class](tales_msa_class.html) picks
up directly from @sec-dom-code and @sec-subsetting, above.
