---
title: "Classifying TALE sequences from genomes"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

The [previous article](tale_mining.html) covered finding TALE loci in one
genome at a time. A real study usually has several genomes, and the
interesting question is not "what TALEs does this strain carry" but
"which TALEs, across strains, are versions of the same thing" -- alleles
of one locus, related by descent, rather than unrelated proteins that
happen to both be TALEs. This article covers that comparison:
`tales_compare()` to quantify how alike every pair of arrays and repeats
is, and `tales_group()` to turn that into discrete groups.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

## Discovering TALEs across several genomes

This is a fresh discovery run, independent of the previous article, over
three related genomes: MAI1, BAI3, and the best-corrected version of
BAI3-1-1 from
[the correction section of the mining article](tale_mining.html#sec-best-correction).

::: {.callout-note}
## Why only three genomes, not the fourth
tantale ships a fourth sample genome, PXO86, a distantly related Asian
outgroup strain. It is left out here deliberately: its TALE repertoire
barely overlaps the African strains' at all, so including it turns a
clean set of one-locus-per-strain groups into a mix of real cross-strain
groups and PXO86-only paralog clusters that illustrate a different
question than the one this article is asking. `tales_group()` on three
related genomes, below, is a considerably better demonstration of what it
does than the same call over four genomes spanning two continents.
:::

Discovery runs independently per genome, so it is written as one function
applied to each genome name in turn:

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_classification"))
genome_files <- c(
  MAI1       = system.file("extdata", "MAI1.fa",       package = "tantale", mustWork = TRUE),
  BAI3       = system.file("extdata", "BAI3.fa",       package = "tantale", mustWork = TRUE),
  `BAI3-1-1` = system.file("extdata", "BAI3-1-1.fa",   package = "tantale", mustWork = TRUE)
)
```

```{r discover_one}
discover_strain <- function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir,
                       cterm_min_score = 300,
                       correct_array = TRUE, max_comparisons = 50))
  tales_from_telltale(strain_dir) |>
    mutate(array_id = paste0(strain, "_", array_id), strain = strain)
}
```

`correct_array = TRUE` with `max_comparisons = 50` is the route
[the mining article](tale_mining.html#sec-best-correction) found clean
for BAI3-1-1; run on MAI1 and BAI3 too, it costs a little extra time for no change,
since neither needed correcting.

```{r discover_all, message = FALSE}
all_tales <- lapply(names(genome_files), discover_strain) |>
  suppressWarnings() |>
  bind_rows()
```

`array_id` is prefixed by strain before the three objects are combined,
since `tell_tales()` numbers regions independently within each genome --
without the prefix, `MAI1`'s `ROI_00001` and `BAI3`'s `ROI_00001` would
collide.

```{r sanitize_all}
all_tales <- tales(all_tales, sanitize = TRUE)
n_distinct(all_tales$array_id)
```

## Quantifying relatedness: `tales_compare()`

`tales_compare()` is tantale's R implementation of the DisTAL comparison
used by the original QueTAL suite: it assigns every distinct domain
sequence a `dom_code` (see [the `tales` class article](tales_class.html)
for what that means), computes pairwise distances between distinct
domains, and from those, distances between whole arrays.

```{r tales_compare, message = FALSE}
cmp <- tales_compare(all_tales, aln_method = "DECIPHER", ncores = 4)
names(cmp)
```

```{r tale_distances_show}
cmp$tale_distances
```

### Does the choice of backend matter? {#sec-backends}

Comparing domain sequences -- the step just run -- is the expensive part,
and three backends can do it: `"DECIPHER"` (the default, and by far the
fastest, which is why it was used above), `"Biostrings"`, and `"mmseq2"`.
They implement the same pairwise protein alignment, so which one to pick
is a question of what is installed and how large the dataset is, not of
which answer is right. Checking that claim, rather than trusting it, on a
small independent fixture so all three run quickly:

```{r backend_fixture}
small <- tales_from_telltale(system.file("extdata", "tellTaleExampleOutput",
                                         package = "tantale"))
```

```{r backend_runs, message = FALSE}
cmp_decipher   <- tales_compare(small, aln_method = "DECIPHER")
cmp_biostrings <- tales_compare(small, aln_method = "Biostrings")
cmp_mmseq2     <- tales_compare(small, aln_method = "mmseq2")
```

`as.matrix()` on a `tale_distances` object always returns ids sorted the
same way, so the three backends' outputs line up without any further
work:

```{r backend_correlate}
backend_dists <- tibble::tibble(
  DECIPHER   = as.vector(as.matrix(cmp_decipher$tale_distances)),
  Biostrings = as.vector(as.matrix(cmp_biostrings$tale_distances)),
  mmseq2     = as.vector(as.matrix(cmp_mmseq2$tale_distances))
)
cor(backend_dists)
```

All three agree closely on this fixture, which is the point: the fast
default is not a shortcut that trades accuracy for speed here, at least
not on data this size.

## Allocating arrays to groups: `tales_group()`

`tales_group()` clusters arrays from `tale_distances` and returns the
`tales` object with the result attached as a `group` column -- taking the
object the distances were computed from is what lets it check that the
two actually correspond (see `?tales_group`).

Two clustering methods are available. `method = "k-medoids"` (the
default) fits `cluster::pam()` across a range of candidate group counts
and picks one by its silhouette value; `k = "auto"` accepts that
automatic pick:

```{r tales_group}
grouped <- tales_group(cmp$tales, cmp$tale_distances,
                       method = "k-medoids", k = "auto", k_range = 2:20)
```

```{r group_sizes}
group_sizes <- grouped |> distinct(array_id, group) |> count(group) |> arrange(desc(n))
group_sizes
```

With PXO86 out of the picture, this is a clean result: `r sum(group_sizes$n == 3)`
of `r nrow(group_sizes)` groups have exactly three members, one from each
strain -- the one-locus-per-strain pattern this comparison is meant to
recover.

`method = "hclust"` cuts a dendrogram at a height chosen to yield exactly
`k` groups instead, and with `plot_tree = TRUE` draws it -- useful for
judging whether the chosen `k` looks reasonable rather than trusting the
silhouette value blindly:

```{r}
#| label: fig-tale-dendrogram
#| fig-cap: "Hierarchical clustering of the same three-genome comparison, cut at the k tales_group() picked automatically above."
#| fig-width: 7
#| fig-height: 5
invisible(tales_group(cmp$tales, cmp$tale_distances,
                      method = "hclust", k = n_distinct(grouped$group),
                      plot_tree = TRUE))
```

## An overview across strains: `talomes_heatmap()`

With arrays assigned to groups, one natural summary is which RVD sequence
variant each strain carries in each group -- a *talome*, by analogy to a
genome, being the whole complement of TALEs a strain carries.

```{r talome_annotation}
tale_annotation <- tibble::tibble(
  array_id = names(tales_rvd_strings(grouped)),
  rvdseq = as.character(tales_rvd_strings(grouped))
) |>
  left_join(distinct(grouped, array_id, group, strain), by = "array_id")
```

```{r}
#| label: fig-talomes-heatmap
#| fig-cap: "RVD sequence variant carried by each strain, in each classification group."
#| fig-width: 8
#| fig-height: 6
talomes_heatmap(tale_annotation, group_col = "group", strain_col = "strain",
                rvd_col = "rvdseq")
```

Each cell in @fig-talomes-heatmap is one strain's RVD sequence variant in
one group; a strain missing from a group carries no member of it, and a
cell with more than one colour means that strain carries more than one
distinct variant in that group.

## Next

Grouping tells you *which* arrays are related. Seeing exactly *how* --
which repeats match, which are inserted or deleted relative to one
another -- needs an alignment, which is the subject of
[the next article](tale_msa.html).
