---
title: "Predicting TALE targets"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

The previous articles found TALE arrays, grouped them, and aligned
related ones. This last article asks the question a TALE's RVD sequence
exists to answer: what DNA sequence does it bind? Each RVD contacts one
base of the target through a well-characterised, largely one-to-one code
(`NI` prefers A, `HD` prefers C, `NG` prefers T, `NN` prefers G or A, and
so on), so a whole RVD sequence predicts an *EBE* -- an Effector Binding
Element -- as a short DNA motif with one position per RVD.

```{r package, message = FALSE}
library(tantale)
library(dplyr)
```

tantale wraps two independent predictors that use this code differently
enough to be worth contrasting:
[TALVEZ](https://doi.org/10.1371/journal.pone.0068464), a
position-weight-matrix scan, and
[PrediTALE](https://doi.org/10.1371/journal.pcbi.1007206), which also
models where a real EBE tends to sit relative to a transcription start
site. Both are wrapped as `talvez()` and `preditale()`, taking the same
inputs and returning the same column layout, so their predictions can be
compared directly rather than translated by hand.

## Getting RVD sequences to predict with

RVD sequences come straight from a `tales` object via
`tales_rvd_strings()` -- the same projection used for alignment input,
just with the termini dropped by default, since prediction concerns the
repeat domain only. This reuses the same three genomes as
[the classification](tale_classification.html) and
[alignment](tale_msa.html) articles.

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "tale_target_prediction"))
genome_files <- c(
  MAI1       = system.file("extdata", "MAI1.fa",     package = "tantale", mustWork = TRUE),
  BAI3       = system.file("extdata", "BAI3.fa",     package = "tantale", mustWork = TRUE),
  `BAI3-1-1` = system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE)
)
```

```{r discover_all, message = FALSE}
all_tales <- lapply(names(genome_files), function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir,
                       cterm_min_score = 300,
                       correct_array = TRUE, max_comparisons = 50))
  tales_from_telltale(strain_dir) |>
    mutate(array_id = paste0(strain, "_", array_id))
}) |>
  suppressWarnings() |>
  bind_rows() |>
  tales(sanitize = TRUE)
```

```{r rvds}
rvds <- tales_rvd_strings(all_tales)
length(rvds)
```

The target sequences are promoter regions of three clade III *SWEET*
genes -- a well-studied class of rice susceptibility genes that TALEs
from *Xanthomonas oryzae* are known to target -- from several rice
varieties:

```{r subj_file}
subj_file <- system.file("extdata", "cladeIII_sweet_promoters.fasta",
                         package = "tantale", mustWork = TRUE)
Biostrings::readBStringSet(subj_file) |> names()
```

## TALVEZ

```{r talvez, message = FALSE}
talvez_preds <- talvez(rvd_seqs = rvds, subj_file = subj_file, opt_param = "-t 0 -l 19")
nrow(talvez_preds)
```

```{r talvez_show}
talvez_preds |> arrange(desc(score)) |> head(5)
```

## PrediTALE

```{r preditale, message = FALSE, results = "hide"}
preditale_preds <- preditale(rvd_seqs = rvds, subj_file = subj_file)
```

```{r preditale_show}
nrow(preditale_preds)
preditale_preds |> arrange(desc(score)) |> head(5)
```

::: {.callout-tip}
## `pval`, not just a score
PrediTALE additionally reports a `pval` per prediction, since it models
the null distribution of scores directly rather than only ranking
candidates against each other -- worth using when the question is "is
this a credible site at all" rather than only "which of these is the
best".
:::

## Visualising predictions against the target sequence

`plot_target_preds()` draws predicted RVD-to-base correspondences
directly against the target DNA, coloured by how well each RVD's known
base preference matches the base actually predicted underneath it -- so
a "good" prediction is visually distinguishable from one held up by only
one or two well-matching RVDs.

::: {.callout-important}
## Keep the plotted window narrow
Every RVD gets its own box, one per base of the target. A window wide
enough to span several unrelated predictions squeezes every box down to
a sliver too narrow for its own label. Pick a window around *one* region
of interest -- here, where two orthologous TALEs from different strains
converge on the same site -- rather than the whole promoter at once.
:::

The same locus turns up more than once across strains -- unsurprising,
since [the classification article](tale_classification.html) already
found this comparison to be dominated by one-locus-per-strain groups.
Three arrays, one from each strain, all predict the *same* site:

```{r best_predictions}
converging <- preditale_preds |>
  filter(taleId %in% c("BAI3_ROI_00008", "BAI3-1-1_ROI_00007", "MAI1_ROI_00008"),
        subjSeqId == "SWEET14p_Nipponbare_Sense")
converging |> select(subjSeqId, start, end, score, taleId)
```

```{r}
#| label: fig-target-preds
#| fig-cap: "Three orthologous TALEs, one from each strain, predicted to bind the exact same site in the SWEET14 promoter."
#| fig-width: 12
#| fig-height: 3.6
plot_target_preds(preds = converging, subj_file = subj_file,
                  filter_range = "SWEET14p_Nipponbare_Sense:460-505")
```

Each RVD is printed in a box over the base it is predicted to contact;
predictions on the sense strand are drawn above the sequence and those on
the antisense strand below it, since a TALE can bind either strand of
its target. All three predictions in @fig-target-preds land on the exact
same site, from the same locus in three different strains, with
identical scores -- exactly the kind of convergence
[the alignment article](tale_msa.html) would call conserved rather than
coincidental.

## Closing the loop

This set of articles followed one path through a TALE study:
`tell_tales()` to find arrays (correcting them, and checking for what
correction could not fix, along the way), `tales_compare()` and
`tales_group()` to relate them, `tales_align()` to see that relatedness
directly, and `talvez()`/`preditale()` to ask what they do. None of these
steps requires the ones before it in code -- each function documented
here takes ordinary `tales` objects or plain sequences -- but together
they cover most of what a study of TALE diversity in a set of genomes
actually needs.
