---
title: "The tales_msa class"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

[The `tales` class article](tales_class.html) covered `tales`: one row
per part, with `dom_code` as the identity layer that makes an array
alignable. This one is a deep dive into what happens once several arrays
*are* aligned -- the `tales_msa` class -- and about the traffic between
the two classes: how a `tales` is promoted into one, how it demotes back,
and what each class draws when plotted.

::: {.callout-note}
## Where this fits
This picks up directly from
[the alignment built in the walkthrough](tale_msa.html#aligning-the-group)
-- same three arrays, same real internal gap -- and goes further into the
mechanics than that article needed to.
:::

```{r setup, message = FALSE}
library(tantale)
library(dplyr)
```

## Building the alignment this article examines

Same three genomes and the same group as
[the walkthrough](tale_msa.html), reused here from the cached result of
[the classification article](tale_classification.html#sec-backends)
rather than recomputed; see
[the getting-started article](getting_started.html) for how these four
articles are linked.

```{r cached_pipeline, include = FALSE}
read_cache <- function(path) {
  if (!fs::file_exists(path)) {
    cli::cli_abort(c(
      "Cached pipeline result not found at {.path {path}}.",
      "i" = "Render {.file tale_classification.qmd} first (or run a full {.code pkgdown::build_site()}), which computes and caches it."
    ))
  }
  readRDS(path)
}
cmp <- read_cache("_cache/compare.rds")
group6 <- read_cache("_cache/group.rds") |>
  filter(group == 6)
```

`tales_align()` takes a `tales` and returns a `tales_msa`: the same rows,
plus one new column, `alignment_position`. Nothing else changes -- every
column the input carried, `rvd` and `dom_code` both included, survives
the trip, because the alignment is computed on *one* residue layer
(`residue_col`) but the object it returns is still the whole table.

```{r build_msa, message = FALSE}
msa <- tales_align(group6, residue_col = "dom_code")
msa
```

## What is new: `alignment_position`

A `tales_msa` is a `tales` with one extra column and one extra stored
number:

- **`alignment_position`** -- where this part sits in the alignment,
  which is not the same coordinate as `position_in_array`. The two agree
  up to the first gap and diverge after it.
- **`tales_width()`** -- the alignment's total column count, stored
  rather than derived, because `max(alignment_position)` over a *subset*
  of arrays can under-report a width whose last columns happen to be all
  gaps in that subset.

```{r width_demo}
tales_width(msa)
```

The array with the internal deletion shows the divergence directly:
`position_in_array` counts its 14 real parts contiguously, while
`alignment_position` jumps by five across the gap.

```{r width_demo_detail}
msa |>
  filter(array_id == "BAI3_ROI_00007") |>
  select(array_id, position_in_array, alignment_position, domain_type)
```

::: {.callout-tip}
## Gaps are implicit, not stored
There is no row, anywhere, whose `alignment_position` is a gap -- a gap
is simply a column with no row for that array. This is why the class
validates cheaply and why every `tales` invariant (the key, the residue
columns, the `dom_code` bijection) keeps holding unchanged on a
`tales_msa`: nothing new was added to *rows*, only a new column that says
where each existing row lands.
:::

What the alignment adds on top is its own, narrower, set of rules --
checked by `validate_tales_msa()` in addition to everything
`validate_tales()` already checks: `alignment_position` is a positive
integer, present on every row; unique within an array; ordered like
`position_in_array` (an alignment may insert gaps, it may never reorder
parts); and never beyond the declared width.

## Views: matrices are computed, not stored

The long table is the only thing actually stored. A rectangular alignment
-- the picture most people have in mind, one row per array, one column
per position -- is a *view*, built on request by `as.matrix()`, and it
has to be asked which layer to render:

```{r as_matrix_dom_code}
as.matrix(msa, value = "dom_code")
```

```{r as_matrix_rvd}
as.matrix(msa, value = "rvd")
```

Both matrices have the same shape and the same gaps -- they are two
readings of one geometry, not two independent alignments. That
equivalence is exactly what lets `plot.tales_msa()` (below) show a
domain's identity as a colour and its RVD as a label on the same cell --
a terminus included, since termini carry an RVD-string entry (their
`NTERM`/`CTERM` marker) just as repeats do.

## Coercion between `tales` and `tales_msa`

**Promotion** is `tales_align()`: give it a complete `tales`, get back a
`tales_msa`. There is no other route, because the alignment column has to
come from an actual alignment run -- nothing else could populate it
honestly.

**Demotion** is `as_tales()`, and it is a real operation, not a
formality:

```{r demote}
demoted <- as_tales(msa)
class(demoted)
```

```{r demote_check}
is_tales_msa(demoted)
"alignment_position" %in% names(demoted)
```

`alignment_position` survives as an ordinary column -- demoting does not
erase where things were aligned, it only stops the object *claiming* to
be a coherent single alignment. That claim is exactly what
`tales_width()` encodes, and it is gone:

```{r demote_width}
tales_width(demoted)
```

::: {.callout-important}
## Why this distinction is not pedantic
A talome-wide summary spanning several *independently* aligned groups
cannot honestly be one `tales_msa` -- each group has its own width and
its own coordinate system, so `alignment_position = 5` would mean
unrelated things in two different groups. Demoting to a plain `tales`
and carrying `alignment_position` as an ordinary column (per group) is
the honest shape for that case; the class is not stretched to cover
something it cannot actually guarantee.
:::

**Subsetting degrades the same way, automatically, one step at a time.**
Row subsetting never touches the alignment contract, since every check
above is closed under keeping a subset of rows. Column subsetting can:
dropping `alignment_position` itself leaves something that is still a
valid `tales` (everything else about a part is unaffected) but can no
longer be a `tales_msa`, and the class notices without being told:

```{r subset_degrade}
no_alignment <- msa |> select(-alignment_position)
class(no_alignment)
```

This is the same graded-degradation mechanism from
[the `tales` class article's](tales_class.html#sec-subsetting) `[.tales`
method, just visible one level up: a `tales_msa` that loses what makes it
an alignment steps down to a `tales` rather than lying about still being
one, and a `tales` that loses its key steps down to a bare tibble for the
same reason. Losing an *optional* column -- `seqnames`, `aa_seq` -- costs
nothing at either level.

## Plotting: two methods, one generic

`plot()` dispatches on class, so the two objects draw different things
from the same call.

### `plot.tales()` before and after alignment

A `tales` without an alignment can only be laid out on
`position_in_array`: every array starts its own count at 1, so an
aberrant repeat scatters to wherever it happens to sit in each array.
`position = "alignment"` needs the column that only an aligned (or
demoted-from-aligned) object carries, and lines features up instead:

```{r}
#| label: fig-plot-before
#| fig-cap: "The same three arrays laid out by position_in_array -- no gaps, because nothing has been aligned."
#| fig-width: 8
#| fig-height: 2.6
plot(group6, position = "array")
```

```{r}
#| label: fig-plot-after
#| fig-cap: "The same three arrays laid out by alignment_position -- the shared deletion now lines up as a column."
#| fig-width: 8
#| fig-height: 2.6
plot(demoted, position = "alignment")
```

### `plot.tales_msa()`: three things decided independently

The alignment plot draws a heatmap -- one row per array, one column per
position -- and three questions about it are answered separately:

- **what each cell *says*** -- `label`, defaulting to `rvd` whenever that
  is not also the fill layer;
- **what colour that text is** -- always whether the cell matches the
  consensus of its column (`tales_consensus()`), cyan for yes, pink for
  no, regardless of what `fill_type` is showing;
- **what colour the block behind it is** -- `fill_type`, the only one of
  the three that can be unavailable.

```{r fill_type_table, echo = FALSE}
knitr::kable(tibble::tribble(
  ~`fill_type`, ~shows, ~needs,
  '`"domain_clust"`', "which cluster the domain falls in, cut at `h_cut`", "`domain_distances`",
  '`"domain_sim"`', "protein-sequence similarity to the reference, 0-100", "`domain_distances`",
  '`"rvd_sim"`', "how alike the RVD's *DNA-binding preference* is to the reference's, -1 to 1", "a `label` layer"
))
```

The `domain_distances` argument takes exactly that -- the `domain_distances`
element of `tales_compare_distal()`'s result. Passing `tale_distances`
(the `tale_distances` element) in addition adds a dendrogram panel
ordering the rows by relatedness:

```{r}
#| label: fig-msa-plot-default
#| fig-cap: "Default plot.tales_msa() view: domain-cluster fill, RVD labels, and a tree from tale_distances."
#| fig-width: 9
#| fig-height: 2.3
plot(msa, tale_distances = cmp$tale_distances, domain_distances = cmp$domain_distances)
```

`"rvd_sim"` asks a different question at a different layer: not "which
cluster" or "how similar a protein", but how alike each RVD's DNA-binding
specificity is to the reference's. It needs no `domain_distances` at all,
only the RVDs themselves:

```{r}
#| label: fig-msa-plot-rvdsim
#| fig-cap: "The same alignment, coloured by RVD specificity relative to the reference rather than by repeat identity."
#| fig-width: 9
#| fig-height: 2.3
plot(msa, fill_type = "rvd_sim")
```

The domain- and RVD-level views genuinely differ: `HD` and `ND` are
distinct domains with identical specificity, while domains differing
only at positions 12-13 are near-identical proteins with opposite target
bases -- so `"domain_sim"` and `"rvd_sim"` are not two colour schemes for
the same fact, they are answers to two different questions.

## Next

Return to [the walkthrough](tale_msa.html) to continue with target
prediction, or to [the `tales` class article](tales_class.html) for the
identity layer this alignment is built on.
