---
title: "The tales class"
format:
  html:
    toc: true
    number-sections: false
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every function in tantale that describes, compares or plots a set of TALEs
takes or returns a `tales` object. This article is about what that object
is, how to build one, and what its most important column -- `dom_code` --
actually means.

```{r setup, message = FALSE}
library(tantale)
library(dplyr)
```

# One row per part, not per TALE

A `tales` object is a long table with **one row per part** of a TALE array:
one row for the N-terminus, one for each central repeat, one for the
C-terminus. It is not one row per TALE.

That shape is forced by the biology. A TALE array is built from a variable
number of repeats -- as few as a handful, more than thirty in some natural
alleles -- so "one row per TALE" would need a ragged number of columns, one
per possible repeat position. A long table sidesteps that entirely: an array
with 12 repeats and one with 27 are both just more rows, distinguished by
`array_id`.

Building one from a real `tell_tales()` run looks like this. The example
below uses the small sequence set shipped with the package, so it runs in a
few seconds without any correction step:

```{r tell_tales_demo, message = FALSE, warning = FALSE}
out <- file.path(tempdir(), "tales_class_article")
invisible(tell_tales(
  subject_file = system.file("extdata", "bai3_sample_tal_genomic_regions.fasta",
                             package = "tantale", mustWork = TRUE),
  output_dir = out))

x <- suppressWarnings(tales_from_telltale(out))
x
```

Two columns locate a part within its array, and they answer different
questions:

- `position_in_array` numbers **every** part, termini included, starting at
  1 (the N-terminus, when present).
- `position_in_crd` numbers only the **repeats**, starting at 1 for the
  first one, and is `NA` on the two termini.

They agree wherever there is no N-terminus to offset them, and differ by
exactly the number of non-repeat parts that come before a given position
otherwise. This is also why keeping only the repeats of an array
(`filter(x, domain_type == "repeat")`) is a perfectly valid `tales` and not
a broken one: `position_in_crd` stays contiguous from 1 even though
`position_in_array` no longer starts there. Completeness -- every part
present, `position_in_array` running `1:n` -- is a precondition some
functions need (alignment, chiefly, checked by `tales_assert_complete()`), not an invariant of the class
itself.

# The column contract

Two columns are required outright, because without them a row cannot be
located at all: `array_id` and `position_in_array`. A third requirement is
looser -- at least one of `rvd` or `dom_code` must be present, since a
`tales` with neither has no residues to describe.

Everything else is optional, but validated for shape when it is there:

| column | holds |
|---|---|
| `domain_type` | `"N-terminus"`, `"repeat"`, or `"C-terminus"` |
| `position_in_crd` | position among repeats only, `NA` on termini |
| `aa_seq`, `dna_seq` | the part's amino acid / nucleotide sequence |
| `seqnames` | the source contig, for parts read off a genome |
| `dom_code` | see below -- minted by comparison, not by discovery |
| `source_directory` | which `tell_tales()` run produced this row |

A `tales` built straight from a fasta of RVD sequences (below) has none of
these beyond `rvd` itself, and that is a legitimate, if minimal, object.
Any further column -- a strain name, a clade, a host range -- is preserved
untouched and never warned about: metadata is expected to accumulate on
this table via ordinary `mutate()` calls.

# Three ways to build one

**From a `tell_tales()` run**, as above, with `tales_from_telltale()`.
This is the richest route: it carries sequences, source coordinates and
both position columns.

**From bare sequences**, with `as_tales()` -- a fasta of RVD or
repeat-code strings, sequence names becoming `array_id`:

```{r as_tales_demo}
rvd_fasta <- system.file("extdata", "TalA_RVDSeqs_AnnoTALE.fasta",
                         package = "tantale")
cat(readLines(rvd_fasta, n = 2), sep = "\n")

as_tales(rvd_fasta, sep = "-")
```

This result is deliberately column-poor -- `array_id`, `position_in_array`
and `rvd`, nothing else -- because that is all a bare sequence file can
supply. It is still a complete, valid `tales`.

**From a comparison**, with `tales_assign_domain_codes()` (see below): the
same object, given back with a `dom_code` column added.

All three produce the same class, checked by the same `validate_tales()`,
usable with the same functions afterwards.

# It is a tibble

A `tales` is a tibble with extra structure layered on top, not a different
kind of thing. Ordinary dplyr verbs work on it directly:

```{r dplyr_demo}
x %>%
  filter(domain_type == "repeat") %>%
  count(array_id)
```

Subsetting that would break the class -- dropping `array_id`, say -- quietly
degrades the result to a plain tibble rather than erroring, so a `tales`
never lies about what it is. The mechanics of that degradation are worth a
closer look once `dom_code` is in the picture -- see below.

`print()` and `format()` follow the same idea Biostrings uses for sequence
sets: a header with the essentials (arrays, parts, which residue columns
are present), then the first and last arrays rendered as sequences, eliding
the middle. `summary()` adds the slower numbers -- how redundant the domains
are, how many repeats per array, whether every array has both termini --
and folds in `tales_anomalies()`:

```{r summary_demo}
summary(x)
```

# Anomalies are reported, not refused

Real TALE predictions are messy: a terminus can be duplicated, misplaced,
or simply not found. `tales()` loads such arrays anyway rather than
refusing them -- a class that insists on clean input forces cleaning
outside the package and throws away exactly the signal a user would want to
inspect.

"Anomalous" is deliberately a narrower category than "wrong". An array
missing a terminus, for instance, is not flagged here at all: it may
genuinely sit at the edge of a contig. What *is* flagged is a part
arrangement that cannot be biologically real -- two N-termini in one array,
say, or one that is not at the start:

```{r anomaly_demo}
odd <- tibble::tibble(
  array_id = c("a1", "a1", "a1", "a2", "a2", "a2"),
  position_in_array = c(1L, 2L, 3L, 1L, 2L, 3L),
  domain_type = c("N-terminus", "N-terminus", "repeat",
                  "N-terminus", "repeat", "C-terminus"),
  rvd = c("NTERM", "NTERM", "HD", "NTERM", "HD", "CTERM")
)
x_odd <- suppressWarnings(tales(odd))
tales_anomalies(x_odd)
```

`tales(sanitize = TRUE)` drops the flagged arrays instead of merely warning
about them:

```{r sanitize_demo}
tales(odd, sanitize = TRUE)
```

Structural problems -- a duplicated key, an `NA` where one is not allowed --
are a different matter and remain hard errors either way: the table cannot
be interpreted at all, rather than merely describing something odd.

# Projections: turning parts back into sequences

Some consumers want a `tales` as a table; others want each array back as a
single string. Three functions bridge the gap:

- `tales_rvd_strings()` -- one RVD string per array, the form
  target-prediction tools consume. Drops the termini by default
  (`rvd_only = TRUE`), since prediction concerns the repeat domain only.
- `tales_coded_strings()` -- the `dom_code` sibling, for ARLEM and MAFFT's
  text mode. Keeps the termini by default, since alignment wants them as
  anchors, and separates codes with a space rather than a dash, because
  that is what those tools split on.
- `tales_domain_codes()` -- not a per-array string but the lookup table
  itself: one row per distinct `dom_code`, with the amino acid sequence
  and RVD it stands for.

```{r projections_demo}
xa <- tales_assign_domain_codes(x)
tales_coded_strings(xa)["ROI_00002"]
tales_domain_codes(xa) %>% head(4)
```

# What a `dom_code` is

This is the column the rest of the comparison machinery is built on, and it
is worth being precise about.

## It names a domain, not a repeat

A TALE part is an N-terminus, a repeat, or a C-terminus, and **all three get
`dom_code`s on the same footing**. "Domain" is the word used here
specifically because it has to cover a repeat and a terminus
indiscriminately -- "repeat code" would be the wrong name for a column that
also numbers C-termini. Two parts with identical amino acid sequences get
the same code, whatever kind of part they are.

The distinction is not pedantic. In the small example used throughout this
article:

```{r dom_code_counts}
n_distinct(xa$dom_code)
table(xa$domain_type[match(unique(xa$dom_code), xa$dom_code)])
```

`r n_distinct(xa$dom_code)` distinct codes, and only
`r sum(xa$domain_type[!duplicated(xa$dom_code)] == "repeat")` of them are
repeats -- the rest are distinct N- and C-termini. Describing the total as a
repeat count would overstate the repeat diversity by a substantial margin.

## It is what makes a TALE alignable

Aligning TALE arrays residue by residue is meaningless: the repeats are
near-identical (~34 amino acids, differing mainly at two positions), so
almost anything matches almost anything else. Giving each distinct repeat a
symbol turns an array into a *sequence of repeat units* instead, which can
be aligned the way a protein sequence is -- with insertions and deletions of
whole repeats rather than of individual residues. This is why `tales_align()` works on `dom_code` (or `rvd`), never on
`aa_seq` directly.

## It is finer than an RVD

The RVD is residues 12-13 of a repeat, and it is what determines which DNA
base the repeat binds. Two repeats can share an RVD -- the same binding
specificity -- while differing elsewhere in the protein, and they get
**different** `dom_code`s. So `rvd` is the functional layer and `dom_code`
the identity layer, and the class carries both because they answer
different questions.

In this example, the RVD `HD` (the code for cytosine) is shared by several
distinct repeat proteins:

```{r hd_example}
reps <- xa %>% filter(domain_type == "repeat")
reps %>% filter(rvd == "HD") %>% distinct(dom_code, aa_seq) %>% head(4)
```

A terminus, meanwhile, has no RVD at all: its `rvd` column holds a
placeholder -- `"NTERM"`, `"CTERM"`, or `"XXXXX"` for a terminus whose CDS
was detected but not identified (see `tales_anchor_codes()`) -- while its
`dom_code` is a real identifier of a real sequence. Another reason the two layers are not
interchangeable.

## How they are computed

Plainly: group the parts by `aa_seq`, number the groups.
`dplyr::cur_group_id()`, nothing cleverer, done by
`tales_assign_domain_codes()`.

## The numbers are only meaningful within one call

Because the numbering depends on which distinct sequences are present, code
`42` from one comparison is not code `42` from another -- add an array,
remove one, or run it on a different day's data, and the same protein can
land on a different number. This is not a wart to work around; it is a fact
to know before using `dom_code` for anything, and the package enforces it
rather than merely documenting it.

Every object `tales_assign_domain_codes()` produces carries a namespace
stamp, readable with `tales_namespace()`, derived from the exact set of
sequences that produced it:

```{r namespace_demo}
tales_namespace(xa)
```

Mixing codes from two different calls is refused rather than silently
computing a wrong answer. Splitting the example array set in two and
re-assigning codes to one part gives it a different namespace, and using it
against distances computed for the other part is an error:

```{r namespace_mismatch, error = TRUE}
dd_all <- suppressMessages(tales_domain_distances(xa))

sub <- xa[xa$array_id != "ROI_00001", ][setdiff(names(xa), "dom_code")]
xa_sub <- tales_assign_domain_codes(sub)

tales_namespace(xa_sub) == tales_namespace(xa)

tales_tale_distances(xa_sub, dd_all)
```

# Where `dom_code` leads: comparing TALEs

Minting codes is the first of three steps that together compare a set of
TALEs -- `tales_assign_domain_codes()` first, then `tales_domain_distances()`
(how different the distinct domains are from each other, pairwise), then
`tales_tale_distances()` (how different the *arrays* are, aligning their
domain-code sequences using the domain distances as the substitution cost).
`tales_compare()` runs all three and returns the corresponding `tales`,
`domain_distances` and `tale_distances` objects together:

```{r tales_compare_demo, message = FALSE, warning = FALSE}
cmp <- tales_compare(x)
names(cmp)
```

Each step is independently useful -- the domain-level distances alone
answer questions about repeat diversity, without ever running the
array-level alignment -- and each is documented in its own right; see
`tales_compare()` for the full chain.

# Subsetting keeps the class only while the contract holds

A `tales` is a tibble, so every subsetting route a tibble supports works on
it too: `[`, `filter()`, `select()`, `arrange()`, `mutate()`. What differs
from a plain tibble is what happens to the *class* along the way.

**Row subsetting never breaks anything.** Every invariant a `tales` checks
is closed under keeping a subset of rows -- which is exactly why
completeness (§1, above) is a *precondition* of specific functions rather
than part of the class itself. Filtering to one array, or to its repeats
only, is still a valid `tales`:

```{r subset_rows}
one_array <- xa[xa$array_id == "ROI_00002", ]
is_tales(one_array)
nrow(one_array)
```

**Column subsetting is where it gets interesting.** Dropping a column the
contract needs -- `array_id`, or both residue columns at once -- leaves
something that can no longer be described as a `tales`, and the class says
so by quietly stepping out of the way rather than by erroring:

```{r subset_drop_key}
no_id <- xa %>% select(-array_id)
is_tales(no_id)
class(no_id)
```

The same happens whichever route removes the load-bearing column --
`select()`, `mutate()`, or plain `[, j]`. This is deliberate: a `tales`
object that kept claiming to be one after losing its key would let every
downstream function trust a promise the object could no longer keep.
Dropping an optional column (`seqnames`, `aa_seq`, ...) has no such
consequence, and the class survives untouched.

`mutate()` earns one extra check the others do not need: it can overwrite
`array_id` or `position_in_array` *in place*, which plain subsetting cannot
do, so the result could keep every contract column while no longer being a
valid key. That case is caught separately, immediately after the mutate:

```{r subset_mutate_break_key, error = TRUE}
xa %>% mutate(position_in_array = 1L)
```

What is **not** re-checked on every verb is the row-level biology --
`tales_anomalies()` does not run again on each `filter()` call, because
anomalies are a property to inspect deliberately (§ above), not a cost to
pay on every pipe.

**Attributes travel with a valid subset**, which matters most for
`dom_code_namespace`. Cutting `xa` down to one array with `[` keeps its
namespace, because the namespace describes *the run the codes came from*,
not which rows happen to be kept right now:

```{r subset_namespace}
tales_namespace(one_array) == tales_namespace(xa)
```

That is only true because `one_array` was cut with `[`, not re-coded --
`tales_assign_domain_codes()` on a subset mints a *different* namespace, as
the mismatch example above already showed.

A `tales_msa` carries one further attribute this way, `alignment_position`,
and losing *that* while keeping everything else degrades it by exactly one
step -- to a plain `tales`, not all the way to a tibble. The next article,
on the `tales_msa` class, covers that graded degradation in full, together
with the reverse direction: how a `tales` is promoted into an alignment in
the first place.

# Next

Aligning several `tales` into a `tales_msa` and what can be read off that
alignment is the subject of the next article.
