---
title: "2. Classifying TALE sequences from genomes with tantale"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{2. Classifying TALE sequences from genomes with tantale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>>")
```

```{r package, warning=FALSE, message=FALSE}
library(tantale)
library(dplyr)
```

The [previous article](1_tale_mining.html) covered finding TALE loci in one
genome at a time. A real study usually has several genomes, and the
interesting question is not "what TALEs does this strain carry" but "which
TALEs, across strains, are versions of the same thing" -- alleles of one
locus, related by descent, rather than unrelated proteins that happen to
both be TALEs. This article covers that comparison: `tales_compare()` to
quantify how alike every pair of arrays and repeats is, and `tales_group()`
to turn that into discrete groups.

# Discovering TALEs across several genomes

This is a fresh discovery run, independent of the previous article, over
all four sample genomes together: MAI1, BAI3, PXO86, and the deliberately
error-prone BAI3-1-1.

```{r discover_all, message = FALSE, warning = FALSE}
out <- fs::dir_create(file.path(tempdir(), "tale_classification"))
genome_files <- c(
  MAI1     = system.file("extdata", "MAI1.fa",     package = "tantale", mustWork = TRUE),
  BAI3     = system.file("extdata", "BAI3.fa",     package = "tantale", mustWork = TRUE),
  `BAI3-1-1` = system.file("extdata", "BAI3-1-1.fa", package = "tantale", mustWork = TRUE),
  PXO86    = system.file("extdata", "PXO86.fa",    package = "tantale", mustWork = TRUE)
)

all_tales <- lapply(names(genome_files), function(strain) {
  strain_dir <- file.path(out, strain)
  invisible(tell_tales(subject_file = genome_files[strain], output_dir = strain_dir,
                       cterm_min_score = 300))
  tales_from_telltale(strain_dir) %>%
    mutate(array_id = paste0(strain, "_", array_id), strain = strain)
}) %>%
  suppressWarnings() %>%
  bind_rows() %>%
  tales(sanitize = TRUE)

n_distinct(all_tales$array_id)
```

`array_id` is prefixed by strain before the four objects are combined, since
`tell_tales()` numbers regions independently within each genome -- without
the prefix, `MAI1`'s `ROI_00001` and `BAI3`'s `ROI_00001` would collide.
`sanitize = TRUE` drops the frameshift-broken BAI3-1-1 arrays flagged in the
previous article, so the comparison below runs on complete arrays only.

# Quantifying relatedness: `tales_compare()`

`tales_compare()` is tantale's R implementation of the DisTAL comparison
used by the original QueTAL suite: it assigns every distinct domain
sequence a `dom_code` (see the [`tales` class article](../articles/tales-class.html)
for what that means), computes pairwise distances between distinct domains,
and from those, distances between whole arrays.

Comparing domain sequences is the expensive step, and three backends do it:
`"DECIPHER"` (the default, and by far the fastest), `"Biostrings"`, and
`"mmseq2"`. They implement the same alignment, so which one to use is a
question of what is installed and how large the dataset is, not of which
answer is right -- [a companion article](p1_tales_compare.html) compares
all three directly on a small fixture and finds them highly correlated.
Here, with `r n_distinct(all_tales$aa_seq)` domain sequences across
`r n_distinct(all_tales$array_id)` arrays, DECIPHER is used outright:

```{r tales_compare, message = FALSE}
cmp <- tales_compare(all_tales, aln_method = "DECIPHER", ncores = 4)
names(cmp)
cmp$tale_distances
```

# Allocating arrays to groups: `tales_group()`

`tales_group()` clusters arrays from `tale_distances` and returns the
`tales` object with the result attached as a `group` column -- taking the
object the distances were computed from is what lets it check that the two
actually correspond (see `?tales_group`).

Two clustering methods are available. `method = "hclust"` cuts a
dendrogram at a height chosen to yield exactly `k` groups. `method =
"k-medoids"` (the default) instead fits `cluster::pam()` across a range of
candidate group counts and picks one by its silhouette value -- `k =
"auto"` accepts that automatic pick, which is what is used here; `k` can
also be set to a specific integer instead:

```{r tales_group, fig.width=6, fig.height=4}
grouped <- tales_group(cmp$tales, cmp$tale_distances,
                       method = "k-medoids", k = "auto", k_range = 2:20)

grouped %>% distinct(array_id, group) %>% count(group) %>% arrange(desc(n))
```

Group sizes range from singletons -- an array with no close relative in
this sample -- up to several members. `plot_tree = TRUE` draws the
dendrogram version instead, useful for judging whether the chosen `k`
looks reasonable rather than trusting the silhouette value blindly:

```{r tales_group_hclust, fig.width=7, fig.height=5}
invisible(tales_group(cmp$tales, cmp$tale_distances,
                      method = "hclust", k = n_distinct(grouped$group),
                      plot_tree = TRUE))
```

# An overview across strains: `talomes_heatmap()`

With arrays assigned to groups, one natural summary is which RVD sequence
variant each strain carries in each group -- a *talome*, by analogy to a
genome, being the whole complement of TALEs a strain carries:

```{r talomes_heatmap, fig.width=9, fig.height=6}
tale_annotation <- tibble::tibble(
  array_id = names(tales_rvd_strings(grouped)),
  rvdseq = as.character(tales_rvd_strings(grouped))
) %>%
  left_join(distinct(grouped, array_id, group, strain), by = "array_id")

talomes_heatmap(tale_annotation, group_col = "group", strain_col = "strain",
                rvd_col = "rvdseq")
```

Each cell is one strain's RVD sequence variant in one group; a strain
missing from a group carries no member of it, and a cell with more than
one colour means that strain carries more than one distinct variant in
that group.

# Next

Grouping tells you *which* arrays are related. Seeing exactly *how* --
which repeats match, which are inserted or deleted relative to one
another -- needs an alignment, which is the subject of the next article.
