---
title: "Genuine truncTALEs and frameshift correction"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

Some TALEs are naturally short. [Ji et al. (2016)](https://doi.org/10.1038/ncomms13435)
and [Read et al. (2016)](https://doi.org/10.3389/fpls.2016.01516)
independently characterised truncated TAL effectors in *Xanthomonas
oryzae* -- "interfering TALEs" (iTALEs) in the first paper, "truncTALEs"
in the second, for what both describe as the same underlying phenomenon.
Both report loss of the C-terminal transcription activation domain while
the central repeat region is retained; Read et al. additionally describe
loss of part of the N-terminal region and the second nuclear
localisation signal in most examples, and a novel 28-amino-acid repeat
length not otherwise seen in TALEs. Functionally, the characterised
examples suppress the rice resistance gene *Xa1* -- Read et al.'s Tal2h
did not bind any tested candidate DNA target, consistent with a
decoy or dominant-negative role for the plant immune receptor rather
than DNA-binding transactivation. Both papers document truncTALEs
directly in PXO86 (named `Tal3`/`Tal6` in Ji et al.), the genome used
throughout this article.

That biology creates a hazard for automatic frameshift correction.
[Mining TALE sequences](tale_mining.html) already shows that neither of
tantale's two correction paths is a silver bullet against real assembly
errors. This article asks a different question: what do they do to a
TALE that is short for real biological reasons, with nothing wrong to
fix?

::: {.callout-note}
## Where this fits
This article extends the correction methods introduced in
[Mining TALE sequences](tale_mining.html#correcting-frameshifts-two-ways):
read that section first. It sits outside that walkthrough's numbered
sequence -- come here once you need to reason about a genuinely short
TALE rather than a broken one.
:::

Two concrete questions follow from the hazard above. Does
`tell_tales(correct_array = TRUE)` treat a genuine truncTALE the way it
is already known to treat a genuine assembly error, extending it
regardless of whether extension reflects real biology? And does
`correct_tales()`, the genome-wide correction route, behave any
differently? PXO86 -- documented in the literature as carrying
truncTALEs, and already shipped with the package -- answers both
directly, on real discovered arrays rather than a constructed example.

```{r setup, message = FALSE}
library(tantale)
library(dplyr)
```

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "trunctale_correction"))
```

# Two genuine truncTALEs in one genome {#sec-two-trunctales}

PXO86 is one of the good-quality assemblies used throughout this set of
articles; BAI3-1-1, used elsewhere, is the deliberately error-prone one.
PXO86 also carries two documented truncTALEs, which makes it the right
genome to answer the questions above. **On a good assembly, a short
array deserves to be taken at face value.**

```{r pxo86_path}
pxo86_fa <- system.file("extdata", "PXO86.fa", package = "tantale", mustWork = TRUE)
pxo86_raw_dir <- file.path(out, "PXO86_raw")
```

```{r discover_pxo86, message = FALSE}
invisible(tell_tales(subject_file = pxo86_fa, output_dir = pxo86_raw_dir))
pxo86 <- suppressWarnings(tales_from_telltale(pxo86_raw_dir))
```

Comparing the amino-acid length of every array's N- and C-terminus finds
them immediately -- two arrays with a dramatically shortened C-terminus,
one of which also has a reduced N-terminus:

```{r termini_widths}
pxo86 %>%
  filter(domain_type %in% c("N-terminus", "C-terminus")) %>%
  mutate(aa_length = nchar(aa_seq)) %>%
  select(array_id, domain_type, aa_length) %>%
  tidyr::pivot_wider(names_from = domain_type, values_from = aa_length) %>%
  arrange(`C-terminus`)
```

`ROI_00019`'s C-terminus is barely a seventh of the ~286 aa every other
array carries, and its N-terminus is reduced too. `ROI_00001`'s
C-terminus is shorter than normal but far less drastically so, and its
N-terminus is unaffected. Neither shows up in `tales_anomalies()`:

```{r anomalies_pxo86}
tales_anomalies(pxo86)
```

That absence is informative. `tales_anomalies()` looks for *impossible*
structure -- missing sequence data, coordinate disagreements, termini
arrangements that cannot be real. It reports nothing for either array
here: both are short, coherent proteins.

# Not the same kind of short {#sec-mechanism}

`array_report.tsv`'s `has_all_domains` column already distinguishes the
two arrays, and the reason is worth tracing back to discovery:

```{r array_report_pxo86}
pxo86_report <- readr::read_tsv(
  file.path(pxo86_raw_dir, "array_report.tsv"), show_col_types = FALSE
)
pxo86_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, has_all_domains, nterm_aa_length, cterm_aa_length, orf_coverage)
```

`tell_tales()` finds TALE loci with three separate profile HMMs -- one
each for the N-terminus, the repeat unit, and the C-terminus -- and
`has_all_domains` records whether every one of the three matched
somewhere in an array's merged hits, independently of how the final
protein turns out. `hits_report.tsv`, one of the three reports
`tell_tales()` always writes (the other two are `domains_report.tsv` and
`array_report.tsv` itself), carries every individual hit with its
profile name and `frameshift_count`, and shows exactly what happened
downstream of each array's last complete repeat:

```{r hit_level_evidence}
hits <- readr::read_tsv(file.path(pxo86_raw_dir, "hits_report.tsv"), show_col_types = FALSE)

hits %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  group_by(array_id) %>%
  summarise(
    cterm_hit_found = any(grepl("C-terminus", query_name)),
    cterm_hit_frameshift_count = frameshift_count[grepl("C-terminus", query_name)][1],
    .groups = "drop"
  )
```

`ROI_00001` has a real C-terminus hit -- full length, ~286 codons -- but
the nhmmer alignment needed two internal reframings to call it at all.
The repeat immediately before it is short and frameshifted the same way.
A C-terminus-shaped signal is there, just out of frame. `ROI_00019` has
no C-terminus hit at any frameshift count: nothing downstream of its last
repeat resembles a TALE C-terminus to any of the three profiles.

That difference carries forward into correction. Frameshift correction
-- either path -- acts on the DNA span `tell_tales()` merges from these
hits. `ROI_00001`'s span already contains a full-length,
C-terminus-shaped template, out of frame; `ROI_00019`'s span never
acquired one. Only one array gives a correction tool material to reframe
into.

# Does correction respect that difference? {#sec-does-correction-respect-it}

## `correct_array = TRUE`

```{r correct_array_path}
pxo86_decipher_dir <- file.path(out, "PXO86_decipher")
```

```{r correct_array_run, message = FALSE}
invisible(tell_tales(
  subject_file = pxo86_fa, output_dir = pxo86_decipher_dir,
  correct_array = TRUE, max_comparisons = 20
))
```

```{r correct_array_report}
decipher_report <- readr::read_tsv(
  file.path(pxo86_decipher_dir, "array_report.tsv"), show_col_types = FALSE
)
decipher_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, has_all_domains, nterm_aa_length, cterm_aa_length, orf_coverage)
```

`ROI_00001` is extended -- its C-terminus grows from 183 to 217 aa, and
`orf_coverage` improves from 83% to 86%. `ROI_00019` is untouched. This
is @sec-mechanism's distinction playing out directly:
`DECIPHER::CorrectFrameshifts()` reframes `ROI_00001` into the
C-terminus-shaped template nhmmer already found, and has no comparable
template to work with for `ROI_00019`.

::: {.callout-important}
## `max_comparisons` does not change this
`max_comparisons = 20` is used above to keep this article's build time
reasonable. The outcome is identical at every value tried, including the
full default reference set (measured, not re-run live here -- ~22
minutes against 1057 references):

| `max_comparisons` | `ROI_00001` C-terminus | `ROI_00019` C-terminus |
|---|---|---|
| 20 | 217 aa | 43 aa |
| 50 | 217 aa | 43 aa |
| 1057 (default) | 217 aa | 43 aa |

Whatever makes `DECIPHER::CorrectFrameshifts()` extend `ROI_00001` is
insensitive to how many reference sequences it is allowed to consider.
:::

## `correct_tales()`

```{r correct_tales_path}
pxo86_java_fa <- file.path(out, "PXO86_java_corrected.fa")
```

```{r correct_tales_run, message = FALSE}
corrections <- correct_tales(
  uncorrected_path = pxo86_fa, corrected_path = pxo86_java_fa,
  return_corrections = TRUE
)
```

Genome-wide, `correct_tales()` needed only `r nrow(corrections)`
correction:

```{r corrections_show}
corrections
```

```{r correct_tales_discover_path}
pxo86_java_dir <- file.path(out, "PXO86_java_tell")
```

```{r correct_tales_discover, message = FALSE}
invisible(tell_tales(
  subject_file = pxo86_java_fa, output_dir = pxo86_java_dir, correct_array = FALSE
))
java_report <- readr::read_tsv(
  file.path(pxo86_java_dir, "array_report.tsv"), show_col_types = FALSE
)
java_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, has_all_domains, nterm_aa_length, cterm_aa_length, orf_coverage)
```

A single-base insertion, and it lands inside `ROI_00001`'s own span --
consistent with @sec-mechanism, since that is the array with
frameshifted hits for the correction to act on. The effect is trivial,
though: `orf_coverage` does not move, and the C-terminus grows by a
single amino acid, against the 34 `correct_array = TRUE` adds.
`ROI_00019` is unaffected, as before.

# Summary {#sec-summary}

```{r summary_table, echo = FALSE}
tibble::tribble(
  ~array_id,   ~kind,                    ~correct_array_TRUE, ~correct_tales,
  "ROI_00019", "clean early stop, no downstream C-terminus hit", "unchanged", "unchanged",
  "ROI_00001", "genuine frameshift, C-terminus hit present (out of frame)", "over-corrected (+34 aa)", "essentially unchanged (+1 aa)"
) %>%
  knitr::kable(col.names = c("Array", "What it actually is", "correct_array = TRUE", "correct_tales()"))
```

Two genuine truncTALEs in the same genome respond differently to the
same correction tool, because they arose from different underlying
DNA-level events. A clean early stop leaves nothing downstream that
resembles a C-terminus, so a frameshift corrector has nothing to reframe
into. A genuine, evolved frameshift leaves exactly the kind of
disrupted-but-recognisable signal these tools exist to fix, and neither
tool distinguishes that signal from a real sequencing error -- so it
gets "fixed" regardless of whether fixing it reflects real biology.

**On this evidence, `correct_tales()` is the more transparent choice for
a genome suspected of carrying real truncTALEs.** It left both arrays'
sequence alone, where `tell_tales(correct_array = TRUE)` rewrote the
genuine frameshift's C-terminus as though it were an assembly error.
One genome and two arrays is a narrow base for a general rule, so the
practical takeaway is procedural rather than a blanket preference:
check `has_all_domains` and `orf_coverage` in `array_report.tsv` before
correcting, and treat any change to a documented or suspected
truncTALE's sequence as something to verify, not assume.

- A short array with `has_all_domains = FALSE` and `orf_coverage` in the
  normal range is very likely a genuine short protein. Correcting it
  fabricates sequence that was never there.
- A short array with `has_all_domains = TRUE` but reduced `orf_coverage`
  may be a genuine, fixed frameshift. `correct_array = TRUE` will extend
  it regardless of whether that is what your analysis needs.

Neither correction path can make the underlying biological judgement
call for you.

# Next

This article stands on its own. To continue the walkthrough it extends,
[return to mining TALE sequences](tale_mining.html).
