---
title: "Genuine truncTALEs and frameshift correction"
format:
  html:
    toc: true
    number-sections: true
    code-fold: show
    lightbox: true
editor_options:
  chunk_output_type: console
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

Some TALEs are naturally short. [Ji et al. (2016)](https://doi.org/10.1038/ncomms13435)
and [Read et al. (2016)](https://doi.org/10.3389/fpls.2016.01516)
independently characterised truncated TAL effectors in *Xanthomonas
oryzae*: "interfering TALEs" (iTALEs) in the first paper, "truncTALEs"
in the second, for what both describe as the same underlying phenomenon.
Both report loss of the C-terminal transcription activation domain while
the central repeat region is retained; Read et al. additionally describe
loss of part of the N-terminal region and the second nuclear
localisation signal in most examples, and a novel 28-amino-acid repeat
length not otherwise seen in TALEs. Functionally, the characterised
examples suppress the rice resistance gene *Xa1*. Read et al.'s Tal2h
did not bind any tested candidate DNA target, consistent with a
decoy or dominant-negative role for the plant immune receptor rather
than DNA-binding transactivation. Both papers document truncTALEs
directly in PXO86 (named `Tal3`/`Tal6` in Ji et al.), the genome used
throughout this article.

That biology creates a hazard for automatic frameshift correction.
[Mining TALE sequences](tale_mining.html) already shows that neither of
tantale's two correction paths is a silver bullet against real assembly
errors. This article asks a different question: what do they do to a
TALE that is short for real biological reasons, with nothing wrong to
fix?

::: {.callout-note}
## Where this fits
This article extends the correction methods introduced in
[Mining TALE sequences](tale_mining.html#correcting-frameshifts-two-ways):
read that section first. This article is a side branch of that
walkthrough, for when you need to reason about a TALE that is genuinely
short.
:::

Two concrete questions follow from the hazard above. Does
`tell_tales(correct_array = TRUE)` treat a genuine truncTALE the way it
is already known to treat a genuine assembly error, extending it
regardless of whether extension reflects real biology? And does
`correct_tales()`, the genome-wide correction route, behave any
differently? PXO86, documented in the literature as carrying truncTALEs
and shipped with the package, answers both directly on real discovered
arrays.

```{r setup, message = FALSE}
library(tantale)
library(dplyr)
```

```{r workdir}
out <- fs::dir_create(file.path(tempdir(), "trunctale_correction"))
```

# Two genuine truncTALEs in one genome {#sec-two-trunctales}

PXO86 is one of the good-quality assemblies shipped with the package;
BAI3-1-1, used in the mining article, is the deliberately error-prone one.
PXO86 also carries two documented truncTALEs, which makes it the right
genome to answer the questions above. **On a good assembly, a short
array deserves to be taken at face value.**

```{r pxo86_path}
pxo86_fa <- tantale_genome("PXO86")
pxo86_raw_dir <- file.path(out, "PXO86_raw")
```

```{r discover_pxo86, message = FALSE}
invisible(tell_tales(subject_file = pxo86_fa, output_dir = pxo86_raw_dir))
pxo86 <- suppressWarnings(tales_from_telltales(pxo86_raw_dir))
```

Comparing the amino-acid length of every array's N- and C-terminus finds
them immediately: two arrays with a shortened C-terminus and a reduced
N-terminus:

```{r termini_widths}
pxo86 %>%
  filter(domain_type %in% c("N-terminus", "C-terminus")) %>%
  mutate(aa_length = nchar(aa_seq)) %>%
  select(array_id, domain_type, aa_length) %>%
  tidyr::pivot_wider(names_from = domain_type, values_from = aa_length) %>%
  arrange(`C-terminus`)
```

`ROI_00019`'s C-terminus is barely a seventh of the 286-297 aa every
other array carries. `ROI_00001`'s is shorter than normal, at 183 aa, but
far less drastically so. Both arrays share the same reduced N-terminus,
230 aa against 283-288 aa elsewhere, in line with the partial N-terminal
loss Read et al. describe. `tales_anomalies()` reports all four termini:

```{r anomalies_pxo86}
tales_anomalies(pxo86)
```

`NTERM` and `CTERM` mark a canonical TALE terminal domain, one that can be
expected to do its usual job; any other terminus is coded `XXXXX` (see
`tales_anchor_codes()`). `tell_tales()` decides it from a search of each
terminus with the protein profile of its TALE domain, and
`array_report.tsv` holds the evidence:

```{r termini_matches}
readr::read_tsv(file.path(pxo86_raw_dir, "array_report.tsv"),
                show_col_types = FALSE) %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, nterm_aa_evalue, nterm_aa_cover, nterm_aa_domains,
         cterm_aa_evalue, cterm_aa_cover, cterm_aa_far_gap) %>%
  mutate(across(ends_with("_evalue"), ~ format(.x, digits = 2))) %>%
  knitr::kable()
```

Each short terminus matches its profile strongly: these segments are TALE
termini. But a canonical terminus must cover at least 90% of the profile
(`terminus_min_cover`), and none of these does. The two 230-aa N-termini
align with the N-terminal profile everywhere except profile positions
106-150, an internal deletion of about 45 residues, so the match comes in
two pieces (`nterm_aa_domains`) covering 84% of the profile. The 183-aa
C-terminus of `ROI_00001` aligns over its full length with profile
positions 1-183, and the 42-aa C-terminus of `ROI_00019` with positions
1-37 over its first 37 residues. These C-termini lack the far end of the
profile, 96 and 242 positions respectively (`cterm_aa_far_gap`). The 90%
threshold comes from the curated TALEs of `tale_annotations`: their
canonical termini cover 93% of the profile or more, the truncTALE termini
84% or less.

Function is what the threshold stands for. The C-terminal region of a
TALE carries the nuclear localisation signals and, at its far end, the
acidic activation domain that recruits the host's transcription
machinery: both C-termini here stop well before it, consistent with the
loss of the activation domain that Ji et al. and Read et al. report for
truncTALEs. The N-terminal region carries the type III secretion signal
and, next to the repeats, the degenerate repeats that bind the thymine
preceding the target; which function, if any, the 45-residue deletion
affects is not something sequence alone can settle.

Both arrays are real TALEs, and truncTALEs have a function of their own:
they suppress the *Xa1* resistance. So
`tales(x, sanitize = TRUE)`, which removes arrays whose data are
inconsistent or incomplete, keeps them; `sanitize = "canonical"` removes
them when an analysis needs canonical TALEs only:

```{r sanitize_pxo86}
pxo86_canonical <- suppressWarnings(tales(pxo86, sanitize = "canonical"))
setdiff(unique(pxo86$array_id), unique(pxo86_canonical$array_id))
```

# Not the same kind of short {#sec-mechanism}

`array_report.tsv`'s `cterm_dna_hit` column already distinguishes the
two arrays, and the reason is worth tracing back to discovery:

```{r array_report_pxo86}
pxo86_report <- readr::read_tsv(
  file.path(pxo86_raw_dir, "array_report.tsv"), show_col_types = FALSE
)
pxo86_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, nterm_dna_hit, cterm_dna_hit, cterm_aa_length, orf_coverage)
```

`tell_tales()` finds TALE loci with three separate profile HMMs (one
each for the N-terminus, the repeat unit, and the C-terminus), and
`nterm_dna_hit` and `cterm_dna_hit` record whether the N- and
C-terminus profiles matched somewhere in an array's merged hits,
independently of how the final protein turns out. `hits_report.tsv`, one of the three reports
`tell_tales()` always writes (the other two are `domains_report.tsv` and
`array_report.tsv` itself), carries every individual hit with its
profile name and `frameshift_count`, and shows exactly what happened
downstream of each array's last complete repeat:

```{r hit_level_evidence}
hits <- readr::read_tsv(file.path(pxo86_raw_dir, "hits_report.tsv"), show_col_types = FALSE)

hits %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  group_by(array_id) %>%
  summarise(
    cterm_hit_found = any(grepl("C-terminus", query_name, fixed = TRUE)),
    cterm_hit_frameshift_count = frameshift_count[grepl("C-terminus", query_name, fixed = TRUE)][1],
    .groups = "drop"
  )
```

`ROI_00001` has a real C-terminus hit, full length at ~286 codons, but
the nhmmer alignment needed two internal reframings to call it at all.
The repeat immediately before it is short and frameshifted the same way.
A C-terminus-shaped signal is there, just out of frame.

`ROI_00019` has no C-terminus hit. The hits that make up an array must
score above a threshold (`cterm_min_score`, 200 bits for the C-terminus
profile), set for finding arrays. Weaker matches next to the repeats are
not part of the array, but `array_report.tsv` still describes them:

```{r weak_cterm}
pxo86_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, cterm_dna_hit, cterm_dna_score, cterm_dna_cover, cterm_dna_pieces)
```

Downstream of `ROI_00019`'s last repeat, the C-terminus profile matches
in two pieces, the best at 94 bits, covering together a fifth of the
profile; a complete C-terminus scores around 1100 bits. The raw nhmmer
output (`hmmer_search_out.txt`) places them at the two ends of the
profile, positions 1-97 and 783-861. The DNA thus keeps the start and the
end of a C-terminus, with about 685 nt between them that match nothing,
and the protein stops 42 residues into the first piece.

That difference carries forward into correction. Frameshift correction,
by either path, acts on the DNA span `tell_tales()` merges from these
hits. `ROI_00001`'s span already contains a full-length,
C-terminus-shaped template, out of frame; `ROI_00019`'s span holds no
C-terminus hit, only these fragments beside it. Only one array gives a
correction tool a template to reframe into.

# Does correction respect that difference? {#sec-does-correction-respect-it}

## `correct_array = TRUE`

```{r correct_array_path}
pxo86_decipher_dir <- file.path(out, "PXO86_decipher")
```

```{r correct_array_run, message = FALSE}
invisible(tell_tales(
  subject_file = pxo86_fa, output_dir = pxo86_decipher_dir,
  correct_array = TRUE, max_comparisons = 20
))
```

```{r correct_array_report}
decipher_report <- readr::read_tsv(
  file.path(pxo86_decipher_dir, "array_report.tsv"), show_col_types = FALSE
)
decipher_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, nterm_dna_hit, cterm_dna_hit, cterm_aa_length, orf_coverage)
```

`ROI_00001` is extended: its C-terminus grows from
183 to 216 aa, and `orf_coverage` improves from 83% to 86%. `ROI_00019` is
untouched. This
is @sec-mechanism's distinction playing out directly:
`DECIPHER::CorrectFrameshifts()` reframes `ROI_00001` into the
C-terminus-shaped template nhmmer already found, and has no comparable
template to work with for `ROI_00019`.

::: {.callout-important}
## `max_comparisons` does not change this
`max_comparisons = 20` is used above to keep this article's build time
reasonable. The outcome is identical at every value tried, including the
default (50) and the full reference set of 494 sequences (measured once,
about 22 minutes, and not re-run here):

| `max_comparisons` | `ROI_00001` C-terminus | `ROI_00019` C-terminus |
|---|---|---|
| 20 | 216 aa | 42 aa |
| 50 (default) | 216 aa | 42 aa |
| all 494 | 216 aa | 42 aa |

Whatever makes `DECIPHER::CorrectFrameshifts()` extend `ROI_00001` is
insensitive to how many reference sequences it is allowed to consider.
:::

## `correct_tales()`

```{r correct_tales_path}
pxo86_java_fa <- file.path(out, "PXO86_java_corrected.fa")
```

```{r correct_tales_run, message = FALSE}
corrections <- correct_tales(
  uncorrected_path = pxo86_fa, corrected_path = pxo86_java_fa,
  return_corrections = TRUE
)
```

Genome-wide, `correct_tales()` needed only `r nrow(corrections)`
correction:

```{r corrections_show}
corrections
```

```{r correct_tales_discover_path}
pxo86_java_dir <- file.path(out, "PXO86_java_tell")
```

```{r correct_tales_discover, message = FALSE}
invisible(tell_tales(
  subject_file = pxo86_java_fa, output_dir = pxo86_java_dir, correct_array = FALSE
))
java_report <- readr::read_tsv(
  file.path(pxo86_java_dir, "array_report.tsv"), show_col_types = FALSE
)
java_report %>%
  filter(array_id %in% c("ROI_00001", "ROI_00019")) %>%
  select(array_id, nterm_dna_hit, cterm_dna_hit, cterm_aa_length, orf_coverage)
```

A single-base insertion, and it lands inside `ROI_00001`'s own span,
consistent with @sec-mechanism, since that is the array with
frameshifted hits for the correction to act on. It has no visible
effect: `ROI_00001`'s C-terminus stays at 183 aa and its `orf_coverage` at
83%, against the 33 amino acids `correct_array = TRUE` adds. `ROI_00019` is
unaffected, as before.

# Summary {#sec-summary}

```{r summary_table, echo = FALSE}
tibble::tribble(
  ~array_id,   ~kind,                    ~correct_array_TRUE, ~correct_tales,
  "ROI_00019", "early stop, only fragments of a C-terminus downstream", "unchanged", "unchanged",
  "ROI_00001", "genuine frameshift, C-terminus hit present (out of frame)", "over-corrected (+33 aa)", "unchanged"
) %>%
  knitr::kable(col.names = c("Array", "What it actually is", "correct_array = TRUE", "correct_tales()"))
```

Two genuine truncTALEs in the same genome respond differently to the
same correction tool, because they arose from different underlying
DNA-level events. In `ROI_00019`, the protein stops early and only
fragments of a C-terminus remain downstream, too weak to count as a hit,
so a frameshift corrector has no template to reframe into. A genuine,
evolved frameshift leaves exactly the kind of
disrupted-but-recognisable signal these tools exist to fix.
`correct_array = TRUE` cannot tell that signal from a sequencing error,
and "fixes" it.

**On this evidence, `correct_tales()` is the more transparent choice for
a genome suspected of carrying real truncTALEs.** It left both arrays'
sequence alone, where `tell_tales(correct_array = TRUE)` rewrote the
genuine frameshift's C-terminus as though it were an assembly error.
One genome and two arrays is a narrow base for a general rule. In
practice: check the terminus columns and `orf_coverage` in
`array_report.tsv` before correcting, and verify any change a correction
makes to a documented or suspected truncTALE's sequence.

- A short C-terminus (`cterm_aa_cover` well below 1) with
  `cterm_dna_hit = FALSE`, only weak DNA fragments (`cterm_dna_cover`
  low) and `orf_coverage` in the normal range is very likely a genuine
  short protein. Any sequence a correction adds to it was never there.
- A short C-terminus with `cterm_dna_hit = TRUE`, a full-length DNA
  match (`cterm_dna_cover` near 1) but reduced `orf_coverage` may carry a
  genuine frameshift that is part of the strain's biology.
  `correct_array = TRUE` will extend it regardless of whether that is
  what your analysis needs.

Neither correction path can make the underlying biological judgement
call for you.

# Next

This article stands on its own. To continue the walkthrough it extends,
[return to mining TALE sequences](tale_mining.html).
